### 2025-05-25 Papers/Paper One/TopicOne.tex
initial commit

\documentclass[12pt]{article}

% Packages
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{amsmath, amssymb, amsfonts}
\usepackage{siunitx}
\usepackage{array}
\usepackage[T1]{fontenc}
\usepackage{textcase}
\usepackage{caption}
\usepackage{fancyhdr}
\usepackage{lettrine}
\usepackage{geometry}
\usepackage{caption}
\usepackage{bm}
\usepackage{float}
\usepackage{enumitem}
\usepackage[backend=biber,style=ieee]{biblatex}
\usepackage{titlesec}
\usepackage[hidelinks]{hyperref}

% Geometry and Paths
\geometry{letterpaper, margin=1in}
\graphicspath{{../../images/}}

% Header and Footer
\makeatletter
\newcommand{\papertitle}{Vision Vs. LiDAR/Radar Fusion in ADS}

% re-define \sectionmark so that only top-level \section updates \leftmark
\renewcommand{\sectionmark}[1]{%
  \markboth{\Roman{section}. #1}{}%
}
% Header and Footer
\pagestyle{fancy}
\fancyhf{}
\fancyhead[R]{\scriptsize\leftmark{} – \thepage}
\fancyfoot[R]{\scriptsize\MakeUppercase{\papertitle}}
\setlength{\headheight}{15pt}
\makeatother

% Bibliography
\addbibresource{sources.bib}

\titleformat{\section}
  {\centering\large\scshape}               % shape
  {\Roman{section}.}                        % label
  {1em}                                     % sep
  {}                                        % before-code
\titlespacing*{\section}{0pt}{1.5em}{1em}

\titleformat{name=\section,numberless}
  {\centering\large\scshape}{}{1em}{}
\titlespacing*{name=\section,numberless}{0pt}{1.5em}{1em}

\titleformat{\subsection}[block]
  {\itshape}                               % italic
  {\Alph{subsection}.}                     % label
  {1em}                                     % sep
  {}                                        % before-code
\titlespacing*{\subsection}{0pt}{1em}{0.5em}

% 3) Tertiary (1), 2), …) — block, no colon
\titleformat{\subsubsection}[block]
  {\itshape}                          % italic shape
  {\arabic{subsubsection})}           % label “1)”
  {1em}                                % sep between label and title
  {}                                   % before-code (empty)
\titlespacing{\subsubsection}{1em}{0.75em}{0.5em}

% 4) Quaternary (a), b), …) — block, no colon
\titleformat{\paragraph}[block]
  {\itshape}                          % italic shape
  {\alph{paragraph})}                 % label “a)”
  {1em}                                % sep
  {}                                   % before-code
\titlespacing{\paragraph}{2em}{0.75em}{0.5em}

% CAPTIONS
\captionsetup{font=footnotesize}
\captionsetup[figure]{%
  labelformat=simple,
  name=Fig.,
  labelsep=period,
  justification=centering,
  singlelinecheck=false
}
\renewcommand{\thetable}{\Roman{table}}
\DeclareCaptionLabelFormat{dropcapsmall}{%
  {\small T}{\footnotesize\scshape able}~#2%
}
\captionsetup[table]{%
  labelformat=dropcapsmall,
  labelsep=newline,
  justification=centering,
  singlelinecheck=false
}
% Title Page
\title{\bfseries\LARGE Vision Vs. LiDAR/Radar Fusion in Autonomous Driving Systems}
\author{Arturo Salinas-Aguayo}

\begin{document}

\begin{titlepage}
    \centering
    \vspace*{3cm}
    {\Large \textsc{Vision Vs. LiDAR/Radar Fusion in Autonomous Driving
		Systems}}\\
		Arturo Salinas-Aguayo, \textit{BS Computer Engineering, Class of 2027}\\
		\today\\
    \vspace{1.5cm}
    ECE 4900W: Communicating Engineering Solutions in a Societal Context\\
    Dr. Shengli Zhou, SEC040-1255\\
    Department of Electrical and Computer Engineering\\
    \vfill
    \includegraphics[scale=0.1]{uconnlogo}\\[1em]
    University of Connecticut College of Engineering\\
    \small\tiny{Coded in \LaTeX} \\
    \vspace{1cm}
\end{titlepage}

\tableofcontents
\newpage

\section*{Abstract}
\addcontentsline{toc}{section}{Abstract}

This paper evaluates the viability of vision-primary perception systems in autonomous driving, focusing on scalability, efficiency, and environmental resilience. Traditional sensor-fusion architectures based on LiDAR and radar offer high-fidelity spatial mapping but suffer from substantial power consumption, mechanical complexity, and high unit costs, which challenge real-time deployment and mass-market feasibility. Advances in neural perception—particularly transformer-based models and tokenized camera inputs—have enabled camera-only systems to approximate LiDAR-level 3D detection accuracy while operating at a fraction of the computational cost. Benchmark evaluations reveal that vision systems achieve depth estimation errors below three percent and perform within five percent of LiDAR/radar fusion baselines in object detection. Critically, performance gaps in adverse conditions such as fog or night driving are recoverable through the selective integration of millimeter-wave radar. This radar-augmented fallback restores over 80 percent of degraded performance with minimal latency or energy impact. The study advocates for a simplified, vision-first architecture complemented by radar only in edge cases. This paradigm enhances scalability, reduces deployment overhead, and aligns with biologically plausible sensing strategies. LiDAR remains useful for simulation and benchmarking but is unnecessary for live inference.

\textbf{Index Terms—Adverse weather robustness, autonomous vehicles, deep
	learning, depth estimation, energy efficiency, Light Detection and Ranging
(LiDAR), Radar, sensor fusion, Transformer networks, vision-based perception.}
\newpage

\section{Introduction}

As vehicles become more technologically advanced, the demand for autonomous capabilities continues to rise. Features such as lane keeping, adaptive cruise control, and highway autopilot—collectively known as advanced driver-assistance systems (ADAS)—form the early building blocks of full self-driving stacks. Industry leaders now face a critical decision: how should vehicles perceive the road ahead?

Historically, high-end systems such as Waymo and Cruise relied on Light Detection and Ranging (LiDAR) and radar to construct dense, 3D maps of their surroundings. These sensors offer precise geometry and reliable velocity estimation, but impose high hardware cost, weight, and power draw. Moreover, integrating data from multiple sensors increases architectural complexity and susceptibility to calibration drift.

In contrast, Tesla, Comma.ai, and others have adopted a vision-first approach. These systems rely on arrays of low-cost cameras and train deep neural networks—often transformers—on large-scale driving datasets. Rather than reconstruct explicit 3D geometry, these models learn to infer structure and intent directly from pixels.

The debate between fusion-heavy and camera-only stacks now defines the frontier of ADS design. While LiDAR remains dominant in academia, vision-based approaches have rapidly improved in real-world performance and outpace fusion in scalability. From a societal perspective, the ability to deliver autonomy without costly sensors is especially important in emerging markets, where a single LiDAR unit may cost more than an entire car.

This report compares the two paradigms by analyzing their theoretical foundations, energy profiles, data fidelity, and robustness under real-world driving conditions. The goal is to determine whether vision-primary perception can provide sufficient safety and performance for large-scale deployment.

\section{Theory and Technical Background}

\subsection{SAE Levels of Automation}

The Society of Automotive Engineers (SAE) J3016 standard defines six levels of
vehicle automation \cite{SAEJ3016_2021}. These range from Level 0 (no
automation) to Level 5 (full automation without any human involvement).

\begin{table}[H]
  \centering
  \caption{SAE J3016 Levels of Driving Automation}
  \label{tab:sae}
  \begin{tabular}{|p{5.5cm}|p{9.5cm}|}
    \hline
    \textbf{Level} & \textbf{Description} \\
    \hline
    Level 0 – No Automation & The human driver is responsible for all tasks; the system may provide basic warnings. \\
    \hline
    Level 1 – Driver Assistance & The system assists with either steering or speed, but not both; driver must stay engaged. \\
    \hline
    Level 2 – Partial Automation & The system can manage both speed and steering, but human oversight is still mandatory. \\
    \hline
    Level 3 – Conditional Automation & The vehicle handles all driving in specific scenarios; the driver must intervene when requested. \\
    \hline
    Level 4 – High Automation & No driver input is required in controlled conditions (e.g., geofenced urban areas). \\
    \hline
    Level 5 – Full Automation & The vehicle is capable of full self-driving in all conditions without human input. \\
    \hline
  \end{tabular}
\end{table}

\subsection{LiDAR: Time-of-Flight (ToF) Principles}

LiDAR estimates object distance using time-of-flight (ToF) measurements, where a light pulse is emitted and its return time is measured. Several ToF techniques are in use:

\subsubsection{Pulsed ToF:}

\begin{equation}
R_{\text{pulse}} = \frac{c}{2} t_{\text{pulse}}
\label{eq:tof_pulse}
\end{equation}

This high-power method provides centimeter accuracy.

\subsubsection{AMCW (Amplitude-Modulated Continuous Wave):}

\begin{equation}
R_{\text{AMCW}} = \frac{c}{2} \cdot \frac{\Delta\Phi}{2\pi f_M}
\label{eq:amcw}
\end{equation}

Here, \( \Delta\Phi \) is the phase shift, and \( f_M \) is the modulation frequency \cite{2019Royo}.

\subsubsection{FMCW (Frequency-Modulated Continuous Wave):}

\begin{equation}
R_{\text{FMCW}} = f_r \cdot \frac{c T}{2B}
\label{eq:fmcw}
\end{equation}

Where \( f_r \) is the beat frequency, \( T \) the chirp duration, and \( B \) the bandwidth \cite{2019Royo}.

\subsubsection{Voxelization:}
What results from the utilization of these methods is a ``point cloud" which
corresponds to the distances the objects are away from the transmitter. The
interpretation of these distances is a process called Voxelization. There are
two forms, hard voxelization and dynamic voxelization. The in-depth explanation
of these are not required for further elaboration in this paper. Details are
shown in Figure~\ref{fig:voxelization}.
\begin{figure}[H]
	\centering
	\includegraphics[width=0.6\textwidth]{voxelization.png}
	\caption{The combining of the point cloud forms an image \autocite{RT3D}}
	\label{fig:voxelization}
\end{figure}

\subsection{Radar: Doppler and Velocity Estimation}

Radar calculates velocity via Doppler frequency shift:

\begin{equation}
v = \frac{f_D \lambda}{2}
\label{eq:radar_doppler}
\end{equation}

Where \( f_D \) is the Doppler shift and \( \lambda \) is the radar wavelength \cite{Han2023FourDRadarSurvey}. Radar’s all-weather capability makes it ideal for fallback use.

\subsection{Vision-Based Depth Estimation}

Depth from stereo vision is estimated by:

\begin{equation}
D = \frac{f b}{d}
\label{eq:depth_stereo}
\end{equation}

Where \( f \) is focal length, \( b \) baseline, and \( d \) disparity.

Monocular depth uses learned inverse depth:

\begin{equation}
\hat{D}^{-1} = g(I; \theta)
\label{eq:depth_mono}
\end{equation}

Where \( g \) is a neural network with parameters \( \theta \) \cite{Li2022BEVFormer}.

\subsection{Sensor Fusion and Noise Correlation}

Noise in sensor fusion is modeled as:

\begin{equation}
\sigma_{D_f}^2 = \sum_{i=1}^{n} w_i^2\,\sigma_{D_i}^2 + 2\sum_{i<j} w_i w_j\,\rho_{ij}\,\sigma_{D_i}\sigma_{D_j}
\label{eq:noise}
\end{equation}

Here, \( w_i \) are weights, \( \sigma_{D_i} \) are variances, and \( \rho_{ij} \) are sensor correlations \cite{Rana2023PerceptionSystems}. It is clearly shown that with the inclusion of more sensors, the potential for
more noise is increased.

\subsection{Energy and Compute Cost}

Power usage scales with input bandwidth \( B \) and model complexity \( C \):

\begin{equation}
P \propto B \cdot C
\label{eq:power_model}
\end{equation}

Vision stacks operate efficiently at \SI{2.1}{\watt}, compared to \SIrange{12}{20}{\watt} for fusion \cite{Chen2024EndToEndAD, Rana2023PerceptionSystems}.
\section{System Design and Methods}

\subsection{Vision-Primary Perception Stack}

\subsubsection{System Architecture:}

The vision-first perception stack is designed for simplicity, modularity, and real-time inference. It consists of three core stages:

\begin{enumerate}[label=\alph*)]
  \item Image Tokenization – converts camera frames into a semantically rich, compressed token representation using a learned encoder.
  \item Transformer-Based Planning – autoregressively predicts future trajectory
		tokens from visual context using a transformer model. This is the key to
		making a completely generalizable model.
  \item Control Decoding – transforms predicted spatial tokens into low-level vehicle commands for actuation.
\end{enumerate}

This pipeline emulates language models for sequential prediction and supports real-time operation on embedded platforms \autocite{goff2025learningdriveworldmodel}.

\subsubsection{Token-Based Image Compression:}

Each frame, typically $128 \times 256$ pixels, is passed through a VQGAN encoder
that produces up to 512 discrete tokens per image. This reduces bandwidth while preserving high-level semantics such as lane markings and moving objects.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{architecture}
    \caption{Vision-based pipeline with VQGAN tokenization and transformer rollout\autocite{Esser2021TamingTransformersHighResolutionImage}.}
    \label{fig:tokenizer}
\end{figure}

This compact form accelerates inference and reduces power draw without sacrificing depth accuracy \autocite{Chen2024EndToEndAD}.

\subsubsection{Autoregressive Planning with Transformers}

The token sequence is processed by a decoder-only transformer, which predicts the next spatial token at each step. This architecture parallels natural language processing and enables generalization to unseen traffic scenarios.
\begin{figure}[H]
    \centering
    \includegraphics[width=0.5\textwidth]{architecture2}
    \caption{Transformer model for sequence prediction from visual tokens \autocite{vaswani2023attentionneed}.}
    \label{fig:transformer}
\end{figure}

The transformer is trained using imitation learning on datasets like commaVQ and can be fine-tuned using reinforcement learning for policy robustness \autocite{goff2025learningdriveworldmodel}.

\subsection{LiDAR/Radar Fusion-Based Architecture}

\subsubsection{Pipeline Overview:}

Legacy ADS stacks rely on multi-modal fusion to mitigate sensor weaknesses. A typical LiDAR/radar pipeline includes:

\begin{enumerate}[label=\alph*), nosep]
  \item Data Acquisition: Captures and synchronizes LiDAR point clouds, radar waveforms, and camera frames.
  \item Preprocessing: Voxelizes LiDAR data, computes Doppler velocities from radar, and encodes image features.
  \item Fusion Layer: Combines sensor features using early, mid, or late neural fusion strategies.
  \item Detection and Tracking: Predicts 3D bounding boxes and tracks objects across time.
\end{enumerate}

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{lidararchitecture}
	\caption{Reference LiDAR/radar fusion architecture \autocite{Haghighi2024}.}
	\label{fig:lidarstack}
\end{figure}

While this pipeline offers redundancy, it incurs significant power and latency costs.

\subsubsection{Fusion Tradeoffs and Challenges:}

Fusion systems are limited by:

\begin{itemize}[nosep]
  \item High Power Draw: LiDAR alone draws \SIrange{10}{20}{\watt}, compared to \SI{2.1}{\watt} for vision-only stacks \autocite{Chen2024EndToEndAD}.
  \item Latency: Voxelization and multi-sensor alignment add \SI{20}{\milli\second} delay, reducing real-time responsiveness \autocite{Rana2023PerceptionSystems}.
  \item Noise Correlation: Overlapping modalities propagate shared errors, increasing total uncertainty as shown in Equation~\ref{eq:noise} \autocite{Rana2023PerceptionSystems}.
  \item Maintenance Complexity: Fusion requires precise calibration and can drift over time.
\end{itemize}

This complexity restricts scalability, particularly in cost-sensitive markets.

\subsection{Radar as a Resilient Fallback}

Radar bridges the gap between vision and LiDAR. It estimates object velocity
using the Doppler effect shown in Equation~\ref{eq:radar_doppler}.

Radar is immune to adverse weather and lighting and can detect motion through occlusions. When fused with vision using a transformer-based late-fusion module, it recovers over \SI{80}{\percent} of lost mAP in foggy conditions \autocite{Liao2024RadarVisionFusion}.

\subsection{A Unified Multimodal Framework}

A hybrid architecture combining vision and radar enables a balance of robustness and efficiency:

\begin{itemize}[nosep]
  \item Vision for primary perception—high-resolution, low-cost, and semantically rich.
  \item Radar for adverse conditions—weather-resilient and velocity-sensitive.
  \item LiDAR for benchmarking only—useful for training ground truth, not live deployment.
\end{itemize}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.6\textwidth]{multimodelarch}
	\caption{Modular architecture integrating vision, radar, and LiDAR inputs \autocite{Hasanujjaman2023}.}
	\label{fig:multiarch}
\end{figure}

This design maximizes generalization while minimizing cost and compute demands, and it reflects a biologically plausible model: humans drive with vision, but radar is the backup.

% -------------------- Results & Discussion ------------------
\section{Results and Discussion}

This section evaluates the empirical performance of vision-primary and LiDAR/radar fusion architectures. The comparison spans five axes: detection accuracy, computational efficiency, robustness, system complexity, and deployment scalability. Results are drawn from benchmarks using KITTI, nuScenes, and commaVQ datasets, as well as simulation and power profiling studies.

\subsection{Detection Accuracy and Depth Estimation}

On KITTI, the BEVFormer vision model achieved an AbsRel of 0.112 and RMSE of \SI{3.97}{\meter}, closely tracking the LiDAR-derived ground truth \autocite{Li2022BEVFormer}. On nuScenes, the vision-only configuration reached 56.2 3D mAP versus 61.5 mAP for the full fusion stack \autocite{Zhang2023MultiSensorFusionSurvey}.

However, when radar fallback was selectively integrated using transformer-based late fusion, the vision model recovered to 59.8 mAP \autocite{Liao2024RadarVisionFusion}. This narrows the performance delta to under 2 points—well within tolerances for safe operation—demonstrating that vision can match LiDAR-fusion with minimal augmentation.

\subsection{Computational Latency and Power Efficiency}

Inference benchmarks revealed stark contrasts. The vision-primary pipeline completed end-to-end perception and control in under \SI{9.8}{\milli\second}, consuming just \SI{2.1}{\watt} on an embedded platform \autocite{Chen2024EndToEndAD}.

In comparison, fusion systems required $\geq$ \SI{30}{\milli\second} and drew
between \SI{12}{\watt} to \SI{20}{\watt} due to voxelization, feature fusion, and redundant sensor streams \autocite{Rana2023PerceptionSystems}.

These efficiency gains are critical for EV battery life and real-time safety margins. The energy savings alone translate into longer range and lower thermal load, which directly impact vehicle design feasibility.

\subsection{Robustness in Adverse Conditions}

Under clear conditions, vision models outperform. In fog, heavy rain, or low-light scenarios, however, performance degraded by up to \SI{12.6}{\percent} in mAP \autocite{Han2023FourDRadarSurvey}. Transformer-based radar fallback restored more than \SI{80}{\percent} of this loss, effectively covering vision’s blind spots.

Unlike LiDAR, which also degrades in poor visibility and consumes continuous
power, radar is always-on but low-power—making it an ideal complement rather
than a redundant primary sensor. Figure~\ref{fig:lidar_noise} shows the
generated results in potentially adverse conditions.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.6\textwidth]{lidar_noise.png}
    \caption{LiDAR scans of a street environment under clear weather conditions (top) and with fog (bottom). The lower image shows significant point cloud degradation and sparsity due to adverse atmospheric interference \autocite{2023dreissig}.}
    \label{fig:lidar_noise}
\end{figure}
\subsection{Noise Propagation and Calibration Drift}

Sensor fusion introduces noise amplification when data is temporally misaligned
or spatially correlated. The fusion error propagation model in
Equation~\ref{eq:noise} confirms this degradation in high-$\rho$ configurations, such as LiDAR-camera overlaps \autocite{Rana2023PerceptionSystems}. Vision stacks, trained end-to-end, maintain internal consistency and minimize accumulated error.

Radar integration avoids this issue: its data is orthogonal in nature—velocity vs. spatial—which introduces minimal correlation when fused intelligently \autocite{Liao2024RadarVisionFusion}.

\subsection{Cost and Complexity of Deployment}

LiDAR units alone account for over \SI{75}{\percent} of the ADS hardware budget,
and their moving parts increase failure risk \autocite{Shetty2022LiDARvsCamera,
Sajjad2021ComparativeDetection}. Vision sensors are passive, solid-state, and
available at consumer scale. This shifts the bottleneck from manufacturing and
calibration to software modeling—an area that benefits from rapid advances in
transformers and simulation. See Figure~\ref{fig:transformer} for more details.

Modern simulators—leveraging NeRFs, 3DGS, and diffusion models—allow vision stacks to be trained entirely in synthetic environments with minimal real-world data. This accelerates iteration and de-risks real-world deployment \autocite{Haghighi2024}.

\subsection{Vision as the Core, Radar as the Shield:}

In the biological analogy, humans navigate using vision as their primary sensor. We don’t emit lasers—we perceive, interpret, and act. This report adopts the same philosophy: with a learned representation pipeline, vision is sufficient for most environments. Radar is reserved not as a crutch, but as a safety net.

By offloading primary perception to vision, we reduce cost, energy, and complexity. By retaining radar for edge cases, we maintain resilience. LiDAR, with its energy burden and calibration demands, becomes increasingly obsolete in this paradigm.

\subsection{Summary}

The data is unambiguous:

\begin{itemize}
  \item Vision stacks perform within $5\%$ of LiDAR/radar fusion across key metrics, achieving 56.2 mAP on nuScenes compared to 61.5 for fusion \autocite{Zhang2023MultiSensorFusionSurvey}.
  \item With radar fallback, this gap shrinks to $<2\%$, recovering performance to 59.8 mAP \autocite{Liao2024RadarVisionFusion}.
  \item Vision systems run 3–5x faster and consume 6–10x less
		power—\SI{9.8}{\milli\second} and \SI{2.1}{\watt} vs. $\geq$ \SI{30}{\milli\second} and \SIrange{12}{20}{\watt} for fusion \autocite{Chen2024EndToEndAD, Rana2023PerceptionSystems}.
  \item Vision stacks avoid fusion noise \autocite{Rana2023PerceptionSystems}, reduce calibration burden \autocite{Han2023FourDRadarSurvey}, and cost significantly less, with LiDAR accounting for over \SI{75}{\percent} of ADS hardware expense \autocite{Shetty2022LiDARvsCamera, Sajjad2021ComparativeDetection}.
\end{itemize}

In short, vision is sufficient for autonomy. Radar makes it resilient. LiDAR
makes it expensive, but can still have its uses in training datasets as
benchmarks. The proposed ideal multi-modal solution is outlined in
Figure~\ref{fig:multimodal}.
\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\textwidth]{multimodal}
	\caption{A Multi-Modal Approach to ADS \autocite{Hasanujjaman2023}}
	\label{fig:multimodal}
\end{figure}
\section{Conclusion}

This study set out to evaluate whether camera-first autonomous driving systems can match or exceed the performance of LiDAR/radar fusion approaches. The answer is clear: for most conditions, they can—and they do so with greater efficiency, lower cost, and fewer integration challenges.

The evidence shows:

\begin{itemize}
  \item Vision-primary pipelines achieve near-parity with fusion systems on 3D object detection and depth accuracy.
  \item Their energy and latency profiles outperform fusion systems by over 5x, enabling real-time inference on embedded hardware.
  \item Environmental failure modes, such as fog or night driving, can be mitigated with selective radar fallback, avoiding the need for costly LiDAR.
  \item End-to-end training on vision stacks reduces noise propagation and obviates sensor calibration.
\end{itemize}

These findings point to a clear path forward: deploy vision-first systems as the default, integrate radar for robustness where needed, and phase out LiDAR except in high-precision edge cases. With advances in simulation, world modeling, and vision transformers, the field now has the tools to deliver scalable autonomy without exotic sensors.

The next generation of autonomous vehicles will not be defined by how many sensors they carry, but by how intelligently they interpret the world with less. Vision, when empowered by the right models, is not just sufficient—it is optimal.
\footnote{LLM was used strictly for formatting IAW IEEE Editorial Style Manual
	for Authors. All thoughts, prose, and conclusions are
my own.}
\newpage
\addcontentsline{toc}{section}{References}
\printbibliography

\end{document}

% vim: set ft=tex tw=80 ts=2 sts=2 sw=2 noet spell:

### 2025-05-25 Slides/Presentation/slides.tex
initial commit

% ECE4900W Communicating Engineering Solutions
% Author: Arturo Salinas-Aguayo <artjsalina5@uconn.edu>
\documentclass{beamer}
\usetheme{Warsaw}
\usecolortheme{dolphin}
\usefonttheme{professionalfonts}
\setbeamertemplate{navigation symbols}{}

\title[Engineering Communication]{Communicating Engineering Solutions in a Societal Context}
\subtitle{ECE 4900W – Summer 2025}
\author[Arturo Salinas]{Arturo Salinas-Aguayo}
\institute[UConn]{University of Connecticut\\College of Engineering}
\date{\today}

\AtBeginSection[]{
  \begin{frame}{Outline}
    \tableofcontents[currentsection]
  \end{frame}
}

\begin{document}

\begin{frame}
	\titlepage
\end{frame}

\begin{frame}{Outline}
	\tableofcontents[pausesections]
\end{frame}

\section{Introduction}

\begin{frame}{Why This Course Matters}
	\begin{itemize}
		\item Engineers must communicate their solutions clearly.
		\item Societal impact matters: ethics, sustainability, inclusion.
		\item Writing and presenting ideas is just as important as building them.
	\end{itemize}
\end{frame}

\section{Course Components}

\begin{frame}{Major Assignments}
	\begin{itemize}
		\item \textbf{Paper One}: Technology Comparison (Draft + Final)
		\item \textbf{Paper Two}: Societal Impacts (Draft + Final)
		\item \textbf{Oral Presentation} based on Paper Two
		\item \textbf{Group Discussions} and \textbf{Exercises}
	\end{itemize}
\end{frame}

\subsection{Grading Breakdown}
\begin{frame}{Assessment Breakdown}
	\begin{itemize}
		\item Drafts \& Finals: \alert{63\%} combined
		\item Oral Presentation: \alert{14\%}
		\item Exercises \& Discussions: \alert{18\%}
		\item Peer Feedback: \alert{5\%}
	\end{itemize}
	\vfill
	\begin{block}{Rubrics Available}
		Check HuskyCT for detailed rubrics and formatting guides.
	\end{block}
\end{frame}

\section{Tools and Strategies}

\begin{frame}{Tips for Success}
	\begin{itemize}
		\item Start drafts early — use feedback!
		\item Visit the Writing Center
		\item Use IEEE citation style
		\item Write for clarity, audience, and purpose
	\end{itemize}
\end{frame}

\section{Summary}

\begin{frame}{Summary}
	\begin{itemize}
		\item \alert{Writing is Engineering}
		\item \alert{You will write, revise, and present}
		\item This course prepares you for communication in your future work
	\end{itemize}
	\vfill
	\begin{block}{Outlook}
		Capstone-ready communication skills that make you stand out.
	\end{block}
\end{frame}

\end{document}

### 2025-05-25 Papers/Paper One/TopicOne.tex
edited paper one

@@
 
-As vehicles become more technologically advanced, the demand for autonomous capabilities continues to rise.
-Features such as lane keeping, adaptive cruise control, and highway autopilot—collectively known as advanced driver-assistance systems (ADAS)—form the early building blocks of full self-driving stacks.
+As vehicles become more technologically advanced, the demand for autonomous
+capabilities continues to rise. Features such as lane keeping, adaptive cruise
+control, and highway autopilot—collectively known as autonomous driving systems (ADS)—form the early building blocks of full self-driving stacks.
 Industry leaders now face a critical decision: how should vehicles perceive the road ahead?
@@
 This report compares the two paradigms by analyzing their theoretical foundations, energy profiles, data fidelity, and robustness under real-world driving conditions.
-The goal is to determine whether vision-primary perception can provide sufficient safety and performance for large-scale deployment.
 
-\section{Theory and Technical Background}
+\section{Technology Description}
 
@@
 vehicle automation \cite{SAEJ3016_2021}. These range from Level 0 (no
-automation) to Level 5 (full automation without any human involvement).
+automation) to Level 5 (full automation without any human involvement). The full
+description of each level is given in Table~\ref{tab:sae}.
 
@@
 
+Currently, most ADS operate at Level 2, which is very much like being a
+driver's ed instructor sitting at secondary controls while a 15-year-old is
+operating the vehicle, ready to take control at any given moment. Level 5 is
+ideally like sitting in a train car, with
+absolutely no attention needed from the driver. This frees the driver to work on
+other important tasks such as grading papers, concentrating on a new idea, or
+taking some real downtime at the end of a stressful day.
+
 \subsection{LiDAR: Time-of-Flight (ToF) Principles}
@@
 \subsubsection{Pulsed ToF:}
-
+Pulsed TOF techniques are the simplest of the three cases in which distance is
+determined by multiplying the speed of light in a medium by the time a light
+pulse takes to travel the distance to the target:
 \begin{equation}
-R_{\text{pulse}} = \frac{c}{2} t_{\text{pulse}}
+R_{\text{pulse}} = \frac{c}{2} t_{\text{oF}},
 \label{eq:tof_pulse}
 \end{equation}
-
-This high-power method provides centimeter accuracy.
+where $R_{oF}$ is the distance the light ray travels, $c$ is the speed of
+light in free space, and $t_{oF}$ is the time it takes for the pulse of energy
+to travel from its emitter to the observed object and then back to the receiver.
+This high-power method provides centimeter accuracy. This method is shown in
+Figure~\ref{fig:pulse}.
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=0.6\textwidth]{lidar.png}
+	\caption{The Pulsed Approach to Determining Object Distance \cite{2019Royo}}
+	\label{fig:pulse}
+\end{figure}
 
 \subsubsection{AMCW (Amplitude-Modulated Continuous Wave):}
-
+The AMCW approach modulates the intensity of a continuous lightwave instead of
+laser pulsing as utilized in Eqn.~\ref{eq:tof_pulse}. This method utilizes the
+corresponding shift in phase in the received signal to calculate the distance to
+the object, $R$:
 \begin{equation}
-R_{\text{AMCW}} = \frac{c}{2} \cdot \frac{\Delta\Phi}{2\pi f_M}
+R = \frac{c}{2} \cdot \frac{\Delta\Phi}{2\pi f_M},
 \label{eq:amcw}
 \end{equation}
-
-Here, \( \Delta\Phi \) is the phase shift, and \( f_M \) is the modulation frequency \cite{2019Royo}.
-
+where \( \Delta\Phi \) is the phase shift, $c$ is the speed of light in free
+space, and \( f_M \) is the modulation
+frequency \cite{2019Royo}. It's simplified principles are shown in
+Figure~\ref{fig:amcw}.
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=0.6\textwidth]{amcw}
+	\caption{Measuring Distance by Phase Difference \cite{2019Royo}}
+	\label{fig:amcw}
+\end{figure}
 \subsubsection{FMCW (Frequency-Modulated Continuous Wave):}
-
+This final method involves a modulating the power applied to the source
+\cite{petermannopto}. The signal returned is mixed with the source signal,
+creating a beat frequency that is a measure of the probed distance, $R$:
 \begin{equation}
-R_{\text{FMCW}} = f_r \cdot \frac{c T}{2B}
+R = f_r\frac{c T}{2B},
 \label{eq:fmcw}
 \end{equation}
-
-Where \( f_r \) is the beat frequency, \( T \) the chirp duration, and \( B \) the bandwidth \cite{2019Royo}.
+where \( f_r \) is the beat frequency, \( T \) the period of the ramp, $c$ is
+the speed of light in free space, and \( B \)
+the bandwidth of the frequency sweep \cite{2019Royo}. The main parameters are
+shown in Figure~\ref{fig:fmcw}.
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=0.6\textwidth]{fmcw}
+	\caption{Measuring Distance by Frequency Modulation \cite{2019Royo}}
+	\label{fig:fmcw}
+\end{figure}
+This method is fundamentally different as there is additional processing
+overhead done in the Fourier domain rather than in the time or complex domain
+for the pulsed and AMCW approach respectively. This ultimately allows precision
+to much higher precision \cite{petermannopto}.
 
@@
 corresponds to the distances the objects are away from the transmitter. The
-interpretation of these distances is a process called Voxelization. There are
-two forms, hard voxelization and dynamic voxelization. The in-depth explanation
-of these are not required for further elaboration in this paper. Details are
-shown in Figure~\ref{fig:voxelization}.
+interpretation of these distances into 3D object boxes is a process called
+Voxelization. The high level process of breaking down point clouds into voxels and
+interpreting them is shown in Figure~\ref{fig:voxelization}.
 \begin{figure}[H]
@@
 	\includegraphics[width=0.6\textwidth]{voxelization.png}
-	\caption{The combining of the point cloud forms an image \autocite{RT3D}}
+	\caption{The combining of the point cloud forms an image where (a) is the
+	interpreted collection of point clouds inferred by the distance calculation in the sensor, (b) is the
+segmentation of this set of distances into a grid to create a 'set of sets' of
+different points, (c) is the transformer pipeline which leads to object
+identification, and (d) is the outcome with identified objects boxed
+\autocite{RT3D}.}
+
 	\label{fig:voxelization}
@@
 
-Radar calculates velocity via Doppler frequency shift:
+The Doppler effect is the change in frequency or wavelength of a wave for an
+observer or sensor, who is moving relative to the wave source. Radar calculates velocity $v$, via Doppler frequency shift:
 
 \begin{equation}
-v = \frac{f_D \lambda}{2}
+v = \frac{f_D \lambda}{2},
 \label{eq:radar_doppler}
 \end{equation}
+where \( f_D \) is the Doppler shift and \( \lambda \) is the radar wavelength
+\cite{Han2023FourDRadarSurvey}. Radar’s all-weather capability makes it ideal
+for fallback use. Figure~\ref{fig:doppler} shows the Doppler effect as it
+pertains to humans, a useful tool in understanding how this phenomena can be
+applied to this.
 
-Where \( f_D \) is the Doppler shift and \( \lambda \) is the radar wavelength \cite{Han2023FourDRadarSurvey}. Radar’s all-weather capability makes it ideal for fallback use.
-
-\subsection{Vision-Based Depth Estimation}
-
-Depth from stereo vision is estimated by:
-
-\begin{equation}
-D = \frac{f b}{d}
-\label{eq:depth_stereo}
-\end{equation}
-
-Where \( f \) is focal length, \( b \) baseline, and \( d \) disparity.
-
-Monocular depth uses learned inverse depth:
-
-\begin{equation}
-\hat{D}^{-1} = g(I; \theta)
-\label{eq:depth_mono}
-\end{equation}
-
-Where \( g \) is a neural network with parameters \( \theta \) \cite{Li2022BEVFormer}.
-
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=0.4\textwidth]{doppler}
+	\caption{The classic Doppler example showing an ambulance emitting sound. An
+	observer, from whom the ambulance leaving from will receive fewer pressure and
+will hear a deeper note, while an observer from whom the ambulance is
+approaching will receive higher pressure fluctuations and hear a higher note.}
+	\label{fig:doppler}
+\end{figure}
 \subsection{Sensor Fusion and Noise Correlation}
 
-Noise in sensor fusion is modeled as:
-
+Noise is a deviation of the response to the ideal signal and manifests itself
+differently depending on the type of sensor in phenomena such as temperature,
+EMF, or visual interference (weather). Noise in sensor fusion is modeled as:
 \begin{equation}
-\sigma_{D_f}^2 = \sum_{i=1}^{n} w_i^2\,\sigma_{D_i}^2 + 2\sum_{i<j} w_i w_j\,\rho_{ij}\,\sigma_{D_i}\sigma_{D_j}
+\sigma_{D_f}^2 = \sum_{i=1}^{n} w_i^2\,\sigma_{D_i}^2 + 2\sum_{i<j} w_i
+w_j\,\rho_{ij}\,\sigma_{D_i}\sigma_{D_j},
 \label{eq:noise}
@@
 
-Here, \( w_i \) are weights, \( \sigma_{D_i} \) are variances, and \( \rho_{ij} \) are sensor correlations \cite{Rana2023PerceptionSystems}.
-It is clearly shown that with the inclusion of more sensors, the potential for
-more noise is increased.
-
-\subsection{Energy and Compute Cost}
-
-Power usage scales with input bandwidth \( B \) and model complexity \( C \):
-
-\begin{equation}
-P \propto B \cdot C
-\label{eq:power_model}
-\end{equation}
-
-Vision stacks operate efficiently at \SI{2.1}{\watt}, compared to \SIrange{12}{20}{\watt} for fusion \cite{Chen2024EndToEndAD, Rana2023PerceptionSystems}.
-\section{System Design and Methods}
+where \( w_i \) are weights, \( \sigma_{D_i} \) are variances, and \( \rho_{ij}
+\) are sensor correlations \cite{Rana2023PerceptionSystems}.
 
@@
 Each frame, typically $128 \times 256$ pixels, is passed through a VQGAN encoder
-that produces up to 512 discrete tokens per image. This reduces bandwidth while preserving high-level semantics such as lane markings and moving objects.
+that produces up to 512 discrete tokens per image. This reduces bandwidth while
+preserving high-level semantics such as lane markings and moving objects. The
+process of encoding these images into tokens is shown in
+Figure~\ref{fig:tokenizer}.
 
@@
     \includegraphics[width=0.8\textwidth]{architecture}
-    \caption{Vision-based pipeline with VQGAN tokenization and transformer rollout\autocite{Esser2021TamingTransformersHighResolutionImage}.}
+    \caption{Vision-based pipeline with VQGAN tokenization and transformer rollout \autocite{Esser2021TamingTransformersHighResolutionImage}.}
     \label{fig:tokenizer}
@@
 \subsubsection{Autoregressive Planning with Transformers}
-
-The token sequence is processed by a decoder-only transformer, which predicts the next spatial token at each step.
-This architecture parallels natural language processing and enables generalization to unseen traffic scenarios.
+Much like how an LLM can predict the next word in a string through tokenization,
+a transformer can predict a variety of outcomes in a driving model from a static
+image that is tokenized. The token sequence is processed by a decoder-only
+transformer, which predicts the next spatial token at each step. This
+architecture parallels natural language processing and enables generalization to
+unseen traffic scenarios \cite{drivingsim}. The modern transformer architecture
+employed, shown in Figure~\ref{fig:transformer}, is taken from one of the most
+important works in recent AI advancement, ``Attention is All You Need"
+\cite{vaswani2023attentionneed}.
 \begin{figure}[H]
@@
 
-The transformer is trained using imitation learning on datasets like commaVQ and can be fine-tuned using reinforcement learning for policy robustness \autocite{goff2025learningdriveworldmodel}.
+The transformer is trained using imitation learning on datasets like commaVQ and
+can be fine-tuned using reinforcement learning for policy robustness
+\autocite{drivingsim}. This way, the training data need not have to contain
+every situation imaginable, as the process itself will be able to infer all
+possible world conditions \cite{goff2025learningdriveworldmodel}. Once all of
+these policies are applied and processing is applied, the details in
+Figure~\ref{fig:visionpipe} can be inferred.
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=0.7\textwidth]{masking}
+	\caption{The Vision Pipeline shows the original camera image, the feature map,
+	and the image mask \cite{wang2024}.}
+	\label{fig:visionpipe}
+\end{figure}
 
@@
 \end{enumerate}
+Figure~\ref{fig:lidarstack} shows how
+Eqns.~\ref{eq:tof_pulse},~\ref{eq:amcw}, and ~\ref{fig:fmcw} can be utilized and
+combined to overcome noise and predict 3D bounding boxes.
 
@@
 	\centering
-	\includegraphics[width=\textwidth]{lidararchitecture}
+	\includegraphics[width=0.7\textwidth]{lidararchitecture}
 	\caption{Reference LiDAR/radar fusion architecture \autocite{Haghighi2024}.}
@@
   \item Latency: Voxelization and multi-sensor alignment add \SI{20}{\milli\second} delay, reducing real-time responsiveness \autocite{Rana2023PerceptionSystems}.
-  \item Noise Correlation: Overlapping modalities propagate shared errors, increasing total uncertainty as shown in Equation~\ref{eq:noise} \autocite{Rana2023PerceptionSystems}.
+  \item Noise Correlation: Overlapping modalities propagate shared errors,
+		increasing total uncertainty as shown in Eqn.~\ref{eq:noise} \autocite{Rana2023PerceptionSystems}.
   \item Maintenance Complexity: Fusion requires precise calibration and can drift over time.
 \end{itemize}
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=\textwidth]{voxels}
+	\caption{Representations of point clouds from LiDAR/Radar are raw point
+		clouds, post voxel-sampling, and the generated the point clouds view in
+	Birds-Eye-View \cite{wang2024}.}
+\label{fig:voxels}
+\end{figure}
 
 This complexity restricts scalability, particularly in cost-sensitive markets.
-
-\subsection{Radar as a Resilient Fallback}
-
-Radar bridges the gap between vision and LiDAR. It estimates object velocity
-using the Doppler effect shown in Equation~\ref{eq:radar_doppler}.
-
-Radar is immune to adverse weather and lighting and can detect motion through occlusions.
-When fused with vision using a transformer-based late-fusion module, it recovers over \SI{80}{\percent} of lost mAP in foggy conditions \autocite{Liao2024RadarVisionFusion}.
-
-\subsection{A Unified Multimodal Framework}
-
-A hybrid architecture combining vision and radar enables a balance of robustness and efficiency:
-
-\begin{itemize}[nosep]
-  \item Vision for primary perception—high-resolution, low-cost, and semantically rich.
-  \item Radar for adverse conditions—weather-resilient and velocity-sensitive.
-  \item LiDAR for benchmarking only—useful for training ground truth, not live deployment.
-\end{itemize}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=0.6\textwidth]{multimodelarch}
-	\caption{Modular architecture integrating vision, radar, and LiDAR inputs \autocite{Hasanujjaman2023}.}
-	\label{fig:multiarch}
-\end{figure}
-
-This design maximizes generalization while minimizing cost and compute demands, and it reflects a biologically plausible model: humans drive with vision, but radar is the backup.
+Figure~\ref{fig:voxels} shows a typical LiDAR/Radar pipeline \cite{wang2024}.
 
 % -------------------- Results & Discussion ------------------
-\section{Results and Discussion}
+\section{Comparisons}
 
-This section evaluates the empirical performance of vision-primary and LiDAR/radar fusion architectures.
-The comparison spans five axes: detection accuracy, computational efficiency, robustness, system complexity, and deployment scalability.
-Results are drawn from benchmarks using KITTI, nuScenes, and commaVQ datasets, as well as simulation and power profiling studies.
+This section evaluates the empirical performance of vision-primary and
+LiDAR/Radar fusion architectures. The comparison spans five axes: detection
+accuracy, computational efficiency, robustness, system complexity, and
+deployment scalability. Results are drawn from benchmarks using
+KITTI \autocite{kitti}, nuScenes \autocite{nuscenes}, and commaVQ
+\autocite{goff2025learningdriveworldmodel} datasets, as well as simulation and
+power profiling studies \autocite{Rana2023PerceptionSystems}.
 
 \subsection{Detection Accuracy and Depth Estimation}
+In order to properly classify and compare these technologies, a method to
+evaluate them must be first established. For this, \textit{Precision} is the
+measure of the probability of a classifier predicting a true example and
+\textit{recall} is the number of positive samples correctly identified versus
+the number of images whose class is really a positive class, expressed by:
+\begin{equation}
+	recall = \frac{TP}{TP + FN},
+	\label{eq:recall}
+\end{equation}
+where $TP$ is the number of true positives, $FP$ is the number of false
+positives, and $FN$ is the number of false negatives.
 
-On KITTI, the BEVFormer vision model achieved an AbsRel of 0.112 and RMSE of \SI{3.97}{\meter}, closely tracking the LiDAR-derived ground truth \autocite{Li2022BEVFormer}.
-On nuScenes, the vision-only configuration reached 56.2 3D mAP versus 61.5 mAP for the full fusion stack \autocite{Zhang2023MultiSensorFusionSurvey}.
+Building off of this, the most commonly used evaluation metric for object
+detection is $AP$, or Average Precision. This is defined by:
+\begin{equation}
+	AP = \int_0^1 p(r)dr,
+	\label{eq:ap}
+\end{equation}
+where $p(r)$ is the function of precision about the recall.
 
-However, when radar fallback was selectively integrated using transformer-based late fusion, the vision model recovered to 59.8 mAP \autocite{Liao2024RadarVisionFusion}.
+However, for a
+better statistical representation of the data, a more popular metric is used
+defined as $mAP$, or Mean Average Position, which can be expressed as:
+\begin{equation}
+	mAP = \frac{\sum_{i=1}^N AP_i}{N},
+	\label{eq:map}
+\end{equation}
+where, $N$ is the total number of classes and $AP_i$ denotes the average
+precision of the $i$-th class \cite{wang2024}.
+
+\subsection{Findings}
+On KITTI, the BEVFormer vision model achieved an AbsRel of 0.112 and RMSE of
+\SI{3.97}{\meter}, closely tracking the LiDAR-derived ground truth
+\autocite{Li2022BEVFormer}. On nuScenes, the vision-only configuration reached
+56.2 3D $mAP$ versus 61.5 $mAP$ for the full fusion stack \autocite{Zhang2023MultiSensorFusionSurvey}.
+
+However, when radar fallback was selectively integrated using transformer-based
+late fusion, the vision model recovered to 59.8 $mAP$ \autocite{Liao2024RadarVisionFusion}.
 This narrows the performance delta to under 2 points—well within tolerances for safe operation—demonstrating that vision can match LiDAR-fusion with minimal augmentation.
@@
 \subsection{Computational Latency and Power Efficiency}
-
 Inference benchmarks revealed stark contrasts.
@@
 
-Under clear conditions, vision models outperform.
-In fog, heavy rain, or low-light scenarios, however, performance degraded by up to \SI{12.6}{\percent} in mAP \autocite{Han2023FourDRadarSurvey}.
-Transformer-based radar fallback restored more than \SI{80}{\percent} of this loss, effectively covering vision’s blind spots.
+Under clear conditions, vision models outperform. In fog, heavy rain, or
+low-light scenarios, however, performance degraded by up to \SI{12.6}{\percent}
+in $mAP$ \autocite{Han2023FourDRadarSurvey}. Transformer-based radar fallback restored more than \SI{80}{\percent} of this loss, effectively covering vision’s blind spots.
 
@@
 or spatially correlated. The fusion error propagation model in
-Equation~\ref{eq:noise} confirms this degradation in high-$\rho$ configurations, such as LiDAR-camera overlaps \autocite{Rana2023PerceptionSystems}.
-Vision stacks, trained end-to-end, maintain internal consistency and minimize accumulated error.
+Eqn.~\ref{eq:noise} confirms this degradation in high-$\rho$ configurations,
+such as LiDAR-camera overlaps \autocite{Rana2023PerceptionSystems}. Vision
+stacks, trained end-to-end, maintain internal consistency and minimize
+accumulated error \cite{goff2025learningdriveworldmodel}.
 
@@
 
-In the biological analogy, humans navigate using vision as their primary sensor.
-We don’t emit lasers—we perceive, interpret, and act.
-This report adopts the same philosophy: with a learned representation pipeline, vision is sufficient for most environments.
-Radar is reserved not as a crutch, but as a safety net.
+Elon Musk famously claimed ``LiDAR is a fool's errand" to much surprise at the
+time \cite{TeslaAutonomyDay2019}. In 2019, this was perhaps too forward
+thinking, but was it without bounds? In the biological analogy, humans navigate using vision as their primary sensor. We don’t emit lasers—we perceive, interpret, and act.
 
@@
 
+\subsection{Radar as a Resilient Fallback}
+
+Radar bridges the gap between vision and LiDAR. It estimates object velocity
+using the Doppler effect shown in Eqn.~\ref{eq:radar_doppler} and in
+Figure~\ref{fig:doppler}.
+
+Radar is immune to adverse weather and lighting and can detect motion through
+occlusions. When fused with vision using a transformer-based late-fusion module,
+it recovers over \SI{80}{\percent} of lost $mAP$ in foggy conditions
+\autocite{Liao2024RadarVisionFusion}. In low speed, high traffic conditions,
+radar can be utilized to better estimate distance to other vehicles with less
+cooling necessary. This is very enticing for high population areas such as Los
+Angeles where stop and go traffic on exceedingly hot roads is a daily occurrence.
+
+\subsection{A Unified Multimodal Framework}
+
+What companies such as Comma.ai and Tesla are doing is employing a hybrid
+architecture combining vision and radar which enable a balance of robustness and efficiency:
+
+\begin{itemize}[nosep]
+  \item Vision for primary perception—high-resolution, low-cost, and semantically rich.
+  \item Radar for adverse conditions—weather-resilient and velocity-sensitive.
+  \item LiDAR for benchmarking only—useful for training ground truth, not live deployment.
+\end{itemize}
+Figure~\ref{fig:multiarch} shows the modular stack which can be tuned via
+hyperparameters during the training stage to give more weight to favor certain
+inputs. Such an approach is utilized by Comma.ai to prioritize key sensor inputs
+to the model deployed in Openpilot \cite{goff2025learningdriveworldmodel}.
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=0.65\textwidth]{multimodelarch}
+	\caption{Modular architecture integrating vision, radar, and LiDAR inputs \autocite{Hasanujjaman2023}.}
+	\label{fig:multiarch}
+\end{figure}
+
+This design maximizes generalization while minimizing cost and computation demands, and it reflects a biologically plausible model: humans drive with vision, but radar is the backup.
+
 \subsection{Summary}
-
-The data is unambiguous:
-
+In order to eventually achieve SAE Level 5 Autonomy, the currently existing data points to the possibility of vision-only stacks
+surpassing fusion models in the near future:
 \begin{itemize}
-  \item Vision stacks perform within $5\%$ of LiDAR/radar fusion across key metrics, achieving 56.2 mAP on nuScenes compared to 61.5 for fusion \autocite{Zhang2023MultiSensorFusionSurvey}.
-  \item With radar fallback, this gap shrinks to $<2\%$, recovering performance to 59.8 mAP \autocite{Liao2024RadarVisionFusion}.
+  \item Vision stacks perform within $5\%$ of LiDAR/Radar fusion across key
+		metrics, achieving 56.2 $mAP$ on nuScenes compared to 61.5 for fusion \autocite{Zhang2023MultiSensorFusionSurvey}.
+  \item With radar fallback, this gap shrinks to $<2\%$, recovering performance
+		to 59.8 $mAP$ \autocite{Liao2024RadarVisionFusion}.
   \item Vision systems run 3–5x faster and consume 6–10x less
@@
 makes it expensive, but can still have its uses in training datasets as
-benchmarks. The proposed ideal multi-modal solution is outlined in
+benchmarks. The ideal multi-modal solution is outlined in
 Figure~\ref{fig:multimodal}.
@@
 
-This study set out to evaluate whether camera-first autonomous driving systems can match or exceed the performance of LiDAR/radar fusion approaches.
-The answer is clear: for most conditions, they can—and they do so with greater efficiency, lower cost, and fewer integration challenges.
-
-The evidence shows:
-
-\begin{itemize}
-  \item Vision-primary pipelines achieve near-parity with fusion systems on 3D object detection and depth accuracy.
-  \item Their energy and latency profiles outperform fusion systems by over 5x, enabling real-time inference on embedded hardware.
-  \item Environmental failure modes, such as fog or night driving, can be mitigated with selective radar fallback, avoiding the need for costly LiDAR.
-  \item End-to-end training on vision stacks reduces noise propagation and obviates sensor calibration.
-\end{itemize}
-
-These findings point to a clear path forward: deploy vision-first systems as the default, integrate radar for robustness where needed, and phase out LiDAR except in high-precision edge cases.
-With advances in simulation, world modeling, and vision transformers, the field now has the tools to deliver scalable autonomy without exotic sensors.
-
-The next generation of autonomous vehicles will not be defined by how many sensors they carry, but by how intelligently they interpret the world with less.
-Vision, when empowered by the right models, is not just sufficient—it is optimal.
-\footnote{LLM was used strictly for formatting IAW IEEE Editorial Style Manual
+This paper sets out to compare the existing technology employed in Level 2 ADS systems
+deployed in production around the world today. While approaches such as Waymo's
+LiDAR/Vision/Radar fusion are commendable and show great promise in deployment, the
+drawbacks mentioned and advancements in the vision stack in particular offer a
+clear alternative that is cheaper and more energy efficient becoming more effective each day.
+\footnote{An LLM was used strictly for formatting IAW IEEE Editorial Style Manual
 	for Authors. All thoughts, prose, and conclusions are

### 2025-05-25 Papers/Paper One/TopicOne.tex
ammend paper one

@@
 
-This paper evaluates the viability of vision-primary perception systems in autonomous driving, focusing on scalability, efficiency, and environmental resilience.
-Traditional sensor-fusion architectures based on LiDAR and radar offer high-fidelity spatial mapping but suffer from substantial power consumption, mechanical complexity, and high unit costs, which challenge real-time deployment and mass-market feasibility.
-Advances in neural perception—particularly transformer-based models and tokenized camera inputs—have enabled camera-only systems to approximate LiDAR-level 3D detection accuracy while operating at a fraction of the computational cost.
-Benchmark evaluations reveal that vision systems achieve depth estimation errors below three percent and perform within five percent of LiDAR/radar fusion baselines in object detection.
+This paper evaluates the viability of vision-primary perception systems in
+autonomous driving, focusing on scalability, efficiency, and environmental
+resilience. Traditional sensor-fusion architectures based on LiDAR and radar
+offer high-fidelity spatial mapping but suffer from substantial power
+consumption, mechanical complexity, and high unit costs, which challenge
+real-time deployment and mass-market feasibility. Advances in neural
+perception—particularly transformer-based models and tokenized camera
+inputs—have enabled camera-only systems to approximate LiDAR-level 3D detection
+accuracy while operating at a fraction of the computational cost. Benchmark
+evaluations reveal that vision systems achieve depth estimation errors below
+three percent and perform within five percent of LiDAR/Radar fusion baselines in object detection.
 Critically, performance gaps in adverse conditions such as fog or night driving are recoverable through the selective integration of millimeter-wave radar.
@@
 	\centering
-	\includegraphics[width=0.7\textwidth]{masking}
+	\includegraphics[width=0.9\textwidth]{masking}
 	\caption{The Vision Pipeline shows the original camera image, the feature map,
@@
 
+\section{Further Discussions}
 \subsection{Vision as the Core, Radar as the Shield:}

### 2025-05-28 Papers/Paper One/TopicOne.tex
work on cleaning up

@@
 \begin{equation}
-R_{\text{pulse}} = \frac{c}{2} t_{\text{oF}},
+R = \frac{c}{2} t_{\text{oF}},
 \label{eq:tof_pulse}
 \end{equation}
-where $R_{oF}$ is the distance the light ray travels, $c$ is the speed of
+where $R$ is the distance the light ray travels, $c$ is the speed of
 light in free space, and $t_{oF}$ is the time it takes for the pulse of energy
 to travel from its emitter to the observed object and then back to the receiver.
-This high-power method provides centimeter accuracy. This method is shown in
+This high-power method provides centimeter accuracy \cite{2019Royo}. This method is shown in
 Figure~\ref{fig:pulse}.

### 2025-06-01 Papers/Paper Two/papertwo.tex
work on cleaning up

--- 2025-05-25 Papers/Paper One/TopicOne.tex
@@
+%Copyright (c) 2025 arturo salinas-aguayo. All Rights Reserved.
 \documentclass[12pt]{article}
@@
 \begin{document}
-
-\begin{titlepage}
-    \centering
-    \vspace*{3cm}
-    {\Large \textsc{Vision Vs. LiDAR/Radar Fusion in Autonomous Driving
-		Systems}}\\
-		Arturo Salinas-Aguayo, \textit{BS Computer Engineering, Class of 2027}\\
-		\today\\
-    \vspace{1.5cm}
-    ECE 4900W: Communicating Engineering Solutions in a Societal Context\\
-    Dr. Shengli Zhou, SEC040-1255\\
-    Department of Electrical and Computer Engineering\\
-    \vfill
-    \includegraphics[scale=0.1]{uconnlogo}\\[1em]
-    University of Connecticut College of Engineering\\
-    \small\tiny{Coded in \LaTeX} \\
-    \vspace{1cm}
-\end{titlepage}
-
-\tableofcontents
-\newpage
-
-\section*{Abstract}
-\addcontentsline{toc}{section}{Abstract}
-
-This paper evaluates the viability of vision-primary perception systems in autonomous driving, focusing on scalability, efficiency, and environmental resilience.
-Traditional sensor-fusion architectures based on LiDAR and radar offer high-fidelity spatial mapping but suffer from substantial power consumption, mechanical complexity, and high unit costs, which challenge real-time deployment and mass-market feasibility.
-Advances in neural perception—particularly transformer-based models and tokenized camera inputs—have enabled camera-only systems to approximate LiDAR-level 3D detection accuracy while operating at a fraction of the computational cost.
-Benchmark evaluations reveal that vision systems achieve depth estimation errors below three percent and perform within five percent of LiDAR/radar fusion baselines in object detection.
-Critically, performance gaps in adverse conditions such as fog or night driving are recoverable through the selective integration of millimeter-wave radar.
-This radar-augmented fallback restores over 80 percent of degraded performance with minimal latency or energy impact.
-The study advocates for a simplified, vision-first architecture complemented by radar only in edge cases.
-This paradigm enhances scalability, reduces deployment overhead, and aligns with biologically plausible sensing strategies.
-LiDAR remains useful for simulation and benchmarking but is unnecessary for live inference.
-
-\textbf{Index Terms—Adverse weather robustness, autonomous vehicles, deep
-	learning, depth estimation, energy efficiency, Light Detection and Ranging
-(LiDAR), Radar, sensor fusion, Transformer networks, vision-based perception.}
-\newpage
-
-\section{Introduction}
-
-As vehicles become more technologically advanced, the demand for autonomous
-capabilities continues to rise. Features such as lane keeping, adaptive cruise
-control, and highway autopilot—collectively known as autonomous driving systems (ADS)—form the early building blocks of full self-driving stacks.
-Industry leaders now face a critical decision: how should vehicles perceive the road ahead?
-
-Historically, high-end systems such as Waymo and Cruise relied on Light Detection and Ranging (LiDAR) and radar to construct dense, 3D maps of their surroundings.
-These sensors offer precise geometry and reliable velocity estimation, but impose high hardware cost, weight, and power draw.
-Moreover, integrating data from multiple sensors increases architectural complexity and susceptibility to calibration drift.
-
-In contrast, Tesla, Comma.ai, and others have adopted a vision-first approach.
-These systems rely on arrays of low-cost cameras and train deep neural networks—often transformers—on large-scale driving datasets.
-Rather than reconstruct explicit 3D geometry, these models learn to infer structure and intent directly from pixels.
-
-The debate between fusion-heavy and camera-only stacks now defines the frontier of ADS design.
-While LiDAR remains dominant in academia, vision-based approaches have rapidly improved in real-world performance and outpace fusion in scalability.
-From a societal perspective, the ability to deliver autonomy without costly sensors is especially important in emerging markets, where a single LiDAR unit may cost more than an entire car.
-
-This report compares the two paradigms by analyzing their theoretical foundations, energy profiles, data fidelity, and robustness under real-world driving conditions.
-
-\section{Technology Description}
-
-\subsection{SAE Levels of Automation}
-
-The Society of Automotive Engineers (SAE) J3016 standard defines six levels of
-vehicle automation \cite{SAEJ3016_2021}. These range from Level 0 (no
-automation) to Level 5 (full automation without any human involvement). The full
-description of each level is given in Table~\ref{tab:sae}.
-
-\begin{table}[H]
-  \centering
-  \caption{SAE J3016 Levels of Driving Automation}
-  \label{tab:sae}
-  \begin{tabular}{|p{5.5cm}|p{9.5cm}|}
-    \hline
-    \textbf{Level} & \textbf{Description} \\
-    \hline
-    Level 0 – No Automation & The human driver is responsible for all tasks; the system may provide basic warnings. \\
-    \hline
-    Level 1 – Driver Assistance & The system assists with either steering or speed, but not both; driver must stay engaged. \\
-    \hline
-    Level 2 – Partial Automation & The system can manage both speed and steering, but human oversight is still mandatory. \\
-    \hline
-    Level 3 – Conditional Automation & The vehicle handles all driving in specific scenarios; the driver must intervene when requested. \\
-    \hline
-    Level 4 – High Automation & No driver input is required in controlled conditions (e.g., geofenced urban areas). \\
-    \hline
-    Level 5 – Full Automation & The vehicle is capable of full self-driving in all conditions without human input. \\
-    \hline
-  \end{tabular}
-\end{table}
-
-Currently, most ADS operate at Level 2, which is very much like being a
-driver's ed instructor sitting at secondary controls while a 15-year-old is
-operating the vehicle, ready to take control at any given moment. Level 5 is
-ideally like sitting in a train car, with
-absolutely no attention needed from the driver. This frees the driver to work on
-other important tasks such as grading papers, concentrating on a new idea, or
-taking some real downtime at the end of a stressful day.
-
-\subsection{LiDAR: Time-of-Flight (ToF) Principles}
-
-LiDAR estimates object distance using time-of-flight (ToF) measurements, where a light pulse is emitted and its return time is measured. Several ToF techniques are in use:
-
-\subsubsection{Pulsed ToF:}
-Pulsed TOF techniques are the simplest of the three cases in which distance is
-determined by multiplying the speed of light in a medium by the time a light
-pulse takes to travel the distance to the target:
-\begin{equation}
-R_{\text{pulse}} = \frac{c}{2} t_{\text{oF}},
-\label{eq:tof_pulse}
-\end{equation}
-where $R_{oF}$ is the distance the light ray travels, $c$ is the speed of
-light in free space, and $t_{oF}$ is the time it takes for the pulse of energy
-to travel from its emitter to the observed object and then back to the receiver.
-This high-power method provides centimeter accuracy. This method is shown in
-Figure~\ref{fig:pulse}.
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=0.6\textwidth]{lidar.png}
-	\caption{The Pulsed Approach to Determining Object Distance \cite{2019Royo}}
-	\label{fig:pulse}
-\end{figure}
-
-\subsubsection{AMCW (Amplitude-Modulated Continuous Wave):}
-The AMCW approach modulates the intensity of a continuous lightwave instead of
-laser pulsing as utilized in Eqn.~\ref{eq:tof_pulse}. This method utilizes the
-corresponding shift in phase in the received signal to calculate the distance to
-the object, $R$:
-\begin{equation}
-R = \frac{c}{2} \cdot \frac{\Delta\Phi}{2\pi f_M},
-\label{eq:amcw}
-\end{equation}
-where \( \Delta\Phi \) is the phase shift, $c$ is the speed of light in free
-space, and \( f_M \) is the modulation
-frequency \cite{2019Royo}. It's simplified principles are shown in
-Figure~\ref{fig:amcw}.
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=0.6\textwidth]{amcw}
-	\caption{Measuring Distance by Phase Difference \cite{2019Royo}}
-	\label{fig:amcw}
-\end{figure}
-\subsubsection{FMCW (Frequency-Modulated Continuous Wave):}
-This final method involves a modulating the power applied to the source
-\cite{petermannopto}. The signal returned is mixed with the source signal,
-creating a beat frequency that is a measure of the probed distance, $R$:
-\begin{equation}
-R = f_r\frac{c T}{2B},
-\label{eq:fmcw}
-\end{equation}
-where \( f_r \) is the beat frequency, \( T \) the period of the ramp, $c$ is
-the speed of light in free space, and \( B \)
-the bandwidth of the frequency sweep \cite{2019Royo}. The main parameters are
-shown in Figure~\ref{fig:fmcw}.
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=0.6\textwidth]{fmcw}
-	\caption{Measuring Distance by Frequency Modulation \cite{2019Royo}}
-	\label{fig:fmcw}
-\end{figure}
-This method is fundamentally different as there is additional processing
-overhead done in the Fourier domain rather than in the time or complex domain
-for the pulsed and AMCW approach respectively. This ultimately allows precision
-to much higher precision \cite{petermannopto}.
-
-\subsubsection{Voxelization:}
-What results from the utilization of these methods is a ``point cloud" which
-corresponds to the distances the objects are away from the transmitter. The
-interpretation of these distances into 3D object boxes is a process called
-Voxelization. The high level process of breaking down point clouds into voxels and
-interpreting them is shown in Figure~\ref{fig:voxelization}.
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=0.6\textwidth]{voxelization.png}
-	\caption{The combining of the point cloud forms an image where (a) is the
-	interpreted collection of point clouds inferred by the distance calculation in the sensor, (b) is the
-segmentation of this set of distances into a grid to create a 'set of sets' of
-different points, (c) is the transformer pipeline which leads to object
-identification, and (d) is the outcome with identified objects boxed
-\autocite{RT3D}.}
-
-	\label{fig:voxelization}
-\end{figure}
-
-\subsection{Radar: Doppler and Velocity Estimation}
-
-The Doppler effect is the change in frequency or wavelength of a wave for an
-observer or sensor, who is moving relative to the wave source. Radar calculates velocity $v$, via Doppler frequency shift:
-
-\begin{equation}
-v = \frac{f_D \lambda}{2},
-\label{eq:radar_doppler}
-\end{equation}
-where \( f_D \) is the Doppler shift and \( \lambda \) is the radar wavelength
-\cite{Han2023FourDRadarSurvey}. Radar’s all-weather capability makes it ideal
-for fallback use. Figure~\ref{fig:doppler} shows the Doppler effect as it
-pertains to humans, a useful tool in understanding how this phenomena can be
-applied to this.
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=0.4\textwidth]{doppler}
-	\caption{The classic Doppler example showing an ambulance emitting sound. An
-	observer, from whom the ambulance leaving from will receive fewer pressure and
-will hear a deeper note, while an observer from whom the ambulance is
-approaching will receive higher pressure fluctuations and hear a higher note.}
-	\label{fig:doppler}
-\end{figure}
-\subsection{Sensor Fusion and Noise Correlation}
-
-Noise is a deviation of the response to the ideal signal and manifests itself
-differently depending on the type of sensor in phenomena such as temperature,
-EMF, or visual interference (weather). Noise in sensor fusion is modeled as:
-\begin{equation}
-\sigma_{D_f}^2 = \sum_{i=1}^{n} w_i^2\,\sigma_{D_i}^2 + 2\sum_{i<j} w_i
-w_j\,\rho_{ij}\,\sigma_{D_i}\sigma_{D_j},
-\label{eq:noise}
-\end{equation}
-
-where \( w_i \) are weights, \( \sigma_{D_i} \) are variances, and \( \rho_{ij}
-\) are sensor correlations \cite{Rana2023PerceptionSystems}.
-
-\subsection{Vision-Primary Perception Stack}
-
-\subsubsection{System Architecture:}
-
-The vision-first perception stack is designed for simplicity, modularity, and real-time inference. It consists of three core stages:
-
-\begin{enumerate}[label=\alph*)]
-  \item Image Tokenization – converts camera frames into a semantically rich, compressed token representation using a learned encoder.
-  \item Transformer-Based Planning – autoregressively predicts future trajectory
-		tokens from visual context using a transformer model. This is the key to
-		making a completely generalizable model.
-  \item Control Decoding – transforms predicted spatial tokens into low-level vehicle commands for actuation.
-\end{enumerate}
-
-This pipeline emulates language models for sequential prediction and supports real-time operation on embedded platforms \autocite{goff2025learningdriveworldmodel}.
-
-\subsubsection{Token-Based Image Compression:}
-
-Each frame, typically $128 \times 256$ pixels, is passed through a VQGAN encoder
-that produces up to 512 discrete tokens per image. This reduces bandwidth while
-preserving high-level semantics such as lane markings and moving objects. The
-process of encoding these images into tokens is shown in
-Figure~\ref{fig:tokenizer}.
-
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=0.8\textwidth]{architecture}
-    \caption{Vision-based pipeline with VQGAN tokenization and transformer rollout \autocite{Esser2021TamingTransformersHighResolutionImage}.}
-    \label{fig:tokenizer}
-\end{figure}
-
-This compact form accelerates inference and reduces power draw without sacrificing depth accuracy \autocite{Chen2024EndToEndAD}.
-
-\subsubsection{Autoregressive Planning with Transformers}
-Much like how an LLM can predict the next word in a string through tokenization,
-a transformer can predict a variety of outcomes in a driving model from a static
-image that is tokenized. The token sequence is processed by a decoder-only
-transformer, which predicts the next spatial token at each step. This
-architecture parallels natural language processing and enables generalization to
-unseen traffic scenarios \cite{drivingsim}. The modern transformer architecture
-employed, shown in Figure~\ref{fig:transformer}, is taken from one of the most
-important works in recent AI advancement, ``Attention is All You Need"
-\cite{vaswani2023attentionneed}.
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=0.5\textwidth]{architecture2}
-    \caption{Transformer model for sequence prediction from visual tokens \autocite{vaswani2023attentionneed}.}
-    \label{fig:transformer}
-\end{figure}
-
-The transformer is trained using imitation learning on datasets like commaVQ and
-can be fine-tuned using reinforcement learning for policy robustness
-\autocite{drivingsim}. This way, the training data need not have to contain
-every situation imaginable, as the process itself will be able to infer all
-possible world conditions \cite{goff2025learningdriveworldmodel}. Once all of
-these policies are applied and processing is applied, the details in
-Figure~\ref{fig:visionpipe} can be inferred.
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=0.7\textwidth]{masking}
-	\caption{The Vision Pipeline shows the original camera image, the feature map,
-	and the image mask \cite{wang2024}.}
-	\label{fig:visionpipe}
-\end{figure}
-
-\subsection{LiDAR/Radar Fusion-Based Architecture}
-
-\subsubsection{Pipeline Overview:}
-
-Legacy ADS stacks rely on multi-modal fusion to mitigate sensor weaknesses. A typical LiDAR/radar pipeline includes:
-
-\begin{enumerate}[label=\alph*), nosep]
-  \item Data Acquisition: Captures and synchronizes LiDAR point clouds, radar waveforms, and camera frames.
-  \item Preprocessing: Voxelizes LiDAR data, computes Doppler velocities from radar, and encodes image features.
-  \item Fusion Layer: Combines sensor features using early, mid, or late neural fusion strategies.
-  \item Detection and Tracking: Predicts 3D bounding boxes and tracks objects across time.
-\end{enumerate}
-Figure~\ref{fig:lidarstack} shows how
-Eqns.~\ref{eq:tof_pulse},~\ref{eq:amcw}, and ~\ref{fig:fmcw} can be utilized and
-combined to overcome noise and predict 3D bounding boxes.
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=0.7\textwidth]{lidararchitecture}
-	\caption{Reference LiDAR/radar fusion architecture \autocite{Haghighi2024}.}
-	\label{fig:lidarstack}
-\end{figure}
-
-While this pipeline offers redundancy, it incurs significant power and latency costs.
-
-\subsubsection{Fusion Tradeoffs and Challenges:}
-
-Fusion systems are limited by:
-
-\begin{itemize}[nosep]
-  \item High Power Draw: LiDAR alone draws \SIrange{10}{20}{\watt}, compared to \SI{2.1}{\watt} for vision-only stacks \autocite{Chen2024EndToEndAD}.
-  \item Latency: Voxelization and multi-sensor alignment add \SI{20}{\milli\second} delay, reducing real-time responsiveness \autocite{Rana2023PerceptionSystems}.
-  \item Noise Correlation: Overlapping modalities propagate shared errors,
-		increasing total uncertainty as shown in Eqn.~\ref{eq:noise} \autocite{Rana2023PerceptionSystems}.
-  \item Maintenance Complexity: Fusion requires precise calibration and can drift over time.
-\end{itemize}
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=\textwidth]{voxels}
-	\caption{Representations of point clouds from LiDAR/Radar are raw point
-		clouds, post voxel-sampling, and the generated the point clouds view in
-	Birds-Eye-View \cite{wang2024}.}
-\label{fig:voxels}
-\end{figure}
-
-This complexity restricts scalability, particularly in cost-sensitive markets.
-Figure~\ref{fig:voxels} shows a typical LiDAR/Radar pipeline \cite{wang2024}.
-
-% -------------------- Results & Discussion ------------------
-\section{Comparisons}
-
-This section evaluates the empirical performance of vision-primary and
-LiDAR/Radar fusion architectures. The comparison spans five axes: detection
-accuracy, computational efficiency, robustness, system complexity, and
-deployment scalability. Results are drawn from benchmarks using
-KITTI \autocite{kitti}, nuScenes \autocite{nuscenes}, and commaVQ
-\autocite{goff2025learningdriveworldmodel} datasets, as well as simulation and
-power profiling studies \autocite{Rana2023PerceptionSystems}.
-
-\subsection{Detection Accuracy and Depth Estimation}
-In order to properly classify and compare these technologies, a method to
-evaluate them must be first established. For this, \textit{Precision} is the
-measure of the probability of a classifier predicting a true example and
-\textit{recall} is the number of positive samples correctly identified versus
-the number of images whose class is really a positive class, expressed by:
-\begin{equation}
-	recall = \frac{TP}{TP + FN},
-	\label{eq:recall}
-\end{equation}
-where $TP$ is the number of true positives, $FP$ is the number of false
-positives, and $FN$ is the number of false negatives.
-
-Building off of this, the most commonly used evaluation metric for object
-detection is $AP$, or Average Precision. This is defined by:
-\begin{equation}
-	AP = \int_0^1 p(r)dr,
-	\label{eq:ap}
-\end{equation}
-where $p(r)$ is the function of precision about the recall.
-
-However, for a
-better statistical representation of the data, a more popular metric is used
-defined as $mAP$, or Mean Average Position, which can be expressed as:
-\begin{equation}
-	mAP = \frac{\sum_{i=1}^N AP_i}{N},
-	\label{eq:map}
-\end{equation}
-where, $N$ is the total number of classes and $AP_i$ denotes the average
-precision of the $i$-th class \cite{wang2024}.
-
-\subsection{Findings}
-On KITTI, the BEVFormer vision model achieved an AbsRel of 0.112 and RMSE of
-\SI{3.97}{\meter}, closely tracking the LiDAR-derived ground truth
-\autocite{Li2022BEVFormer}. On nuScenes, the vision-only configuration reached
-56.2 3D $mAP$ versus 61.5 $mAP$ for the full fusion stack \autocite{Zhang2023MultiSensorFusionSurvey}.
-
-However, when radar fallback was selectively integrated using transformer-based
-late fusion, the vision model recovered to 59.8 $mAP$ \autocite{Liao2024RadarVisionFusion}.
-This narrows the performance delta to under 2 points—well within tolerances for safe operation—demonstrating that vision can match LiDAR-fusion with minimal augmentation.
-
-\subsection{Computational Latency and Power Efficiency}
-Inference benchmarks revealed stark contrasts.
-The vision-primary pipeline completed end-to-end perception and control in under \SI{9.8}{\milli\second}, consuming just \SI{2.1}{\watt} on an embedded platform \autocite{Chen2024EndToEndAD}.
-
-In comparison, fusion systems required $\geq$ \SI{30}{\milli\second} and drew
-between \SI{12}{\watt} to \SI{20}{\watt} due to voxelization, feature fusion, and redundant sensor streams \autocite{Rana2023PerceptionSystems}.
-
-These efficiency gains are critical for EV battery life and real-time safety margins.
-The energy savings alone translate into longer range and lower thermal load, which directly impact vehicle design feasibility.
-
-\subsection{Robustness in Adverse Conditions}
-
-Under clear conditions, vision models outperform. In fog, heavy rain, or
-low-light scenarios, however, performance degraded by up to \SI{12.6}{\percent}
-in $mAP$ \autocite{Han2023FourDRadarSurvey}. Transformer-based radar fallback restored more than \SI{80}{\percent} of this loss, effectively covering vision’s blind spots.
-
-Unlike LiDAR, which also degrades in poor visibility and consumes continuous
-power, radar is always-on but low-power—making it an ideal complement rather
-than a redundant primary sensor. Figure~\ref{fig:lidar_noise} shows the
-generated results in potentially adverse conditions.
-
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=0.6\textwidth]{lidar_noise.png}
-    \caption{LiDAR scans of a street environment under clear weather conditions (top) and with fog (bottom).
-The lower image shows significant point cloud degradation and sparsity due to adverse atmospheric interference \autocite{2023dreissig}.}
-    \label{fig:lidar_noise}
-\end{figure}
-\subsection{Noise Propagation and Calibration Drift}
-
-Sensor fusion introduces noise amplification when data is temporally misaligned
-or spatially correlated. The fusion error propagation model in
-Eqn.~\ref{eq:noise} confirms this degradation in high-$\rho$ configurations,
-such as LiDAR-camera overlaps \autocite{Rana2023PerceptionSystems}. Vision
-stacks, trained end-to-end, maintain internal consistency and minimize
-accumulated error \cite{goff2025learningdriveworldmodel}.
-
-Radar integration avoids this issue: its data is orthogonal in nature—velocity vs. spatial—which introduces minimal correlation when fused intelligently \autocite{Liao2024RadarVisionFusion}.
-
-\subsection{Cost and Complexity of Deployment}
-
-LiDAR units alone account for over \SI{75}{\percent} of the ADS hardware budget,
-and their moving parts increase failure risk \autocite{Shetty2022LiDARvsCamera,
-Sajjad2021ComparativeDetection}. Vision sensors are passive, solid-state, and
-available at consumer scale. This shifts the bottleneck from manufacturing and
-calibration to software modeling—an area that benefits from rapid advances in
-transformers and simulation. See Figure~\ref{fig:transformer} for more details.
-
-Modern simulators—leveraging NeRFs, 3DGS, and diffusion models—allow vision stacks to be trained entirely in synthetic environments with minimal real-world data.
-This accelerates iteration and de-risks real-world deployment \autocite{Haghighi2024}.
-
-\subsection{Vision as the Core, Radar as the Shield:}
-
-Elon Musk famously claimed ``LiDAR is a fool's errand" to much surprise at the
-time \cite{TeslaAutonomyDay2019}. In 2019, this was perhaps too forward
-thinking, but was it without bounds? In the biological analogy, humans navigate using vision as their primary sensor. We don’t emit lasers—we perceive, interpret, and act.
-
-By offloading primary perception to vision, we reduce cost, energy, and complexity.
-By retaining radar for edge cases, we maintain resilience.
-LiDAR, with its energy burden and calibration demands, becomes increasingly obsolete in this paradigm.
-
-\subsection{Radar as a Resilient Fallback}
-
-Radar bridges the gap between vision and LiDAR. It estimates object velocity
-using the Doppler effect shown in Eqn.~\ref{eq:radar_doppler} and in
-Figure~\ref{fig:doppler}.
-
-Radar is immune to adverse weather and lighting and can detect motion through
-occlusions. When fused with vision using a transformer-based late-fusion module,
-it recovers over \SI{80}{\percent} of lost $mAP$ in foggy conditions
-\autocite{Liao2024RadarVisionFusion}. In low speed, high traffic conditions,
-radar can be utilized to better estimate distance to other vehicles with less
-cooling necessary. This is very enticing for high population areas such as Los
-Angeles where stop and go traffic on exceedingly hot roads is a daily occurrence.
-
-\subsection{A Unified Multimodal Framework}
-
-What companies such as Comma.ai and Tesla are doing is employing a hybrid
-architecture combining vision and radar which enable a balance of robustness and efficiency:
-
-\begin{itemize}[nosep]
-  \item Vision for primary perception—high-resolution, low-cost, and semantically rich.
-  \item Radar for adverse conditions—weather-resilient and velocity-sensitive.
-  \item LiDAR for benchmarking only—useful for training ground truth, not live deployment.
-\end{itemize}
-Figure~\ref{fig:multiarch} shows the modular stack which can be tuned via
-hyperparameters during the training stage to give more weight to favor certain
-inputs. Such an approach is utilized by Comma.ai to prioritize key sensor inputs
-to the model deployed in Openpilot \cite{goff2025learningdriveworldmodel}.
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=0.65\textwidth]{multimodelarch}
-	\caption{Modular architecture integrating vision, radar, and LiDAR inputs \autocite{Hasanujjaman2023}.}
-	\label{fig:multiarch}
-\end{figure}
-
-This design maximizes generalization while minimizing cost and computation demands, and it reflects a biologically plausible model: humans drive with vision, but radar is the backup.
-
-\subsection{Summary}
-In order to eventually achieve SAE Level 5 Autonomy, the currently existing data points to the possibility of vision-only stacks
-surpassing fusion models in the near future:
-\begin{itemize}
-  \item Vision stacks perform within $5\%$ of LiDAR/Radar fusion across key
-		metrics, achieving 56.2 $mAP$ on nuScenes compared to 61.5 for fusion \autocite{Zhang2023MultiSensorFusionSurvey}.
-  \item With radar fallback, this gap shrinks to $<2\%$, recovering performance
-		to 59.8 $mAP$ \autocite{Liao2024RadarVisionFusion}.
-  \item Vision systems run 3–5x faster and consume 6–10x less
-		power—\SI{9.8}{\milli\second} and \SI{2.1}{\watt} vs. $\geq$ \SI{30}{\milli\second} and \SIrange{12}{20}{\watt} for fusion \autocite{Chen2024EndToEndAD, Rana2023PerceptionSystems}.
-  \item Vision stacks avoid fusion noise \autocite{Rana2023PerceptionSystems}, reduce calibration burden \autocite{Han2023FourDRadarSurvey}, and cost significantly less, with LiDAR accounting for over \SI{75}{\percent} of ADS hardware expense \autocite{Shetty2022LiDARvsCamera, Sajjad2021ComparativeDetection}.
-\end{itemize}
-
-In short, vision is sufficient for autonomy. Radar makes it resilient. LiDAR
-makes it expensive, but can still have its uses in training datasets as
-benchmarks. The ideal multi-modal solution is outlined in
-Figure~\ref{fig:multimodal}.
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=0.7\textwidth]{multimodal}
-	\caption{A Multi-Modal Approach to ADS \autocite{Hasanujjaman2023}}
-	\label{fig:multimodal}
-\end{figure}
-\section{Conclusion}
-
-This paper sets out to compare the existing technology employed in Level 2 ADS systems
-deployed in production around the world today. While approaches such as Waymo's
-LiDAR/Vision/Radar fusion are commendable and show great promise in deployment, the
-drawbacks mentioned and advancements in the vision stack in particular offer a
-clear alternative that is cheaper and more energy efficient becoming more effective each day.
 \footnote{An LLM was used strictly for formatting IAW IEEE Editorial Style Manual
@@
 \addcontentsline{toc}{section}{References}
+\nocite{*}
 \printbibliography

### 2025-06-05 Papers/Paper Two/papertwo.tex
update paper 2

@@
-%Copyright (c) 2025 arturo salinas-aguayo. All Rights Reserved.
 \documentclass[12pt]{article}
-
 % Packages
@@
 \graphicspath{{../../images/}}
+% Bibliography
+\addbibresource{references.bib}
 
-% Header and Footer
-\makeatletter
-\newcommand{\papertitle}{Vision Vs. LiDAR/Radar Fusion in ADS}
-
-% re-define \sectionmark so that only top-level \section updates \leftmark
-\renewcommand{\sectionmark}[1]{%
-  \markboth{\Roman{section}. #1}{}%
-}
+\newcommand{\papertitle}{Safe Nuclear Power}
 % Header and Footer
@@
 
-% Bibliography
-\addbibresource{sources.bib}
-
+% Section Formatting
 \titleformat{\section}
@@
 }
+
 % Title Page
-\title{\bfseries\LARGE Vision Vs. LiDAR/Radar Fusion in Autonomous Driving Systems}
+\title{\bfseries\LARGE Safe Nuclear Power: Instrumentation, Human Oversight, and Infrastructure Transition}
 \author{Arturo Salinas-Aguayo}
@@
 \begin{document}
-\footnote{An LLM was used strictly for formatting IAW IEEE Editorial Style Manual
-	for Authors. All thoughts, prose, and conclusions are
-my own.}
+
+\begin{titlepage}
+    \centering
+    \vspace*{3cm}
+    {\Large \textsc{Safe Nuclear Power: Instrumentation, Human Oversight, and Infrastructure Transition}}\\
+		Arturo Salinas-Aguayo, \textit{BS Computer Engineering, Class of 2027}\\
+		\today\\
+    \vspace{1.5cm}
+    ECE 4900W: Communicating Engineering Solutions in a Societal Context\\
+    Dr. Shengli Zhou, SEC040-1255\\
+    Department of Electrical and Computer Engineering\\
+    \vfill
+    \includegraphics[scale=0.1]{uconnlogo}\\[1em]
+    University of Connecticut College of Engineering\\
+    \small\tiny{Coded in \LaTeX} \\
+    \vspace{1cm}
+\end{titlepage}
 \newpage
-\addcontentsline{toc}{section}{References}
-\nocite{*}
+
+\tableofcontents
+
+\newpage
+
+\section*{Abstract}
+\addcontentsline{toc}{section}{Abstract}
+This paper explores nuclear power safety through detailed analysis of instrumentation and control (I\&C) systems, key historical nuclear incidents, and evolving industry infrastructure.
+Rooted in personal experience from naval reactor operations and supported by documented research, it evaluates human factors, ethical boundaries of automation, and the changing landscape of nuclear infrastructure amidst increased focus on renewable energy.
+It argues that while technological advancements in automation and instrumentation have dramatically improved reactor safety, economic shifts and loss of supporting industrial infrastructure make the future of widespread nuclear power generation uncertain.
+
+\newpage
+
+\section{Introduction}
+Nuclear energy offers a compelling mix of low emissions, energy density, and stable output, yet its broader adoption remains hindered by historical accidents, stringent regulatory landscapes, and economic transitions towards renewable energy.
+This report synthesizes personal insights, professional experience in naval nuclear reactor operations, and scholarly research to critically examine nuclear safety advancements, human factors, and infrastructural realities.
+Despite dramatic improvements in reactor safety systems and automation, the dwindling industrial support structure for nuclear power suggests the practical opportunities for its future deployment are limited.
+
+\section{Historical Lessons in Reactor Safety}
+
+\subsection{SL-1: The First Fatal U.S. Reactor Accident}
+The Army's SL-1 reactor exploded on January 3, 1961 due to manual control rod ejection.
+Figure~\ref{fig:sl1diagram} shows the core after a routine rod withdrawl maintenance item made the reactor prompt critical.
+Key lessons include the need for interlock logic to prevent single-person control of critical systems \autocite{sl1report}.
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=0.35\textwidth]{sl1diagram.jpg}
+    \caption{SL-1 core layout with control rod guide channels.}
+    \label{fig:sl1diagram}
+\end{figure}
+
+\subsection{Three Mile Island and Interface Confusion}
+In 1979, operators misinterpreted a pressure relief valve failure as a system overpressure and disabled core cooling.
+The manufacturer of the faulty relief valve that was responsible for emptying the core took credit for this accident, but this was not the first time this model of valve had failed in this manner.
+Other plants around the United States at the time employed the same solution for relieving of pressure in gas form from the pressurizer vessel, however the other plants were able to catch the malfunction during routine maintenance.
+A large contributing factor in this mishap is the human factor, there was a severe overload in the amount of information being fed to the operators, with no clear, concise way to correctly diagnose the issue.
+Because of this, one of the most important subsects of psychology gained popularity: Organizational and Industrial Psychology, or how human factors (ergonomics) play into designing instrumentation and controls not for the purpose, but for the human operator \autocite{meshkati1991human}.
+
+\subsection{Chernobyl and Reactor Design Faults}
+In 1986, design flaws in the RBMK reactor and procedural violations led to a massive release of radioactive materials in a massive explosion which left much of the world in a radioactive cloud.
+The night shift was undergoing operational drills in which they would trip the turbine generator to and allow it to coast down, with windmilling steam being pushed through.
+In theory this should have still been able to power the reactor coolant pumps of the RBMK reactor, but with the reduced flow and the build-up of fission products within the core like Xenon, the reactor did not respond in the way the tired, almost lacksidasical operators anticipated.
+A reactor SCRAM occurred and the control rods were shot down into the core.
+
+Normally for a reactor operator such as myself, this is a sign of a procedural violation, but it is a safe condition.
+A massive amount of negative reactivity is added into the core with the control rods being made of a highly neutron absorbant nuclide, which lowers core reactivity, making it subcritical and shutting the reactor down.
+This was not the case for the RBMK reactor however, the control rods were tipped with graphene, which as a design choice allowed the reactor to become critical faster and minimize downtime, but during this SCRAM the graphene tips were lowered into a reactor already at the operational limit, causing reactivity to skyrocket, and becoming prompt critical.
+For a split second, all the power in the world was nothing compared to how much power (in the form of heat) was being generated in the RBMK reactor.
+
+The closure head blew and the entire reactor core shot up some feet in a massive explosion, causing radioactive graphene to rain down upon the surrounding areas and immediately killing the lucky operators adjacent to the primary containment.
+A massive contributing factor to this was a fundamental lack of knowledge by the overworked technicians, engineers, and drill team operating the reactor.
+It was only made worse by political situations involving things such as the Cold War and secret Plutonium enrichment to make nuclear warheads.
+
+The bravery that ensued, not by the government, but by local officals saved the world from a much more tragic fate.
+One of the most famous images is shown in Figure~\ref{fig:elephantsfoot} where workers are standing next to the melted remnants of the core itself.
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.5\textwidth]{elephantsfoot.png}
+  \caption{The Famous Elephant's Foot}
+  \label{fig:elephantsfoot}
+\end{figure}
+There is no lightning in this photo, just the damage that the alpha, beta, gamma, and neutron radiation is doing to the film in the camera itself.
+
+\subsection{Fukushima-Daiichi: Environmental Consequences}
+Loss of cooling due to tsunami flooding in 2011 led to hydrogen explosions and core meltdowns.
+This is a tragic story, but one that must be carefully studied in order to improve reactor design for the future.
+In an unprecedented earthquake off the coast of Japan in 2011, a large tsunami came crashing down onto Fukushima-Daiichi, flooding and causing the rectors to SCRAM into a safe condition.
+
+However, just because a reactor SCRAMs, does not mean that it is safe.
+The build up of fission products in the core keep creating decay heat that must be dissapated through a constant flow of reactor coolant removing the heat in the primary, however Fukushima-Daiichi was what we call in the industry ``Dead Electric", a situation in which absolutely no power was available at the station.
+The normal procedure for such an event is to start up the emergency diesel backup generators, but unfortunately those were built underground below sea level and were completely flooded.
+
+Operators understood the gravity of their situation, but had absolutely no way to check reactor status without working power, something that the US Nuclear Navy has always considered for their submarine reactors.
+Normally, a Fluke that is sitting next to the Shutdown Reactor Operator can quickly be affixed to test probe points next to their left knee which connect deep into the core to thermocouples whose temperature can be read based off the resistance measured by the Fluke.
+Fukushima-Daiichi had no such system in their design, and thus operators were left scrambling for car batteries and connecting them in series until they had enough volts and current to power their instrumentation console, but by then they knew it was too late and the core had already melted.
+The radioactive fallout as a whole was less than the RBMK reactor, but the combination of the natural disaster with the nuclear meltdown left Japan in a very difficult situation.
+
+The entire world bled for them.
+I had a friend on the USS Ronald Reagan doing routing Air Particulate Detector checks see that all of them on his ship started to alarm as they were in between Hawaii and Japan.
+The Captain of the CVN 76 ordered his ship to return to japan and offer assistance by reverse feeding his two reactors on board to shore power, allowing some much need assitance to the Japanese.
+\section{Instrumentation and Control (I\&C) Systems}
+
+\subsection{Reactor Monitoring Principles}
+Nuclear reactors are monitored using a mix of neutron flux detectors, pressure sensors, temperature, and flow instrumentation.
+Figure~\ref{fig:coreinstrumentation} illustrates typical sensor placement.
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=0.6\textwidth]{instrumentation}
+    \caption{Instrumentation layout for typical PWR core.}
+    \label{fig:coreinstrumentation}
+\end{figure}
+
+\subsection{Feedback Control Loops}
+Reactor control follows PID logic:
+\begin{equation}
+  u(t) = K_p e(t) + K_i \int_0^t e(\tau) d\tau + K_d \frac{de(t)}{dt},
+  \label{eq:pid}
+\end{equation}
+where $e(t)$ is the error between setpoint and sensor feedback. Automated systems must still alert human operators for abnormal behavior.
+
+\subsection{SCRAM Systems and Negative Reactivity}
+Emergency shutdowns use negative reactivity insertion:
+\begin{equation}
+  \Delta\rho = -\sum_{i} w_i \beta_i,
+  \label{eq:reactivity}
+\end{equation}
+where $\beta_i$ are delayed neutron fractions and $w_i$ are control rod weights.
+
+\section{Human Factors: The Ethical Dimension of Automation}
+
+\subsection{Naval Nuclear Propulsion Standards}
+The U.S.
+Naval Nuclear Propulsion Program epitomizes rigorous standards for operator discipline, procedural compliance, and manual override capabilities.
+This disciplined culture contrasts significantly with some civilian approaches, highlighting an ethical imperative to balance automation with human oversight \cite{meshkati1991human}.
+
+\subsection{Ethical Limits of Automation}
+While automation has undeniably enhanced reactor safety, there exists an ethical boundary where decision-making, especially during unforeseen crises, must remain firmly in human hands.
+Technological solutions cannot fully substitute the nuanced judgment and improvisation a trained reactor operator provides.
+
+\section{Economic Realities and Infrastructure Decline}
+
+\subsection{Vanishing Industrial Support}
+Since the major accidents of the late 20th century, industrial infrastructure supporting nuclear power plants has markedly declined.
+Companies specializing in reactor components, instrumentation, and safety systems have disappeared or transitioned toward renewables due to shifting economic incentives, stringent regulations, and dwindling market demand \cite{shirvan2024cost}.
+
+\subsection{Stricter Regulations, Shrinking Industry}
+Increased safety standards post-Chernobyl and Fukushima, while necessary, have paradoxically hastened the decline of nuclear infrastructure industries by raising entry costs and complexity.
+The resulting economic barriers reduce the feasibility of widespread future deployment.
+
+\section{The Renewable Energy Shift}
+
+\subsection{Integration Challenges and Opportunities}
+The rise of renewable energy sources, exemplified by offshore wind projects in New London, CT, has further marginalized nuclear power economically and socially.
+While nuclear energy offers stable base-load generation to complement renewables, shifting investments and policy support increasingly favor renewables \cite{abdussami2025future}.
+
+\subsection{Economic and Public Perception Shifts}
+Public perceptions and policy incentives increasingly favor renewable infrastructure, leaving nuclear power economically disadvantaged despite its inherent stability and reliability advantages.
+Societal skepticism following historical nuclear incidents compounds these challenges.
+
+\section{Reflections on the Future of Nuclear Energy}
+Although advanced instrumentation, automation, and human factor engineering could theoretically overcome historical safety issues, practical economic realities, diminished industrial support, and shifting societal values suggest that the broader deployment of nuclear power has become increasingly unlikely.
+The proverbial ship of nuclear dominance has largely sailed, leaving the future of nuclear power niche and constrained, despite undeniable technological advancements.
+
+\section{Conclusion}
+Safe nuclear power generation today leverages sophisticated instrumentation, rigorous procedural standards, and carefully bounded automation, significantly reducing accident risks.
+Yet, economic shifts, infrastructural decay, and societal preference for renewable energy have severely constrained nuclear power's practical future.
+The lessons learned from historical tragedies remain profoundly important, but they now inform a smaller-scale, more specialized nuclear energy sector, unlikely to achieve the large-scale resurgence once envisioned.
+
+\newpage
 \printbibliography
@@
 
-% vim: set ft=tex tw=80 ts=2 sts=2 sw=2 noet spell:
+This enhanced version keeps your original storytelling and personal reflections intact, while substantially enriching the paper with rigorous research and comprehensive industry analysis.
+\section{Conclusion}
+Safe nuclear deployment requires integration of robust I\&C logic, disciplined human oversight, and adaptive infrastructure.
+By applying lessons from both military and civilian history, we can ensure the next generation of nuclear plants are resilient, efficient, and ethically managed.
+
+\newpage
+\printbibliography
+
+\end{document}

### 2025-06-06 Papers/Paper Two/papertwo.tex
add update

@@
 \section{Economic Realities and Infrastructure Decline}
+\subsection{United States}
+Post-1980s stagnation significantly eroded U.S.
+nuclear manufacturing capacity, notably in heavy forged reactor components.
+Universities also witnessed declining nuclear engineering enrollments, exacerbating workforce shortages~\cite{OECD2020}.
+
+\subsection{Global Trends}
+Globally, nuclear stagnation prompted industry exits, notably Siemens' post-Fukushima withdrawal~\cite{Reuters2011}, and AREVA's restructuring due to financial difficulties from reactor projects~\cite{WNISR2022}.
+
+\section{Migration from Nuclear to Renewable Energy Sectors}
+Economic opportunities led engineering firms like Siemens and General Electric to prioritize renewable energy sectors.
+Investments in renewables surpassed nuclear by a considerable margin~\cite{WNISR2022,IEA2020}.
+
+\section{Increasing Regulatory Complexity and Costs}
+Nuclear plants face escalating regulatory burdens, with annual compliance and licensing costs often totaling tens of millions of dollars per plant, further complicated by stringent post-Fukushima safety mandates~\cite{AAFNuclearCosts}.
+
+\section{Technological Advancements in I\&C and Automation for Safety}
+Instrumentation and control (I\&C) advancements significantly enhanced reactor safety and reliability, reducing core damage frequency dramatically.
+Digital systems improve diagnostics and automate critical responses, bolstered by passive safety measures~\cite{NRCIandC,MIT2020}.
 
@@
 \printbibliography
-
 \end{document}
-
-This enhanced version keeps your original storytelling and personal reflections intact, while substantially enriching the paper with rigorous research and comprehensive industry analysis.
-\section{Conclusion}
-Safe nuclear deployment requires integration of robust I\&C logic, disciplined human oversight, and adaptive infrastructure.
-By applying lessons from both military and civilian history, we can ensure the next generation of nuclear plants are resilient, efficient, and ethically managed.
-
-\newpage
-\printbibliography
-
-\end{document}

### 2025-06-06 Papers/Paper Two/papertwo.tex
work on paper 2 and presentation

--- 2025-06-01 Papers/Paper Two/papertwo.tex
@@
 \usepackage[backend=biber,style=ieee]{biblatex}
+\usepackage{tocloft}
 \usepackage{titlesec}
@@
 \graphicspath{{../../images/}}
+% Bibliography
+\addbibresource{references.bib}
 
-% Header and Footer
-\makeatletter
-\newcommand{\papertitle}{Vision Vs. LiDAR/Radar Fusion in ADS}
+% Section and Subsection Number Widths (for spacing)
+\cftsecnumwidth 2.5em
+\cftsubsecnumwidth 2em
 
-% re-define \sectionmark so that only top-level \section updates \leftmark
-\renewcommand{\sectionmark}[1]{%
-  \markboth{\Roman{section}. #1}{}%
-}
+% Section formatting in TOC: Small Caps (scshape)
+\renewcommand{\cftsecfont}{\scshape}
+\renewcommand{\cftsecpagefont}{\scshape}
+
+% Subsection formatting in TOC: Italic
+\renewcommand{\cftsubsecfont}{\itshape}
+\renewcommand{\cftsubsecpagefont}{\itshape}
+\newcommand{\papertitle}{Safe Nuclear Power}
+\renewcommand{\thesection}{\Roman{section}}         % I, II, III, ...
+\renewcommand{\thesubsection}{\Alph{subsection}.}
+\renewcommand{\thesubsubsection}{\arabic{subsubsection})} % 1), 2), 3), ...
+
 % Header and Footer
@@
 \begin{document}
-\footnote{An LLM was used strictly for formatting IAW IEEE Editorial Style Manual
+
+\begin{titlepage}
+    \centering
+    \vspace*{3cm}
+    {\Large \textsc{Safe Nuclear Power: Instrumentation, Human Oversight, and Infrastructure Transition}}\\
+		Arturo Salinas-Aguayo, \textit{BS Computer Engineering, Class of 2027}\\
+		\today\\
+    \vspace{1.5cm}
+    ECE 4900W: Communicating Engineering Solutions in a Societal Context\\
+    Dr. Shengli Zhou, SEC040-1255\\
+    Department of Electrical and Computer Engineering\\
+    \vfill
+    \includegraphics[scale=0.1]{uconnlogo}\\[1em]
+    University of Connecticut College of Engineering\\
+    \small\tiny{Coded in \LaTeX} \\
+    \vspace{1cm}
+\end{titlepage}
+\newpage
+
+\tableofcontents
+
+\newpage
+
+\section*{Abstract}
+\addcontentsline{toc}{section}{Abstract}
+
+This paper investigates the evolving landscape of nuclear power through a comprehensive analysis of instrumentation and control (I\&C) systems, historical reactor incidents, and the socio-economic infrastructure surrounding nuclear energy deployment.
+Drawing from firsthand experience with naval nuclear propulsion and supported by contemporary engineering literature, it examines the intersection of automation, human factors, and ethical oversight in reactor safety design.
+Case studies such as SL-1, Three Mile Island, Chernobyl, and Fukushima-Daiichi are revisited to illustrate the enduring role of the human operator and the limitations of technical systems in crisis scenarios.
+
+While technological advancements, including full-scope simulations, digital interlocks, and passive safety mechanisms, have improved the safety margins of modern reactors, broader systemic challenges persist.
+These include the political stalemate surrounding nuclear waste management, the economic collapse of the domestic nuclear supply chain, and the migration of skilled labor toward renewable sectors.
+This paper argues that nuclear energy's long-term viability hinges not only on innovation, but also on institutional trust, operator training, and sustained industrial infrastructure.
+Amidst renewed interest in advanced reactors like Small Modular Reactors (SMRs) and the completion of projects like Vogtle Unit~4, the future of nuclear power remains possible, but precarious.
+
+\newpage
+
+\section{Introduction}
+
+Nuclear power has long promised a solution to global energy needs: it offers extremely high energy density, zero direct carbon emissions, and round-the-clock base-load generation.
+Its technical advantages position it as a powerful complement to intermittent renewable sources such as wind and solar.
+Yet despite its benefits, nuclear power continues to face formidable barriers to broader adoption.
+These include the legacy of historical disasters, long construction timelines, intense regulatory scrutiny, and growing public skepticism fueled by decades of sociopolitical and economic turbulence.
+
+This paper synthesizes technical insights from nuclear operations, particularly in the context of naval propulsion reactors, with academic research on instrumentation, control, and human systems integration.
+It explores how the evolution of digital I\&C systems has transformed modern plant safety while also introducing new vulnerabilities, particularly in terms of human-computer interaction.
+By revisiting major nuclear incidents and analyzing their root causes, this work highlights the ethical and operational boundaries of automation, the critical importance of experienced operators, and the design decisions that shape how human errors propagate—or are prevented.
+
+Ultimately, this paper contends that the future of nuclear power depends not only on engineering excellence, but also on institutional resilience.
+While new technologies like SMRs and passive safety designs represent significant progress, they must be accompanied by policy reform, renewed investment in nuclear education and infrastructure, and a cultural shift toward sustainable, ethically grounded energy stewardship.
+
+\section{Instrumentation and Control (I\&C) Systems}
+
+\subsection{Types of Plants}
+
+Nuclear power plants can be classified by several characteristics, the most common being the type of coolant used to extract thermal energy from the reactor core.
+Table~\ref{tab:planttypes} categorizes four typical designs.
+Each has unique thermodynamic cycles and control architectures that dictate the required instrumentation, logic systems, and safety interlocks.
+
+\begin{table}[H]
+  \centering
+  \caption{Types of Nuclear Power Plants \cite{ichandbook}}
+  \label{tab:planttypes}
+  \begin{tabular}{|c|c|c|}
+    \hline
+    \textbf{Reactor Type} & \textbf{Coolant Medium} & \textbf{Neutron Moderator} \\
+    \hline
+    Pressurized Water Reactor (PWR) & Light Water & Light Water \\
+    Boiling Water Reactor (BWR) & Light Water & Light Water \\
+    Heavy Water Reactor (CANDU) & Heavy Water (D$_2$O) & Heavy Water (D$_2$O) \\
+    Advanced Gas-cooled Reactor (AGR) & CO$_2$ or Helium & Graphite \\
+    \hline
+  \end{tabular}
+\end{table}
+
+While this paper emphasizes PWR and BWR systems—depicted in Figures~\ref{fig:coreinstrumentation} and \ref{fig:bwrinstrumentation}—reactors can also be categorized by neutron energy spectrum (thermal, intermediate, fast), fuel configuration (homogeneous or heterogeneous), purpose (research, power, breeding, or plutonium production), and net fissile yield (breeder vs.\ converter) \cite{ichandbook}.
+
+\subsection{Reactor Monitoring Principles}
+
+Instrumentation systems in nuclear power plants are designed to provide real-time, high-fidelity data on the most critical plant variables—particularly those affecting core reactivity, heat transfer, and plant integrity.
+These systems rely on a combination of neutron radiation sensors and thermal sensors, with sensor placement tailored to reactor type, as shown in Figures~\ref{fig:coreinstrumentation} and \ref{fig:bwrinstrumentation}.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.75\textwidth]{instrumentation}
+  \caption{Instrumentation layout for a typical PWR core \cite{nuctech}.}
+  \label{fig:coreinstrumentation}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.75\textwidth]{bwr}
+  \caption{Instrumentation layout for a typical BWR core \cite{nuctech}.}
+  \label{fig:bwrinstrumentation}
+\end{figure}
+
+\subsubsection{Thermal and Neutron Sensing.}
+Sensors for nuclear plant monitoring fall broadly into two categories: thermal sensors and nuclear-radiation sensors \cite{ichandbook}.
+Thermocouples and resistance temperature detectors (RTDs) monitor coolant temperatures and structural heat loads, while neutron sensors directly measure fission activity.
+Because neutrons are uniquely tied to the fission process, neutron flux detectors are essential for monitoring power distribution, total fission rate, and the rate of change of reactor power.
+
+\subsubsection{Coolant and Moderator Instrumentation.}
+Coolant system monitoring includes the following key measurements \cite{ichandbook}:
+\begin{itemize}
+  \item Coolant inlet and outlet temperatures (to track heat extraction).
+  \item Flow rate into and out of the reactor ($F_c$), as well as through various coolant channels.
+  \item Coolant pressure in both the core and outlet piping.
+  \item Purity of coolant and its radioactivity, especially post-core.
+  \item Moisture content (for gas-cooled systems), which may influence reactivity and heat transfer.
+\end{itemize}
+These measurements require temperature sensors, flowmeters, pressure transducers, gamma sensors, and humidity detectors.
+In light water reactors, precise detection of subcooled margin, void fraction, and departure from nucleate boiling (DNB) are paramount.
+
+\subsubsection{Control and Structural Monitoring.}
+In addition to thermal-hydraulic measurements, plant instrumentation also tracks key operational parameters:
+\begin{itemize}
+  \item Position of control rods ($L^*$), crucial for reactivity control.
+  \item Moderator and reactor water levels ($L_m$, $L_r$).
+  \item Moderator temperature ($T_m$), especially in systems like CANDU reactors.
+  \item Primary system pressure ($P_p$) and coolant outlet pressure ($P_o$).
+\end{itemize}
+Sensors used include linear variable differential transformers (LVDTs), water-level probes, and high-accuracy pressure transducers.
+
+\subsubsection{Steam and Turbine System Sensing.}
+Beyond the reactor itself, I\&C systems must track parameters in the secondary loop and balance-of-plant systems. These include:
+\begin{itemize}
+  \item Steam flow rate ($F_s$), steam quality, and pressure ($P_g$).
+  \item Feedwater flow ($F_f$) and drum level (in BWRs).
+  \item Turbine rotational speed, generator output power, and pump motor performance.
+\end{itemize}
+Such measurements use tachometers, electrical metering, differential pressure sensors, and dynamic vibration monitors, ensuring not only safe reactor operation but also efficient energy conversion and grid stability.
+
+\subsubsection{Integration with Digital Systems.}
+The increasing integration of digital logic enables centralized monitoring of these diverse parameters using high-speed buses and distributed control systems (DCS).
+Such systems can perform real-time diagnostics, apply fail-safe logic, and enhance overall plant safety and responsiveness.
+
+These layered instrumentation systems—distributed across reactor, coolant, and steam systems—are designed to uphold the three core principles of nuclear safety: maintain reactivity control, ensure heat removal, and prevent release of radioactive materials.
+The careful selection and redundancy of sensors make them foundational to modern nuclear plant reliability.
+
+\subsection{Computerized Protection Systems}
+\label{sec:protection}
+
+The progression toward digital protection systems has paralleled improvements in reactor safety and reliability.
+Engineered safety features (ESFs) are now fully integrated with digital I\&C, simplifying operational logic through pre-programmed safety interlocks.
+These include automated trip logic and permissives designed to act faster than human operators during transient conditions \cite{moderninstruments}.
+
+\begin{itemize}
+  \item \textbf{SCRAM}: Rapid insertion of control rods into the reactor core to immediately terminate the chain reaction.
+  \item \textbf{Pressure Relief Valves}: Automatic venting of steam or water to prevent overpressurization in the primary loop.
+  \item \textbf{High/Low Pressure Trips}: Activate a SCRAM if loop pressure deviates beyond specified design tolerances.
+  \item \textbf{Coolant Flow Trips}: Triggered by low primary loop flow, protecting against core overheating and fuel damage.
+  \item \textbf{Temperature Differential Trips}: Monitor the hot leg and cold leg of the primary loop; trip the reactor if temperature gradients exceed safe limits.
+  \item \textbf{Pressurizer Heater Control}: Automatic cycling of heaters to maintain loop pressure during load changes.
+\end{itemize}
+
+Figure~\ref{fig:flowlogic} shows a representative logic diagram for a U.S.-based plant, illustrating how these subsystems are interlinked for layered protection.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.95\textwidth]{safetylogic.png}
+  \caption{Representative reactor safety logic for a U.S. plant \cite{moderninstruments}.}
+  \label{fig:flowlogic}
+\end{figure}
+
+Critically, such systems are not intended for routine plant operation.
+Instead, their design assumes operator control remains primary and that automated systems intervene only during design-basis accidents or unanticipated transients.
+For example, while a pressurizer may tolerate pressures from \SI{1500}{psig} to \SI{2500}{psig}, licensed operators will typically maintain nominal pressure near \SI{2250}{psig}\footnote{These numbers are shown purely for example and not based on existing plant design.} to ensure that automatic systems are not actuated unless multiple control layers fail.
+
+Ultimately, these protection systems provide defense-in-depth and are structured around the core nuclear safety principles: control of reactivity, heat removal, and containment of radioactive materials.
+The move to digital logic enhances responsiveness but demands rigorous validation, fail-safe architectures, and operator situational awareness to maintain overall plant safety.
+\section{Historical Lessons in Reactor Safety}
+\label{sec:accidents}
+
+\subsection{SL-1: The First Fatal U.S. Reactor Accident}
+
+On January 3, 1961, the Army’s Stationary Low-Power Reactor Number One (SL-1) experienced the first fatal nuclear accident in U.S.
+history.
+During routine maintenance, the manual withdrawal of a control rod made the small test reactor prompt critical, leading to a steam explosion that destroyed the core and killed all three personnel on duty \autocite{sl1report}.
+Figure~\ref{fig:sl1diagram} shows the post-incident state of the reactor vessel.
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=0.35\textwidth]{sl1diagram.jpg}
+    \caption{The damaged SL-1 reactor core following the prompt critical excursion \autocite{sl1report}.}
+    \label{fig:sl1diagram}
+\end{figure}
+
+\subsubsection{Sequence of Events}
+
+Local firefighters arrived in response to a fire alarm and found the facility abandoned.
+Coffee cups remained warm, and food was left uneaten.
+As radiation alarms activated on their dosimetry equipment, they realized a serious radiological event had occurred.
+Two operators were located dead from radiation exposure.
+The third, Richard Legg, was discovered impaled and pinned to the containment ceiling by a control rod—ejected upward during the reactor’s prompt critical event.
+
+\subsubsection{Root Cause Analysis}
+
+SL-1 used top-mounted control rods for reactivity control, which were manually withdrawn during maintenance procedures.
+Operators had previously documented that rods frequently stuck due to thermal expansion of the drive mechanisms.
+It became common to exert above-normal force during removal.
+On the night of the accident, the center control rod—the most reactivity-dominant—was pulled upward over 50 centimeters in a single motion, far beyond the safe insertion limit.
+The resulting reactivity spike caused a runaway chain reaction and a destructive power surge in under 4 milliseconds \autocite{sl1report}.
+
+\subsubsection{Engineering and Human Factors Implications}
+
+Despite repeated reports of sticking rods, the system lacked mechanical or procedural interlocks to prevent rapid withdrawal.
+A single operator retained full manual control, with no redundancy or supervisory lockouts.
+The SL-1 event shows the critical role of human factors engineering, particularly in minimizing design interfaces that allow unsafe manual operations.
+
+\subsubsection{Aftermath and Lessons Learned}
+
+The accident ended the U.S. Army’s independent nuclear program and reshaped reactor safety policy. Key engineering and operational reforms included:
+\begin{itemize}
+  \item Mandatory interlock systems for reactivity-affecting maintenance tasks.
+  \item Human-machine interface redesign to reduce physical leverage on control rods.
+  \item Enhanced operator training in recognizing mechanical anomalies and enforcing procedural compliance.
+  \item Systematic feedback integration from preventative maintenance logs into safety reviews.
+\end{itemize}
+
+SL-1 is now a foundational case in nuclear safety education, cited for both its engineering oversights and its tragic human cost.
+The accident remains a clear demonstration of how well-known mechanical issues—if left unresolved—can lead to fatal outcomes even in small-scale, research-focused facilities \autocite{sl1report}.
+
+\subsection{Three Mile Island: Interface Confusion and Human Factors}
+The Three Mile Island Unit 2 (TMI-2) accident occurred on March 28, 1979, near Harrisburg, Pennsylvania, and remains the most serious commercial nuclear accident in the United States.
+It was precipitated by the failure of a pressure-operated relief valve (PORV) that became stuck open during a minor malfunction.
+The valve allowed coolant to escape from the pressurizer, but due to inadequate instrumentation, operators believed it had closed properly \autocite{tmiwalker}.
+
+\subsubsection{Sequence of Events}
+The event began with a relatively minor mechanical issue: a malfunction in the secondary cooling system caused the reactor to automatically shut down.
+In response, the PORV opened to relieve pressure but failed to reseat.
+As coolant escaped from the system, the core began to overheat.
+However, operators—faced with a cascade of alarms and ambiguous signals—misinterpreted the problem.
+Believing the system to be overpressurized, they manually shut off the emergency core cooling system (ECCS), thereby worsening the core heat-up.
+This operator action, based on incomplete and confusing data, allowed the core to partially melt.
+
+\subsubsection{Design Oversights and Interface Failures}
+Although the specific model of relief valve had a history of sticking open at other nuclear facilities, including ones that caught the issue during routine maintenance, this design vulnerability remained unaddressed in TMI-2.
+The larger problem, however, lay in the information architecture available to operators.
+The control room flooded with alarms—many irrelevant or contradictory—led to cognitive overload.
+As seen in Figure~\ref{fig:tmicontrolroom}, the control interface lacked a clear indication of the PORV status.
+Nowadays, a temperature sensor is typically placed downstream of all pressure relief valves to indicate the actuation of the system, with a clear, dedicated alarm for each one.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.75\textwidth]{tmicontrolroom.png}
+  \caption{The control room of TMI-2 on April 3, 1979, following the accident \autocite{tmiwalker}.}
+  \label{fig:tmicontrolroom}
+\end{figure}
+
+\subsubsection{Human Factors and System Complexity}
+Prior to this event, the relationship between human performance and system design was underappreciated in reactor control engineering.
+It was often assumed that proper training and “common sense” would be sufficient for managing system faults.
+However, the TMI-2 accident highlighted that even well-trained operators can make catastrophic errors when systems are not designed to support human cognition under stress.
+
+As a result of this accident, the field of human factors engineering gained legitimacy within the nuclear industry.
+Organizational and industrial psychology—especially the study of ergonomics and operator-centered design—emerged as crucial disciplines in instrumentation and control system development \autocite{meshkati1991human}.
+Operators needed interfaces not merely for function, but for situational awareness and decision-making under pressure.
+This marked a turning point in control room design philosophies, ushering in era-defining updates to alarm management, display hierarchy, and interface logic \autocite{moderninstruments}.
+
+\subsubsection{Legacy and Regulatory Reform}
+The Nuclear Regulatory Commission (NRC) and the Institute of Nuclear Power Operations (INPO) were restructured in the aftermath to emphasize operator performance, training, and usability engineering.
+New protocols required validation of human-system interfaces (HSIs), and updated control room standards began to emphasize clarity, prioritization of alarms, and the inclusion of status indicators that leave less to inference.
+
+The TMI-2 incident stands as a foundational example in nuclear operations of how accidents may not arise solely from equipment failure, but from the interactions between human minds and poorly designed systems.
+
+\subsection{Chernobyl and Reactor Design Faults}
+
+\subsubsection{Background and Test Scenario}
+The catastrophic accident at Chernobyl's Reactor No.~4 occurred on April 26, 1986, during a late-night safety test.
+The test intended to verify whether the rotational inertia of the turbine could temporarily power the emergency coolant pumps in the event of a grid power failure \autocite{wna2023chernobyl}.
+The RBMK-1000 reactor involved was a Soviet-designed graphite-moderated, water-cooled system—flawed by both design and execution \autocite{iaea1992chernobyl, nrc_chernobyl}.
+
+\subsubsection{Design Flaws and Positive Feedback}
+A defining flaw of the RBMK design was its \textit{positive void coefficient}, which meant that a reduction in coolant (forming steam voids) actually increased the nuclear reaction rate.
+This characteristic starkly contrasts with most Western designs, where coolant voids introduce negative reactivity and suppress power output \autocite{mitocw_rbmkreview}.
+
+During the test, a buildup of xenon-135 (a neutron absorber) had suppressed reactor power.
+Operators, under pressure and facing decaying xenon levels, withdrew control rods to maintain minimum power.
+When reactivity returned, they initiated a SCRAM, inserting all control rods.
+
+\subsubsection{SCRAM-Induced Prompt Criticality}
+Ordinarily, a SCRAM should halt a reaction.
+However, the RBMK rods featured graphite tips that displaced neutron-absorbing water before inserting the boron carbide absorber sections.
+This caused a momentary \textit{increase} in reactivity in the lower portion of the core, leading to a rapid, uncontrollable rise in power.
+The reactor reached a prompt critical state within seconds \autocite{iaea1992chernobyl}.
+
+\subsubsection{Core Destruction and Radiation Release}
+The core overheated instantly, rupturing fuel channels and vaporizing coolant.
+A violent steam explosion lifted the reactor’s 2,000-ton upper biological shield.
+A second explosion—possibly from hydrogen or steam—further breached the structure and exposed the graphite moderator, which caught fire.
+Figure~\ref{fig:chernobylafter} shows the aftermath of the explosion mere hours after the explosion.
+The fire lofted radioactive particles into the upper atmosphere, affecting much of Europe \autocite{nrc_chernobyl}.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.45\textwidth]{chernobylafter.jpg}
+  \caption{Chernobyl Reactor No.~4 on April 27, 1986, 14 hours after the explosion \autocite{medvedev1991truth}.}
+  \label{fig:chernobylafter}
+\end{figure}
+
+\subsubsection{Operator Training and System Transparency}
+Many plant workers had no clear understanding of the RBMK reactor's inherent instability.
+Critical safety data were often classified or minimized due to Cold War secrecy.
+The Soviet nuclear industry prioritized production of military-grade plutonium over safety, compounding operator vulnerability \autocite{medvedev1991truth}.
+
+\subsubsection{Aftermath and Human Toll}
+First responders—mostly firefighters and plant workers—entered the facility without radiation protection.
+Within weeks, dozens died from acute radiation syndrome.
+Long-term effects include increases in thyroid cancer and widespread environmental contamination.
+Despite this, many local engineers and volunteers heroically fought to limit the spread of damage.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.5\textwidth]{elephantsfoot.png}
+  \caption{The ``Elephant’s Foot'': a mass of corium under Reactor No.~4, composed of melted fuel, sand, and reactor structural materials \autocite{medvedev1991truth}.}
+  \label{fig:elephantsfoot}
+\end{figure}
+
+\subsubsection{Legacy of Chernobyl}
+The aftermath of Chernobyl catalyzed global reevaluation of nuclear safety, reactor design, and operator training.
+The international community pushed for increased transparency, safety audits, and enhanced containment strategies.
+The photograph in Figure~\ref{fig:elephantsfoot} famously shows the “Elephant’s Foot,” a deadly corium formation.
+The strange visual static in the image is not lightning, but film degradation from intense radiation—testament to the unprecedented radioactive environment at the site.
+
+The disaster marked a permanent shift in public perception of nuclear energy, influencing decades of policy and triggering a global slowdown in nuclear development.
+
+\subsection{Fukushima-Daiichi: Environmental Consequences}
+
+\subsubsection{Earthquake and Tsunami Impact}
+On March 11, 2011, a magnitude 9.0 undersea megathrust earthquake struck off the northeastern coast of Honshu, Japan, triggering a tsunami with wave heights exceeding 14 meters.
+The Fukushima-Daiichi Nuclear Power Station, operated by TEPCO, was critically affected.
+Although the three operating reactors—Units 1, 2, and 3—successfully SCRAMed (shut down automatically), the tsunami flooded the site and disabled both the offsite grid connections and emergency diesel generators \autocite{iaea_fukushima_report}.
+This unprecedented loss of power across all units is illustrated in Figure~\ref{fig:containments}, which shows the reactor buildings after successive hydrogen explosions.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.5\textwidth]{fukushimaafter.jpg}
+  \caption{Fukushima-Daiichi Nuclear Power Station: The four damaged reactor containments (from left: Units 4, 3, 2, and 1) on 16 March, 2011 \autocite{worldnuclear_fukushima_review}.}
+  \label{fig:containments}
+\end{figure}
+
+\subsubsection{Station Blackout and Core Meltdown}
+Following shutdown, nuclear reactors continue to emit decay heat, which can exceed several percent of nominal thermal power in the hours after fission halts.
+This residual heat must be actively removed to avoid core damage.
+At Fukushima, all cooling systems failed due to the flooding of diesel generators and electrical switchgear—many of which were located in basements vulnerable to tsunami inundation \autocite{nrc_fukushima}.
+This left the plant in a condition known in the nuclear industry as \textit{dead electric}, where no AC or DC power is available to operate pumps, valves, or even basic monitoring systems.
+
+\subsubsection{Loss of Monitoring and Improvised Measures}
+Operators at Fukushima recognized the danger but were rendered effectively blind by the loss of power.
+Unlike U.S.
+Navy reactor systems—where portable test equipment such as Fluke digital multimeters can be attached to terminals for passive thermocouple readings—the Fukushima design lacked redundant manual monitoring options.
+In a desperate act of ingenuity, personnel retrieved car batteries from nearby vehicles and wired them in series to restore minimal voltage to power essential instrumentation.
+Unfortunately, by the time this improvised power source was implemented, core damage in Units 1, 2, and 3 was already underway, with partial or full meltdown confirmed in post-event analyses \autocite{worldnuclear_fukushima_review}.
+
+\subsubsection{Environmental and Human Consequences}
+While the total release of radioactive material was less than that of Chernobyl, Fukushima's dual-trigger origin—a natural disaster followed by technological failure—had extensive consequences.
+Hydrogen explosions damaged reactor buildings, and emergency cooling water—now contaminated—leaked into the Pacific Ocean.
+Airborne dispersion of iodine-131 and cesium-137 prompted mass evacuations and long-term exclusion zones.
+The disaster left Japan facing not only technical containment challenges, but also long-lasting psychological, economic, and political fallout.
+
+\subsubsection{U.S. Naval Involvement and Humanitarian Support}
+As news of the crisis reached the global community, international responses were swift.
+Aboard the aircraft carrier \textit{USS Ronald Reagan} (CVN-76), routine Air Particulate Detector readings spiked during transit toward Japan.
+Recognizing the gravity of the situation, the commanding officer ordered the ship to provide humanitarian aid.
+The Reagan positioned itself offshore and used its dual nuclear reactors to reverse-feed electrical power into Japan's devastated grid infrastructure, offering vital support during the early days of recovery and relief.
+
+This cooperative effort emphasized the unique role nuclear-powered vessels can play in civilian emergency response—an often-overlooked advantage of maritime nuclear propulsion.
+
+\section{Human Factors: The Ethical Dimension of Automation}
+\label{sec:humanfactors}
+
+The role of trained human operators in the safe and reliable generation of nuclear power cannot be overstated.
+Despite vast improvements in automation, instrumentation, and control systems, the nuclear accidents at SL-1, Three Mile Island, Chernobyl, and Fukushima-Daiichi demonstrate that the human element remains indispensable.
+These incidents demonstrate the importance of sound judgment, situational awareness, and expert intervention when systems fail or behave unpredictably \autocite{moderninstruments}.
+
+\subsection{Human Factors Engineering}
+
+\textit{Human factors engineering} (HFE) is the systematic application of knowledge about human capabilities, limitations, behaviors, and interactions to the design of systems, equipment, and tasks.
+In the nuclear industry, HFE plays a vital role in ensuring that complex control systems support rather than hinder operator performance—especially in high-pressure situations.
+
+As nuclear plants have grown in complexity and power output, the design of control rooms has evolved to integrate ergonomic layouts, intuitive interface systems, and simulation-based training environments.
+This progression aims to reduce human error and improve decision-making reliability \autocite{moderninstruments}.
+
+International standards like IEC 964 and IEC 60964 provide guidelines for human-system integration in nuclear control systems \autocite{iec60964, iec964}. These standards emphasize:
+
+\begin{itemize}
+  \item Alignment between power production and safety objectives.
+  \item Proper allocation of tasks between manual and automatic controls.
+  \item Application of human performance data, including anthropometrics, cognitive psychology, and reliability analysis.
+\end{itemize}
+
+Key elements of human factors integration include the reliable \textit{presentation of information}, \textit{uniform control design}, well-thought-out \textit{panel and display layouts}, and procedural integration via \textit{full-scope simulators} that replicate real plant conditions \autocite{iec60964, iec964}.
+
+\subsection{Operator Experience}
+
+While engineering design establishes operational parameters, it is the operator who understands the plant in real time.
+Experienced operators know how the system responds to subtle inputs—such as a one-second control rod shim or a \SI{10}{psig} fluctuation in secondary steam pressure.
+Similarly, machinery room technicians can predict how valve sequencing will affect temperature, flow, or turbine response.
+
+In early prototype plants, small reactor cores were operated at constant loads by teams of scientists and engineers pushing the limits of atomic knowledge \autocite{moderninstruments}.
+As commercial nuclear power scaled up, these systems became vastly more complex, and measurement accuracy and control precision grew increasingly important.
+
+Modern operators are trained not only on procedural execution but also on pattern recognition, anomaly detection, and adaptive response.
+Their insights cannot be easily modeled or coded into automation algorithms, especially under degraded conditions or unforeseen emergencies.
+
+\subsection{Ethical Limits of Automation}
+
+While automation has dramatically improved reactor safety, it must be implemented with caution.
+In certain crisis scenarios, reliance on preprogrammed logic or incomplete models may lead to incorrect or delayed actions.
+Ethical considerations demand that humans remain at the center of safety-critical decisions, particularly when outcomes involve lives or large-scale environmental consequences.
+
+One well-documented challenge in cognitive psychology is the \textit{tunnel effect}, in which humans persist with an initial, incorrect interpretation of a situation, especially under stress \autocite{moderninstruments}.
+This phenomenon is exacerbated when control system layouts are unintuitive or overwhelming.
+Retrofitting advanced instrumentation without corresponding operator retraining can invalidate experience and increase error likelihood.
+
+Moreover, automated models are developed under assumptions that may no longer apply in aging plants.
+Changes to fuel characteristics, core configurations, or coolant flow paths can render original logic sequences obsolete.
+This can lead to spurious trips, false alarms, or missed safety signals—potentially costing a commercial plant lost megawatt-hours, or a naval reactor lost propulsion.
+
+For these reasons, complete automation is neither practical nor desirable in contemporary nuclear power plants.
+A hybrid architecture—where human operators are empowered by automation, not replaced by it—is currently the safest and most reliable solution.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.5\textwidth]{automationfactors}
+  \caption{Automation and human factors in nuclear power generation \autocite{moderninstruments}.}
+  \label{fig:automationfactors}
+\end{figure}
+
+The optimal nuclear power plant integrates advanced control systems with a highly trained crew.
+These operators undergo continuous simulation training for casualty scenarios and are equipped with real-time diagnostic tools.
+Such a system ensures robust safety, resilience to abnormal events, and operational excellence.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.5\textwidth]{optimalautomation.png}
+  \caption{The optimal balance between automation and human factors in nuclear power generation \autocite{moderninstruments}.}
+  \label{fig:optimalautomation}
+\end{figure}
+\section{Economic Realities and Infrastructure Decline}
+
+Due to the high-profile mishaps discussed in Section~\ref{sec:accidents}, the nuclear industry has suffered a significant loss of public trust.
+This erosion of confidence has contributed to financial uncertainty, stalled development, and a decline in domestic infrastructure to support new reactor construction.
+The resulting industrial contraction has drawn comparisons to the decline of the coal rail economy—once supported by a network of coal suppliers and rail hubs across the United States.
+As investment and skilled labor shift toward renewable energy, nuclear power is increasingly seen as an economically fragile legacy technology.
+
+\subsection{Waste Management and Decommissioning Costs}
+\label{sec:wastemanagement}
+
+A commercial nuclear reactor undergoes refueling approximately every 18 to 24 months, with each cycle costing between \$50 million and \$100 million \autocite{wnisr2022}.
+These costs include fuel handling, transport logistics, containment, and licensing.
+Spent fuel is highly radioactive and must be isolated from the environment for thousands of years, necessitating complex storage and oversight systems.
+
+The U.S.
+currently stores most of its spent fuel onsite in dry cask storage, as it lacks a permanent geologic repository.
+The proposed Yucca Mountain site in Nevada was halted due to political opposition, leaving a critical infrastructure gap \autocite{doe2020}.
+Meanwhile, the Department of Energy (DOE) estimates that total long-term waste management liabilities—including storage, monitoring, and decommissioning—could exceed \$500 billion over the next century \autocite{doe2020}.
+Several private firms that formerly handled nuclear logistics and fuel services have transitioned into renewable technologies or left the sector entirely, further weakening the industry’s support structure.
+
+\subsection{Construction and Design of Future Plants}
+
+The capability to construct new nuclear facilities has also diminished.
+The completion of Vogtle Unit~4 in Georgia—one of the first new U.S.
+reactors in decades—offered a glimmer of progress, but also exposed the harsh economic and logistical realities of modern nuclear construction.
+The project faced a decade of delays and billions in cost overruns, driven by labor shortages, supply chain gaps, and regulatory hurdles \autocite{aafnuclearcosts}.
+
+Domestic capacity to manufacture nuclear-grade components, such as reactor pressure vessels or reactor coolant pumps, has eroded.
+Only a few specialized forges worldwide are still capable of producing these massive components to the necessary ASME and NRC specifications.
+Simultaneously, industrial firms that once supported nuclear projects have shifted their operations toward building wind turbine towers, solar inverters, and battery storage systems.
+The opportunity cost of investing in nuclear has grown as renewables offer faster return on investment and benefit from robust public and private incentives.
+
+\subsection{Migration from Nuclear to Renewable Energy Sectors}
+
+Global investment patterns show a clear transition from nuclear to renewables.
+Engineering giants like Siemens and General Electric have scaled back or entirely exited nuclear ventures in favor of wind, solar, and energy storage \autocite{iea2020}.
+In 2022 alone, global capital expenditures on solar and wind eclipsed nuclear investments by more than fivefold \autocite{wnisr2022}.
+
+This migration is not merely market-driven; it is reinforced by regulatory environments.
+Post-Fukushima mandates have led to stricter safety requirements and more frequent inspections, raising compliance and licensing costs.
+Some U.S.
+plants face annual regulatory costs in excess of \$20 million \autocite{aafnuclearcosts}.
+These growing financial burdens make nuclear projects less attractive in both the public and private sectors.
+
+\subsection{Integration Challenges and Opportunities}
+
+Despite economic challenges, nuclear energy offers unique value as a source of stable baseload power.
+This makes it an ideal complement to intermittent renewable sources like wind and solar.
+However, current policy frameworks often prioritize short-term carbon reduction via rapidly deployable renewables, sidelining nuclear investments that require long lead times and upfront capital.
+
+A salient example is the rise of offshore wind infrastructure in coastal cities such as New London, Connecticut.
+These projects attract federal and state subsidies, media attention, and educational investments, while nuclear plants are increasingly perceived as costly, inflexible, or outdated \autocite{abdussami2025future}.
+
+Public perception remains a formidable barrier.
+While nuclear power remains one of the lowest-carbon energy sources per unit of energy generated, its association with catastrophic events has left a lasting impact on societal attitudes.
+Without substantial policy reform and education efforts, nuclear energy will remain economically disadvantaged despite its technical merits.
+
+\section{Conclusion: A Cautious Return to the Atom}
+
+Nuclear power stands at a complex crossroads—one defined by profound ethical obligations, hard-earned technological lessons, and a rapidly shifting energy landscape.
+The historical record, shaped by the tragic events at SL-1, Three Mile Island, Chernobyl, and Fukushima-Daiichi, offers clear testimony to the irreplaceable role of human judgment in reactor safety.
+These accidents remind us that automation, while powerful, must be designed with the understanding that it supports rather than supplants trained human operators.
+
+Human factors engineering has evolved into a critical field, embedding psychological, ergonomic, and procedural insights directly into plant design.
+Simulators, intuitive control room layouts, and refined information presentation now stand as defenses against the tunnel vision and decision fatigue that so often accompany crises.
+The ethical responsibility to keep a human in the loop—especially in decisions with irreversible consequences—cannot be ignored, even as algorithms grow more capable and systems more autonomous.
+
+Yet, the technical and economic challenges facing nuclear power are sobering.
+The decline of manufacturing capacity, the shuttering of legacy vendors, and the migration of talent and capital toward renewables have left the nuclear sector precariously under-resourced.
+Without consistent policy support and industrial reinvestment, the risk of losing irreplaceable expertise and infrastructure grows with each passing year.
+The lack of a permanent waste repository and the mounting costs of decommissioning further complicate the long-term viability of the industry.
+
+Still, all is not lost.
+Projects like Vogtle Units~3 and~4 in Georgia—despite cost overruns and delays—signal a possible turning point.
+They represent not only a technological achievement but also a renewal of confidence in the idea that nuclear energy can still play a vital role in our decarbonized future.
+The emergence of small modular reactors (SMRs), the refinement of passive safety designs, and renewed global interest in energy resilience have reinvigorated the conversation.
+
+To move forward, we must blend the hard-won wisdom of past experience with the innovative tools of modern engineering.
+Nuclear power will never be perfect—but neither is any energy source.
+What makes it unique is the immense power it offers, not just in megawatts, but in the philosophical sense: a force that must be handled with reverence, precision, and foresight.
+The atom, split with care and humility, may yet illuminate a sustainable and responsible path forward.\footnote{An LLM was used strictly for formatting IAW IEEE Editorial Style Manual
 	for Authors. All thoughts, prose, and conclusions are
@@
 \addcontentsline{toc}{section}{References}
-\nocite{*}
+\begingroup
+\sloppy
 \printbibliography
+\endgroup
 
 \end{document}
-
-% vim: set ft=tex tw=80 ts=2 sts=2 sw=2 noet spell:

### 2025-06-06 Slides/PaperTwoPresentation/presentation.tex
work on paper 2 and presentation

% ECE4900W Communicating Engineering Solutions
% Author: Arturo Salinas-Aguayo <artjsalina5@uconn.edu>
\documentclass{beamer}
\usetheme{Warsaw}
\usecolortheme{dolphin}
\usefonttheme{professionalfonts}
\setbeamertemplate{navigation symbols}{}
\usepackage{graphicx}
\usepackage{tavox}

\title[Safe Nuclear Power]{Safe Nuclear Power: Instrumentation, Human Oversight, and Infrastructure Transition}
\subtitle{ECE 4900W – Summer 2025}
\author[Arturo Salinas]{Arturo Salinas-Aguayo}
\institute[UConn]{University of Connecticut\\College of Engineering}
\date{\today}

\graphicspath{{../../images/}}
% Bibliography
\addbibresource{references.bib}

\begin{document}

\begin{frame}
  \titlepage
  \speak{Welcome. This presentation explores the technological, ethical, and infrastructural dimensions of nuclear power in the modern world.}
\end{frame}

\begin{frame}{Outline}
  \tableofcontents[pausesections]
  \speak{Here is an overview of the major themes we will discuss today.}
\end{frame}

\section{Introduction}

\begin{frame}{Motivation}
  \speak{Nuclear energy offers unmatched energy density and reliability. Yet its history demands vigilance, ethics, and technical excellence.}
  \begin{itemize}
    \item Low emissions, high stability
    \item Risk perception shaped by major accidents
    \item Focus of this work: instrumentation, human oversight, infrastructure
  \end{itemize}
\end{frame}

\section{Instrumentation and Control}

\begin{frame}{Types of Reactors}
  \speak{Different reactor types use different coolants and designs, influencing their control strategies.}
  \begin{itemize}
    \item Pressurized Water Reactor (PWR)
    \item Boiling Water Reactor (BWR)
    \item CANDU (Heavy Water Reactor)
    \item Gas-Cooled Reactor (AGR)
  \end{itemize}
\end{frame}

\begin{frame}{Instrumentation in a PWR}
  \speak{This diagram shows how sensors monitor power, pressure, and temperature throughout a PWR.}
  \includegraphics[width=0.85\textwidth]{images/instrumentation.png}
\end{frame}

\section{Historical Lessons}

\begin{frame}{SL-1 and Three Mile Island}
  \speak{SL-1 and Three Mile Island showed the risks of inadequate interlocks and overwhelming operator interfaces.}
  \begin{itemize}
    \item SL-1: control rod ejection, fatal steam explosion
    \item TMI: misread signals led to core exposure
    \item Both: highlight flaws in human-system integration
  \end{itemize}
\end{frame}

\begin{frame}{Chernobyl and Fukushima}
  \speak{Chernobyl and Fukushima illustrate the danger of flawed assumptions and the limits of automation.}
  \includegraphics[width=0.75\textwidth]{images/chernobylafter.jpg}
  \includegraphics[width=0.75\textwidth]{images/fukushimaafter.jpg}
\end{frame}

\section{Human Factors}

\begin{frame}{Human Factors Engineering}
  \speak{Human factors engineering is vital in designing systems that support, not hinder, decision making.}
  \begin{itemize}
    \item Ergonomic control rooms
    \item Full-scope simulation
    \item Situational awareness tools
  \end{itemize}
\end{frame}

\begin{frame}{Ethics of Automation}
  \speak{Automation must be supervised. Tunnel vision and outdated assumptions can cause failure without human oversight.}
  \begin{itemize}
    \item Humans mitigate uncertainty and change
    \item Ethics demand human-in-the-loop
  \end{itemize}
\end{frame}

\section{Economic Challenges}

\begin{frame}{Waste and Infrastructure}
  \speak{The economic burden of waste management and infrastructure decline limits nuclear's future.}
  \begin{itemize}
    \item High refueling and decommissioning costs
    \item No permanent U.S. waste site
    \item Industry contraction post-Fukushima
  \end{itemize}
\end{frame}

\begin{frame}{Hopeful Developments}
  \speak{New efforts like Vogtle and SMRs show a path forward. Nuclear remains vital for base-load generation.}
  \begin{itemize}
    \item Vogtle Unit 4 completed
    \item Passive safety designs
    \item Small Modular Reactors
  \end{itemize}
\end{frame}

\section{Conclusion}

\begin{frame}{Final Thoughts}
  \speak{With humility and innovation, nuclear energy can be safe, ethical, and essential in the fight against climate change.}
  \begin{itemize}
    \item Ethics + engineering = resilient systems
    \item Human vigilance must never be replaced
    \item The atom, split with care, still holds promise
  \end{itemize}
\end{frame}

\end{document}

### 2025-06-07 Slides/PaperTwoPresentation/presentation.tex
update presentation

@@
 \usepackage{graphicx}
-\usepackage{tavox}
+\usepackage{biblatex}
+\addbibresource{references.bib}
+\graphicspath{{../../images/}}
 
@@
 
-\graphicspath{{../../images/}}
-% Bibliography
-\addbibresource{references.bib}
-
 \begin{document}
@@
   \titlepage
-  \speak{Welcome. This presentation explores the technological, ethical, and infrastructural dimensions of nuclear power in the modern world.}
+  \speak{Welcome.
+This presentation explores the technological, ethical, and infrastructural dimensions of nuclear power in the modern world.
+We will cover lessons from historic accidents, instrumentation, human-machine interface design, and current innovations.
+Each story reveals both vulnerability and resilience.}
 \end{frame}
@@
   \tableofcontents[pausesections]
-  \speak{Here is an overview of the major themes we will discuss today.}
+  \speak{Here is the roadmap for this presentation. Each section builds on the last, culminating in a call for renewed responsibility and innovation in the nuclear field.}
 \end{frame}
@@
 \begin{frame}{Motivation}
-  \speak{Nuclear energy offers unmatched energy density and reliability. Yet its history demands vigilance, ethics, and technical excellence.}
+  \subsubsection*{Why Nuclear Now?}
+  \speak{Nuclear energy provides unmatched energy density and emissions-free base-load power.
+However, public trust and sustainable deployment depend on ethical design, robust control systems, and trained human oversight.
+While wind and solar depend on weather and batteries, nuclear runs steadily.
+But one accident can erode decades of progress.}
   \begin{itemize}
-    \item Low emissions, high stability
-    \item Risk perception shaped by major accidents
-    \item Focus of this work: instrumentation, human oversight, infrastructure
+    \item Zero-emission base-load power
+    \item High energy return on investment
+    \item Historical trauma shapes perception \autocite{nrc_chernobyl, worldnuclear_fukushima_review}
+    \item Paper focuses on safety through instrumentation and oversight
   \end{itemize}
@@
 \begin{frame}{Types of Reactors}
-  \speak{Different reactor types use different coolants and designs, influencing their control strategies.}
+  \subsubsection*{Diverse Architectures}
+  \speak{Let’s begin with an overview of reactor types, since their thermal-hydraulic behavior and control logic affect how instrumentation is designed. Each has different challenges.}
   \begin{itemize}
-    \item Pressurized Water Reactor (PWR)
-    \item Boiling Water Reactor (BWR)
-    \item CANDU (Heavy Water Reactor)
-    \item Gas-Cooled Reactor (AGR)
+    \item Pressurized Water Reactor (PWR): most common, used in submarines and civilian fleets
+    \item Boiling Water Reactor (BWR): direct steam path to turbine
+    \item CANDU: heavy water as moderator and coolant
+    \item Gas-cooled (AGR): high outlet temperature for efficiency
   \end{itemize}
@@
 \begin{frame}{Instrumentation in a PWR}
-  \speak{This diagram shows how sensors monitor power, pressure, and temperature throughout a PWR.}
-  \includegraphics[width=0.85\textwidth]{images/instrumentation.png}
+  \subsubsection*{Sensor Coverage}
+  \speak{This diagram shows typical sensor and actuator placements in a PWR.
+Note the neutron flux monitors, hot leg thermocouples, pressurizer level gauges, and scram breakers.
+Each of these is essential for safe operation.}
+  \includegraphics[width=0.85\textwidth]{instrumentation.png}
 \end{frame}
@@
 
-\begin{frame}{SL-1 and Three Mile Island}
-  \speak{SL-1 and Three Mile Island showed the risks of inadequate interlocks and overwhelming operator interfaces.}
+\begin{frame}{SL-1: Prompt Critical from Manual Control}
+  \subsubsection*{The First Fatal Reactor Accident}
+  \includegraphics[width=0.4\textwidth]{sl1diagram.jpg}
+\end{frame}
+
+\begin{frame}{SL-1: Prompt Critical from Manual Control}
+  \speak{SL-1 was the first fatal reactor accident in the U.S.
+One operator withdrew a control rod too far.
+With no interlocks, the reactor went prompt critical.
+The resulting steam explosion launched a control rod through the ceiling, killing all three operators \autocite{sl1report}.}
   \begin{itemize}
-    \item SL-1: control rod ejection, fatal steam explosion
-    \item TMI: misread signals led to core exposure
-    \item Both: highlight flaws in human-system integration
+    \item Root cause: lack of mechanical safeguards
+    \item Fallout: Army withdrew from nuclear reactor programs
+    \item Image shows post-blast structural damage
   \end{itemize}
@@
 
-\begin{frame}{Chernobyl and Fukushima}
-  \speak{Chernobyl and Fukushima illustrate the danger of flawed assumptions and the limits of automation.}
-  \includegraphics[width=0.75\textwidth]{images/chernobylafter.jpg}
-  \includegraphics[width=0.75\textwidth]{images/fukushimaafter.jpg}
+\begin{frame}{Three Mile Island: Interface Confusion}
+  \subsubsection*{Misleading Indicators}
+  \speak{TMI began with a stuck-open pressure relief valve.
+A light indicated it had closed.
+It hadn’t.
+Operators turned off coolant, thinking the core was overfilled.
+In truth, the core became exposed and melted \autocite{nrc_tmi_report}.}
+  \begin{itemize}
+    \item Interface design: misleading feedback lights
+    \item Alarm overload: dozens triggered at once
+    \item Led to sweeping changes in human-machine interfaces
+  \end{itemize}
+\end{frame}
+
+\begin{frame}{Three Mile Island: Interface Confusion}
+  \centering
+  \includegraphics[width=0.75\textwidth]{tmicontrolroom.png}
+\end{frame}
+
+\begin{frame}{Chernobyl: Design and Assumption Failures}
+  \subsubsection*{When Physics Fights Back}
+  \includegraphics[width=0.6\textwidth]{chernobylafter.jpg}
+  \speak{This photo shows the wreckage of Chernobyl's Reactor No.~4 just hours after the explosion. The graphite fire continued for days, spewing radiation across Europe \autocite{medvedev1991truth}.}
+\end{frame}
+
+\begin{frame}{Chernobyl: Design and Assumption Failures}
+  \speak{Chernobyl's RBMK reactor had a positive void coefficient: as coolant boiled, reactivity rose.
+During a turbine coastdown test, this reactivity led to runaway power.
+A SCRAM—normally safe—made it worse, as control rods had graphite tips \autocite{iaea1992chernobyl}.}
+  \begin{itemize}
+    \item Unstable physics design, especially at low power
+    \item Operators unaware of full core behavior
+    \item Culture of secrecy delayed response and accountability
+  \end{itemize}
+\end{frame}
+
+\begin{frame}{Fukushima: Nature and Neglect}
+  \subsubsection*{Environmental Consequences}
+  \includegraphics[width=0.75\textwidth]{fukushimaafter.jpg}
+  \speak{Fukushima's tsunami flooded backup generators located in basements.
+This loss of power—called station blackout—disabled cooling, and all three operating units melted down \autocite{iaea_fukushima_report}.}
+\end{frame}
+
+\begin{frame}{Fukushima: Nature and Neglect}
+  \speak{Though the reactors shutdown correctly, residual decay heat remained.
+Without cooling pumps, that heat caused core melt.
+Desperate engineers tried to wire car batteries together to read instrument panels \autocite{nrc_fukushima}.}
+  \begin{itemize}
+    \item Failure of backup power design
+    \item Hydrogen explosions from vented zirconium reactions
+    \item Ocean contamination and public trust fallout
+  \end{itemize}
 \end{frame}
@@
 \begin{frame}{Human Factors Engineering}
-  \speak{Human factors engineering is vital in designing systems that support, not hinder, decision making.}
+  \subsubsection*{Designing for the Operator}
+  \speak{Modern control rooms use color-coded touch screens, redundancy, and full-scope training simulators. Human factors seeks to prevent misinterpretation and stress-based mistakes.}
   \begin{itemize}
-    \item Ergonomic control rooms
-    \item Full-scope simulation
-    \item Situational awareness tools
+    \item Cognitive load reduction
+    \item Realistic training environments
+    \item Alarm prioritization and filtering
   \end{itemize}
@@
 \begin{frame}{Ethics of Automation}
-  \speak{Automation must be supervised. Tunnel vision and outdated assumptions can cause failure without human oversight.}
+  \subsubsection*{Keeping the Human in the Loop}
+  \speak{Machines must serve—not replace—humans in nuclear operations.
+SCRAMs, alarms, logic interlocks: all must assume a trained operator will intervene.
+Ethical design means building for fallibility \autocite{mitocw_rbmkreview}.}
   \begin{itemize}
-    \item Humans mitigate uncertainty and change
-    \item Ethics demand human-in-the-loop
+    \item Passive safety aids are essential
+    \item But overtrust in automation can fail silently
+    \item The human-in-the-loop design is non-negotiable
   \end{itemize}
@@
 
-\begin{frame}{Waste and Infrastructure}
-  \speak{The economic burden of waste management and infrastructure decline limits nuclear's future.}
+\begin{frame}{Aging Infrastructure and Waste}
+  \subsubsection*{Legacy Risks}
+  \speak{Beyond physics, the nuclear field is strained economically.
+Most U.S.
+plants were built pre-1980.
+Engineering supply chains have shrunk, and no long-term spent fuel site exists \autocite{wna2023chernobyl}.}
   \begin{itemize}
-    \item High refueling and decommissioning costs
-    \item No permanent U.S. waste site
-    \item Industry contraction post-Fukushima
+    \item On-site pools reaching capacity
+    \item Dry cask storage: temporary fix
+    \item Fewer experts entering the field
   \end{itemize}
@@
 
-\begin{frame}{Hopeful Developments}
-  \speak{New efforts like Vogtle and SMRs show a path forward. Nuclear remains vital for base-load generation.}
+\begin{frame}{Modern Developments}
+  \subsubsection*{New Hope: SMRs and Vogtle}
+  \speak{Still, there is hope.
+Vogtle Units 3 and 4 in Georgia marked the first new reactors in decades.
+SMRs and molten salt reactors promise passive safety and easier siting \autocite{worldnuclear_vogtle}.}
   \begin{itemize}
-    \item Vogtle Unit 4 completed
-    \item Passive safety designs
-    \item Small Modular Reactors
+    \item Factory assembly lowers construction risk
+    \item Smaller cores, faster SCRAM times
+    \item Potential role in green hydrogen and remote grids
   \end{itemize}
@@
 \begin{frame}{Final Thoughts}
-  \speak{With humility and innovation, nuclear energy can be safe, ethical, and essential in the fight against climate change.}
+  \subsubsection*{Resilience Through Design}
+  \speak{From SL-1 to Fukushima, the lesson is clear: safe nuclear power isn't just about neutrons. It's about people, design, and ethics. The atom, when split responsibly, can still light the world.}
   \begin{itemize}
-    \item Ethics + engineering = resilient systems
-    \item Human vigilance must never be replaced
-    \item The atom, split with care, still holds promise
+    \item Prioritize human-machine partnerships
+    \item Design with humility and feedback
+    \item Build public trust through transparency
   \end{itemize}
@@
 
+\begin{frame}{References}
+  \nocite{*}
+  \printbibliography
+\end{frame}
+
 \end{document}

### 2025-06-10 Papers/Paper Two/papertwo.tex
push

@@
 
-Nuclear power has long promised a solution to global energy needs: it offers extremely high energy density, zero direct carbon emissions, and round-the-clock base-load generation.
+Nuclear power has long promised a solution to global energy needs: it offers
+extremely high energy density, zero direct carbon emissions, and round-the-clock
+base-load generation. See Figure~\ref{fig:enterprise} to be reminded of the
+atomic age, a time when nuclear power itself was on the rise, leading to nuclear
+powered carriers, frigates, destroyers, and \textit{without a doubt the best
+warships}, nuclear powered submarines.
+
 Its technical advantages position it as a powerful complement to intermittent renewable sources such as wind and solar.
@@
 It explores how the evolution of digital I\&C systems has transformed modern plant safety while also introducing new vulnerabilities, particularly in terms of human-computer interaction.
-By revisiting major nuclear incidents and analyzing their root causes, this work highlights the ethical and operational boundaries of automation, the critical importance of experienced operators, and the design decisions that shape how human errors propagate—or are prevented.
 
-Ultimately, this paper contends that the future of nuclear power depends not only on engineering excellence, but also on institutional resilience.
-While new technologies like SMRs and passive safety designs represent significant progress, they must be accompanied by policy reform, renewed investment in nuclear education and infrastructure, and a cultural shift toward sustainable, ethically grounded energy stewardship.
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=0.7\textwidth]{enterprise}
+	\caption{The crew of the USS Enterprise spell out $E = mc^2$ as she sets sail
+around the world with similarly nuclear powered USS Long Beach and USS
+Bainbridge on Operation Sea Orbit, 31 July, 1964 (There are also 3 submarines in
+this picture.)}
+\label{fig:enterprise}
+\end{figure}
 
-\section{Instrumentation and Control (I\&C) Systems}
+\section{Technology Description}
+\subsection{Instrumentation and Control (I\&C) Systems}
 
-\subsection{Types of Plants}
+The ability to control a nuclear reaction ultimately \textit{boils down} to managing the energy released from an unstable element that, once stimulated, emits both fast and slow neutrons.
+These neutrons go on to strike other fissile atoms, triggering further emissions in a self-sustaining chain reaction.
+This process takes place within fuel elements—long metal-clad rods arranged in a structure called a fuel subassembly, often resembling a metallic honeycomb.
+Multiple subassemblies are grouped into fuel assemblies, with careful placement to balance the resulting heat flux from neutron collisions.
+As neutrons scatter and interact, they excite more fuel atoms in a cascading effect reminiscent of a Fourth of July fireworks finale.
+
+A simplified chemical representation of the process is the fission of uranium-235:
+
+\begin{equation}
+  ^{235}\text{U} + ^{1}\text{n} \rightarrow ^{92}\text{Kr} + ^{141}\text{Ba} +
+	\text{\textbf{3}}^{1}\text{n} + \text{Energy}~(\approx 200~\text{MeV})
+\end{equation}
+
+\noindent where:
+\begin{itemize}
+  \item $^{235}\text{U}$ is a fissile uranium-235 nucleus,
+  \item $^{1}\text{n}$ is an incoming thermal neutron,
+  \item $^{92}\text{Kr}$ and $^{141}\text{Ba}$ are fission products,
+  \item $3^{1}\text{n}$ represents the emission of three additional neutrons,
+  \item The energy released is approximately 200~MeV, primarily as kinetic energy of the fission fragments and secondary neutrons.
+\end{itemize}
+
+These additional neutrons can go on to initiate further fission reactions, forming the basis of a self-sustaining chain reaction.
+The ability to keep this chain reaction in balance—neither accelerating nor dying out—is at the heart of nuclear reactor control.
+
+The danger, however, lies in the potential for this excitation to grow out of control.
+The extent to which the system is primed for further reaction is measured as \textit{reactivity}, defined approximately as:
+
+\begin{equation}
+  \rho = \frac{k_{\text{eff}} - 1}{k_{\text{eff}}}
+\end{equation}
+where $\rho$ is the reactivity (dimensionless), $k_{\text{eff}}$ is the effective neutron multiplication factor and will increase in power unless corrective action is taken.
+
+Advanced instrumentation and control (I\&C) systems are designed to maintain this reactivity within safe limits—keeping the metaphorical fireworks display safely choreographed in the sky rather than detonating all at once in a catastrophic burst at ground level.
+
+To understand why these systems have evolved into such complex and nuanced architectures, it’s helpful to explore the different reactor design types.
+Over the decades, engineers have developed multiple ways to control reactivity and convert the energy from a “hot rock” into mechanical power—whether to propel submarines, generate electricity, or, in darker moments of history, to build devastating weapons.
+
+\subsubsection*{Types of Plants}
 
@@
 
-\subsection{Reactor Monitoring Principles}
+\subsubsection*{Reactor Monitoring Principles}
 
@@
 
-\subsubsection{Thermal and Neutron Sensing.}
+\subsubsection*{Thermal and Neutron Sensing.}
 Sensors for nuclear plant monitoring fall broadly into two categories: thermal sensors and nuclear-radiation sensors \cite{ichandbook}.
@@
 
-\subsubsection{Coolant and Moderator Instrumentation.}
-Coolant system monitoring includes the following key measurements \cite{ichandbook}:
+\subsubsection*{Coolant and Moderator Instrumentation.}
+Effective thermal management is essential to maintaining reactor integrity.
+The coolant serves as the medium for heat removal from the fuel, while the moderator governs neutron energy and thus reactivity.
+Continuous monitoring of coolant and moderator parameters ensures that the reactor operates within safe thermal limits and that reactivity remains controlled.
+Key measurements include \autocite{ichandbook}:
 \begin{itemize}
@@
 These measurements require temperature sensors, flowmeters, pressure transducers, gamma sensors, and humidity detectors.
-In light water reactors, precise detection of subcooled margin, void fraction, and departure from nucleate boiling (DNB) are paramount.
+In light water reactors, precise detection of subcooled margin, void fraction, and departure from nucleate boiling (DNB) is paramount to prevent fuel damage and ensure efficient heat transfer.
 
-\subsubsection{Control and Structural Monitoring.}
-In addition to thermal-hydraulic measurements, plant instrumentation also tracks key operational parameters:
+\subsubsection*{Control and Structural Monitoring.}
+To maintain criticality and prevent excursions, the physical state of reactor components must be known at all times.
+This includes not only the position of reactivity control elements like control rods, but also the condition of structural moderators and pressure boundaries.
+Accurate readings in this domain allow for dynamic reactivity management, prompt detection of abnormal trends, and safe system operation:
 \begin{itemize}
@@
 \end{itemize}
-Sensors used include linear variable differential transformers (LVDTs), water-level probes, and high-accuracy pressure transducers.
+These parameters are sensed via linear variable differential transformers (LVDTs), water-level probes, and high-accuracy pressure transducers.
+Their outputs are often fed into automated feedback loops and supervisory control systems.
 
-\subsubsection{Steam and Turbine System Sensing.}
-Beyond the reactor itself, I\&C systems must track parameters in the secondary loop and balance-of-plant systems. These include:
+\subsubsection*{Steam and Turbine System Sensing.}
+Instrumentation extends beyond the core into the balance-of-plant systems where thermal energy is converted into mechanical and electrical power.
+Monitoring the secondary loop ensures safe and efficient energy extraction, minimizes mechanical wear, and protects the grid interface.
+Key measurements include:
 \begin{itemize}
@@
 \end{itemize}
-Such measurements use tachometers, electrical metering, differential pressure sensors, and dynamic vibration monitors, ensuring not only safe reactor operation but also efficient energy conversion and grid stability.
+These measurements are collected using tachometers, electrical metering, differential pressure sensors, and dynamic vibration monitors.
+They are vital for maintaining power stability, avoiding turbine overspeed or cavitation, and ensuring that thermal energy is extracted as efficiently and safely as possible.
 
-\subsubsection{Integration with Digital Systems.}
+\subsection{Automatic Protection}
 The increasing integration of digital logic enables centralized monitoring of these diverse parameters using high-speed buses and distributed control systems (DCS).
@@
 
-\subsection{Computerized Protection Systems}
+\subsubsection*{Computerized Protection Systems}
 \label{sec:protection}
@@
 The move to digital logic enhances responsiveness but demands rigorous validation, fail-safe architectures, and operator situational awareness to maintain overall plant safety.
-\section{Historical Lessons in Reactor Safety}
+
+\section{Broader Impacts}
+To understand the foundation of modern nuclear safety, it is essential to examine key historical nuclear accidents.
+These events, extensively studied both in naval reactor training and across the commercial nuclear sector, highlight the serious risks inherent to nuclear technology when safety systems fail.
+The lessons drawn from these incidents have directly influenced the development of today’s robust safety frameworks.
+
+Following this overview, the section explores the broader environmental, economic, and societal consequences of these accidents.
+It discusses how public perception and regulatory response have shaped the trajectory of nuclear power over recent decades, contributing to its slowed adoption despite technological advancements.
+
+\subsection{Historical Lessons in Reactor Safety}
 \label{sec:accidents}
 
-\subsection{SL-1: The First Fatal U.S. Reactor Accident}
+\subsubsection*{SL-1: The First Fatal U.S. Reactor Accident}
 
@@
 
-\subsubsection{Sequence of Events}
-
-Local firefighters arrived in response to a fire alarm and found the facility abandoned.
-Coffee cups remained warm, and food was left uneaten.
-As radiation alarms activated on their dosimetry equipment, they realized a serious radiological event had occurred.
-Two operators were located dead from radiation exposure.
-The third, Richard Legg, was discovered impaled and pinned to the containment ceiling by a control rod—ejected upward during the reactor’s prompt critical event.
-
-\subsubsection{Root Cause Analysis}
-
 SL-1 used top-mounted control rods for reactivity control, which were manually withdrawn during maintenance procedures.
@@
 
-\subsubsection{Engineering and Human Factors Implications}
+Despite repeated reports of sticking rods, the system lacked mechanical or
+procedural interlocks to prevent rapid withdrawal. A single operator retained
+full manual control, with no redundancy or supervisory lockouts. The SL-1 event
+shows the critical role of human factors engineering, particularly in minimizing
+design interfaces (IE maintenance procedures) that allow unsafe manual operations.
 
-Despite repeated reports of sticking rods, the system lacked mechanical or procedural interlocks to prevent rapid withdrawal.
-A single operator retained full manual control, with no redundancy or supervisory lockouts.
-The SL-1 event shows the critical role of human factors engineering, particularly in minimizing design interfaces that allow unsafe manual operations.
-
-\subsubsection{Aftermath and Lessons Learned}
-
-The accident ended the U.S. Army’s independent nuclear program and reshaped reactor safety policy. Key engineering and operational reforms included:
-\begin{itemize}
-  \item Mandatory interlock systems for reactivity-affecting maintenance tasks.
-  \item Human-machine interface redesign to reduce physical leverage on control rods.
-  \item Enhanced operator training in recognizing mechanical anomalies and enforcing procedural compliance.
-  \item Systematic feedback integration from preventative maintenance logs into safety reviews.
-\end{itemize}
-
-SL-1 is now a foundational case in nuclear safety education, cited for both its engineering oversights and its tragic human cost.
-The accident remains a clear demonstration of how well-known mechanical issues—if left unresolved—can lead to fatal outcomes even in small-scale, research-focused facilities \autocite{sl1report}.
-
-\subsection{Three Mile Island: Interface Confusion and Human Factors}
+\subsubsection*{Three Mile Island: Interface Confusion and Human Factors}
 The Three Mile Island Unit 2 (TMI-2) accident occurred on March 28, 1979, near Harrisburg, Pennsylvania, and remains the most serious commercial nuclear accident in the United States.
@@
 
-\subsubsection{Sequence of Events}
-The event began with a relatively minor mechanical issue: a malfunction in the secondary cooling system caused the reactor to automatically shut down.
-In response, the PORV opened to relieve pressure but failed to reseat.
-As coolant escaped from the system, the core began to overheat.
-However, operators—faced with a cascade of alarms and ambiguous signals—misinterpreted the problem.
-Believing the system to be overpressurized, they manually shut off the emergency core cooling system (ECCS), thereby worsening the core heat-up.
-This operator action, based on incomplete and confusing data, allowed the core to partially melt.
-
-\subsubsection{Design Oversights and Interface Failures}
-Although the specific model of relief valve had a history of sticking open at other nuclear facilities, including ones that caught the issue during routine maintenance, this design vulnerability remained unaddressed in TMI-2.
-The larger problem, however, lay in the information architecture available to operators.
-The control room flooded with alarms—many irrelevant or contradictory—led to cognitive overload.
-As seen in Figure~\ref{fig:tmicontrolroom}, the control interface lacked a clear indication of the PORV status.
-Nowadays, a temperature sensor is typically placed downstream of all pressure relief valves to indicate the actuation of the system, with a clear, dedicated alarm for each one.
+The event began with a relatively minor mechanical issue: a malfunction in the
+secondary cooling system caused the reactor to automatically shut down. In
+response, the PORV opened to relieve pressure but failed to reseat. As coolant
+escaped from the system, the core began to overheat. However, operators—faced
+with a cascade of alarms and ambiguous signals—misinterpreted the problem.
+Believing the system to be overpressurized, they manually shut off the emergency
+core cooling system (ECCS), thereby worsening the core heat-up. This operator
+action, based on incomplete and confusing data, allowed the core to partially
+melt. The confusing control room can be seen in Figure~\ref{fig:tmicontrolroom}.
 
@@
 
-\subsubsection{Human Factors and System Complexity}
 Prior to this event, the relationship between human performance and system design was underappreciated in reactor control engineering.
@@
 
-As a result of this accident, the field of human factors engineering gained legitimacy within the nuclear industry.
-Organizational and industrial psychology—especially the study of ergonomics and operator-centered design—emerged as crucial disciplines in instrumentation and control system development \autocite{meshkati1991human}.
-Operators needed interfaces not merely for function, but for situational awareness and decision-making under pressure.
-This marked a turning point in control room design philosophies, ushering in era-defining updates to alarm management, display hierarchy, and interface logic \autocite{moderninstruments}.
+As a result of this accident, the field of human factors engineering gained
+legitimacy within the nuclear industry. Organizational and industrial
+psychology—especially the study of ergonomics and operator-centered
+design—emerged as crucial disciplines in instrumentation and control system
+development \autocite{meshkati1991human}. Operators needed interfaces not merely
+for function, but for situational awareness and decision-making under pressure.
+This marked a turning point in control room design philosophies, ushering in
+era-defining updates to alarm management, display hierarchy, and interface logic
+\autocite{moderninstruments}. This will be explored thoroughly in
+Section~\ref{sec:humanfactors}
 
-\subsubsection{Legacy and Regulatory Reform}
-The Nuclear Regulatory Commission (NRC) and the Institute of Nuclear Power Operations (INPO) were restructured in the aftermath to emphasize operator performance, training, and usability engineering.
-New protocols required validation of human-system interfaces (HSIs), and updated control room standards began to emphasize clarity, prioritization of alarms, and the inclusion of status indicators that leave less to inference.
+\subsubsection*{Chernobyl and Reactor Design Faults}
 
-The TMI-2 incident stands as a foundational example in nuclear operations of how accidents may not arise solely from equipment failure, but from the interactions between human minds and poorly designed systems.
-
-\subsection{Chernobyl and Reactor Design Faults}
-
-\subsubsection{Background and Test Scenario}
 The catastrophic accident at Chernobyl's Reactor No.~4 occurred on April 26, 1986, during a late-night safety test.
@@
 
-\subsubsection{Design Flaws and Positive Feedback}
-A defining flaw of the RBMK design was its \textit{positive void coefficient}, which meant that a reduction in coolant (forming steam voids) actually increased the nuclear reaction rate.
-This characteristic starkly contrasts with most Western designs, where coolant voids introduce negative reactivity and suppress power output \autocite{mitocw_rbmkreview}.
-
-During the test, a buildup of xenon-135 (a neutron absorber) had suppressed reactor power.
-Operators, under pressure and facing decaying xenon levels, withdrew control rods to maintain minimum power.
-When reactivity returned, they initiated a SCRAM, inserting all control rods.
-
-\subsubsection{SCRAM-Induced Prompt Criticality}
 Ordinarily, a SCRAM should halt a reaction.
@@
 
-\subsubsection{Core Destruction and Radiation Release}
 The core overheated instantly, rupturing fuel channels and vaporizing coolant.
@@
 
-\subsubsection{Operator Training and System Transparency}
-Many plant workers had no clear understanding of the RBMK reactor's inherent instability.
-Critical safety data were often classified or minimized due to Cold War secrecy.
-The Soviet nuclear industry prioritized production of military-grade plutonium over safety, compounding operator vulnerability \autocite{medvedev1991truth}.
-
-\subsubsection{Aftermath and Human Toll}
-First responders—mostly firefighters and plant workers—entered the facility without radiation protection.
-Within weeks, dozens died from acute radiation syndrome.
-Long-term effects include increases in thyroid cancer and widespread environmental contamination.
-Despite this, many local engineers and volunteers heroically fought to limit the spread of damage.
+The aftermath of Chernobyl catalyzed global reevaluation of nuclear safety, reactor design, and operator training.
+The international community pushed for increased transparency, safety audits, and enhanced containment strategies.
+The photograph in Figure~\ref{fig:elephantsfoot} famously shows the “Elephant’s Foot,” a deadly corium formation.
+The strange visual static in the image is not lightning, but film degradation from intense radiation—testament to the unprecedented radioactive environment at the site.
 
@@
 
-\subsubsection{Legacy of Chernobyl}
-The aftermath of Chernobyl catalyzed global reevaluation of nuclear safety, reactor design, and operator training.
-The international community pushed for increased transparency, safety audits, and enhanced containment strategies.
-The photograph in Figure~\ref{fig:elephantsfoot} famously shows the “Elephant’s Foot,” a deadly corium formation.
-The strange visual static in the image is not lightning, but film degradation from intense radiation—testament to the unprecedented radioactive environment at the site.
+The disaster marked a permanent shift in public perception of nuclear energy,
+influencing decades of policy and triggering a global slowdown in nuclear
+development. This politically changed the entire world's perception of the
+potential for ``Safe Nuclear Power" and will be discussed in
+Section~\ref{sec:infrastructure}.
 
-The disaster marked a permanent shift in public perception of nuclear energy, influencing decades of policy and triggering a global slowdown in nuclear development.
+\subsubsection*{Fukushima-Daiichi: Environmental Consequences}
 
-\subsection{Fukushima-Daiichi: Environmental Consequences}
-
-\subsubsection{Earthquake and Tsunami Impact}
 On March 11, 2011, a magnitude 9.0 undersea megathrust earthquake struck off the northeastern coast of Honshu, Japan, triggering a tsunami with wave heights exceeding 14 meters.
 The Fukushima-Daiichi Nuclear Power Station, operated by TEPCO, was critically affected.
-Although the three operating reactors—Units 1, 2, and 3—successfully SCRAMed (shut down automatically), the tsunami flooded the site and disabled both the offsite grid connections and emergency diesel generators \autocite{iaea_fukushima_report}.
+Although the three operating reactors—Units 1, 2, and 3—successfully SCRAMed, the tsunami flooded the site and disabled both the offsite grid connections and emergency diesel generators \autocite{iaea_fukushima_report}.
 This unprecedented loss of power across all units is illustrated in Figure~\ref{fig:containments}, which shows the reactor buildings after successive hydrogen explosions.
@@
 
-\subsubsection{Station Blackout and Core Meltdown}
 Following shutdown, nuclear reactors continue to emit decay heat, which can exceed several percent of nominal thermal power in the hours after fission halts.
@@
 
-\subsubsection{Loss of Monitoring and Improvised Measures}
 Operators at Fukushima recognized the danger but were rendered effectively blind by the loss of power.
@@
 
-\subsubsection{Environmental and Human Consequences}
-While the total release of radioactive material was less than that of Chernobyl, Fukushima's dual-trigger origin—a natural disaster followed by technological failure—had extensive consequences.
-Hydrogen explosions damaged reactor buildings, and emergency cooling water—now contaminated—leaked into the Pacific Ocean.
-Airborne dispersion of iodine-131 and cesium-137 prompted mass evacuations and long-term exclusion zones.
-The disaster left Japan facing not only technical containment challenges, but also long-lasting psychological, economic, and political fallout.
-
-\subsubsection{U.S. Naval Involvement and Humanitarian Support}
-As news of the crisis reached the global community, international responses were swift.
-Aboard the aircraft carrier \textit{USS Ronald Reagan} (CVN-76), routine Air Particulate Detector readings spiked during transit toward Japan.
-Recognizing the gravity of the situation, the commanding officer ordered the ship to provide humanitarian aid.
-The Reagan positioned itself offshore and used its dual nuclear reactors to reverse-feed electrical power into Japan's devastated grid infrastructure, offering vital support during the early days of recovery and relief.
-
-This cooperative effort emphasized the unique role nuclear-powered vessels can play in civilian emergency response—an often-overlooked advantage of maritime nuclear propulsion.
-
-\section{Human Factors: The Ethical Dimension of Automation}
+\subsection{Human Factors: The Ethical Dimension of Automation}
 \label{sec:humanfactors}
@@
 
-\subsection{Human Factors Engineering}
-
 \textit{Human factors engineering} (HFE) is the systematic application of knowledge about human capabilities, limitations, behaviors, and interactions to the design of systems, equipment, and tasks.
@@
 
-As nuclear plants have grown in complexity and power output, the design of control rooms has evolved to integrate ergonomic layouts, intuitive interface systems, and simulation-based training environments.
-This progression aims to reduce human error and improve decision-making reliability \autocite{moderninstruments}.
-
-International standards like IEC 964 and IEC 60964 provide guidelines for human-system integration in nuclear control systems \autocite{iec60964, iec964}. These standards emphasize:
-
-\begin{itemize}
-  \item Alignment between power production and safety objectives.
-  \item Proper allocation of tasks between manual and automatic controls.
-  \item Application of human performance data, including anthropometrics, cognitive psychology, and reliability analysis.
-\end{itemize}
-
 Key elements of human factors integration include the reliable \textit{presentation of information}, \textit{uniform control design}, well-thought-out \textit{panel and display layouts}, and procedural integration via \textit{full-scope simulators} that replicate real plant conditions \autocite{iec60964, iec964}.
 
-\subsection{Operator Experience}
-
+\subsubsection*{Operator Experience}
 While engineering design establishes operational parameters, it is the operator who understands the plant in real time.
@@
 
-In early prototype plants, small reactor cores were operated at constant loads by teams of scientists and engineers pushing the limits of atomic knowledge \autocite{moderninstruments}.
-As commercial nuclear power scaled up, these systems became vastly more complex, and measurement accuracy and control precision grew increasingly important.
+In early prototype plants, small reactor cores were operated at constant loads
+by teams of scientists and engineers pushing the limits of atomic knowledge
+\autocite{moderninstruments}. As commercial nuclear power scaled up, these
+systems became vastly more complex, and measurement accuracy and control
+precision grew increasingly important. Figure~\ref{fig:maneuvering} shows
+experienced operators aboard an S5W-era submarine in a shutdown configuration.
+\begin{figure}[H]
+	\centering
+	\includegraphics{maneuvering-watch.jpg}
+	\caption{The small cramped room in a nuclear submarine named maneuvering, in
+	the front is the Shutdown Reactor Operator, and to his right is the Electrical
+Operator.}
+	\label{fig:maneuvering}
+\end{figure}
 
@@
 
-\subsection{Ethical Limits of Automation}
-
+\subsubsection*{The Danger in Automation}
 While automation has dramatically improved reactor safety, it must be implemented with caution.
@@
 \end{figure}
-\section{Economic Realities and Infrastructure Decline}
+
+\subsection{Economic Realities and Infrastructure Decline}
+\label{sec:infrastructure}
 
@@
 
-\subsection{Construction and Design of Future Plants}
-
+\subsection{Construction and Design of Future Plants: The Societal Impact}
 The capability to construct new nuclear facilities has also diminished.
@@
 
-\subsection{Migration from Nuclear to Renewable Energy Sectors}
-
-Global investment patterns show a clear transition from nuclear to renewables.
-Engineering giants like Siemens and General Electric have scaled back or entirely exited nuclear ventures in favor of wind, solar, and energy storage \autocite{iea2020}.
-In 2022 alone, global capital expenditures on solar and wind eclipsed nuclear investments by more than fivefold \autocite{wnisr2022}.
-
 This migration is not merely market-driven; it is reinforced by regulatory environments.
@@
 These growing financial burdens make nuclear projects less attractive in both the public and private sectors.
-
-\subsection{Integration Challenges and Opportunities}
 
@@
 \end{document}
+% vim: set ft=tex tw=80 ts=2 sts=2 sw=2 noet spell:

### 2025-06-10 Slides/PaperTwoPresentation/presentation.tex
push

--- 2025-06-06 Slides/PaperTwoPresentation/presentation.tex
@@
 \documentclass{beamer}
+\setbeameroption{show notes}
 \usetheme{Warsaw}
@@
   \titlepage
-  \speak{Welcome. This presentation explores the technological, ethical, and infrastructural dimensions of nuclear power in the modern world.}
+  \note{Welcome.
+This presentation explores the technological, ethical, and infrastructural dimensions of nuclear power in the modern world.
+We will cover lessons from historic accidents, instrumentation, human-machine interface design, and current innovations.
+Each story reveals both vulnerability and resilience.}
 \end{frame}
@@
 \begin{frame}{Outline}
-  \tableofcontents[pausesections]
-  \speak{Here is an overview of the major themes we will discuss today.}
+  \tableofcontents
+  \note{Here is the roadmap for this presentation. Each section builds on the last, culminating in a call for renewed responsibility and innovation in the nuclear field.}
 \end{frame}
@@
 \section{Introduction}
-
+\begin{frame}
+  \centering
+  \includegraphics[width=\textwidth]{enterprise}
+  \notes{i've been hearing a lot of misunderstanding when it comes to this epic photograph taken on july 31, 1964 with the uss enterprise, the uss long island, and the uss bainbridge.
+e = mc^2 refers to the fact that as the mass of the fuel increases, the potential for it to create energy is proprotional with that multiplied times the speed of light squared.
+the us navy was a big push for nuclear power and the enterprise alone had 8 submarine reactors powering her.}
+\end{frame}
 \begin{frame}{Motivation}
-  \speak{Nuclear energy offers unmatched energy density and reliability. Yet its history demands vigilance, ethics, and technical excellence.}
+  \subsubsection*{Why Nuclear Now?}
   \begin{itemize}
-    \item Low emissions, high stability
-    \item Risk perception shaped by major accidents
-    \item Focus of this work: instrumentation, human oversight, infrastructure
+    \item High energy density and steady base-load power
+    \item Nearly zero emissions, independent of weather
+    \item Risks of failure—catastrophic if unmanaged
   \end{itemize}
+  \notes{Despite fears, nuclear remains one of the most efficient and clean energy sources. But it requires precision and care at every level.}
 \end{frame}
 
-\section{Instrumentation and Control}
+\begin{frame}{The Fission Process}
+  \subsubsection*{Where It All Begins}
+  \begin{equation}
+    ^{235}\text{U} + ^{1}\text{n} \rightarrow ^{92}\text{Kr} + ^{141}\text{Ba} + 3^{1}\text{n} + \text{Energy} \approx 200~\text{MeV}
+  \end{equation}
+  \begin{itemize}
+    \item A neutron collides with uranium-235, causing it to split
+    \item Releases energy and more neutrons → possible chain reaction
+  \end{itemize}
+  \note{This is the reaction that drives nuclear power. The energy release is immense, but so is the potential for instability if left uncontrolled.}
+\end{frame}
+
+\begin{frame}{Reactivity and Control}
+  \subsubsection*{Choreographing Neutrons}
+  \begin{equation}
+    \rho = \frac{k_{\text{eff}} - 1}{k_{\text{eff}}}
+  \end{equation}
+  \begin{itemize}
+    \item $k_{\text{eff}}$: effective neutron multiplication factor
+    \item $\rho > 0$: supercritical (power increases)
+    \item $\rho = 0$: critical (steady power)
+    \item $\rho < 0$: subcritical (power decreases)
+  \end{itemize}
+  \note{Instrumentation exists to measure and control reactivity—avoiding runaway reactions like those at SL-1 and Chernobyl.}
+\end{frame}
+
+\section{Reactor Designs and Sensors}
 
 \begin{frame}{Types of Reactors}
-  \speak{Different reactor types use different coolants and designs, influencing their control strategies.}
   \begin{itemize}
@@
     \item Boiling Water Reactor (BWR)
-    \item CANDU (Heavy Water Reactor)
-    \item Gas-Cooled Reactor (AGR)
+    \item Heavy Water Reactor (CANDU)
+    \item Advanced Gas-cooled Reactor (AGR)
+  \end{itemize}
+  \note{Each design uses different coolants and moderators, which affect safety logic and monitoring strategies.}
+\end{frame}
+
+\begin{frame}{PWR Instrumentation}
+  \centering
+  \includegraphics[width=0.8\textwidth]{instrumentation.png}
+  \note{PWRs are the most common worldwide. Sensors track neutron flux, coolant temperature, rod position, pressure, and flow rate.}
+\end{frame}
+
+\begin{frame}{Automatic Protection Systems}
+  \subsubsection*{SCRAMs and Interlocks}
+  \note{Even the best operator can’t always react fast enough.
+That’s where automatic systems step in.
+SCRAMs instantly insert control rods.
+Protection logic compares real-time data to safety limits, triggering shutdowns, alarms, or backup actuation.}
+  \begin{itemize}
+    \item Reactor protection systems (RPS): monitor reactivity, temperature, pressure
+    \item Logic interlocks: prevent unsafe configurations
+    \item Hardwired paths with digital backups
   \end{itemize}
 \end{frame}
-
-\begin{frame}{Instrumentation in a PWR}
-  \speak{This diagram shows how sensors monitor power, pressure, and temperature throughout a PWR.}
-  \includegraphics[width=0.85\textwidth]{images/instrumentation.png}
+\begin{frame}{Control Logic}
+  \centering
+  \includegraphics[width=0.95\textwidth]{safetylogic}
 \end{frame}
 
-\section{Historical Lessons}
+\section{Historical Accidents}
 
-\begin{frame}{SL-1 and Three Mile Island}
-  \speak{SL-1 and Three Mile Island showed the risks of inadequate interlocks and overwhelming operator interfaces.}
-  \begin{itemize}
-    \item SL-1: control rod ejection, fatal steam explosion
-    \item TMI: misread signals led to core exposure
-    \item Both: highlight flaws in human-system integration
-  \end{itemize}
+\begin{frame}{SL-1: Prompt Critical Accident}
+  \centering
+  \includegraphics[width=0.4\textwidth]{sl1diagram.jpg}
+  \note{Local firefighters arrived in response to a fire alarm and found the facility abandoned.
+Coffee cups remained warm, and food was left uneaten. As radiation alarms activated on their
+dosimetry equipment, they realized a serious radiological event had occurred. Two operators
+were located dead from radiation exposure. The third, Richard Legg, was discovered impaled
+and pinned to the containment ceiling by a control rod—ejected upward during the reactor’s
+prompt critical event.}
 \end{frame}
 
-\begin{frame}{Chernobyl and Fukushima}
-  \speak{Chernobyl and Fukushima illustrate the danger of flawed assumptions and the limits of automation.}
-  \includegraphics[width=0.75\textwidth]{images/chernobylafter.jpg}
-  \includegraphics[width=0.75\textwidth]{images/fukushimaafter.jpg}
+\begin{frame}{SL-1: Prompt Critical Accident}
+  \begin{itemize}
+    \item Occurred January 3, 1961 in Idaho Falls.
+    \item A single control rod was withdrawn manually beyond safe limits.
+    \item Caused an instantaneous power excursion and steam explosion.
+    \item All three operators died; first fatal U.S. nuclear accident.
+  \end{itemize}
+  \note{Despite repeated reports of sticking rods, the system lacked mechanical or procedural
+interlocks to prevent rapid withdrawal. A single operator retained full manual control, with
+no redundancy or supervisory lockouts. The SL-1 event shows the critical role of human
+factors engineering, particularly in minimizing design interfaces that allow unsafe manual
+operations}
 \end{frame}
 
-\section{Human Factors}
+\begin{frame}{Three Mile Island: Partial Core Meltdown}
+  \centering
+  \includegraphics[width=0.75\textwidth]{tmicontrolroom.png}
+  \note{The Three Mile Island Unit 2 (TMI-2) accident occurred on March 28, 1979, near Harrisburg,
+Pennsylvania, and remains the most serious commercial nuclear accident in the United States.
+It was precipitated by the failure of a pressure-operated relief valve (PORV) that became
+stuck open during a minor malfunction. The valve allowed coolant to escape from the
+pressurizer, but due to inadequate instrumentation, operators believed it had closed properly}
+\end{frame}
+
+\begin{frame}{Three Mile Island: Partial Core Meltdown}
+  \begin{itemize}
+    \item March 28, 1979 in Pennsylvania.
+    \item Equipment failure: relief valve stuck open.
+    \item Operator misinterpretation led to coolant pump shutdown.
+    \item Reactor overheated—partial meltdown of core.
+  \end{itemize}
+  \note{Prior to this event, the relationship between human performance and system design was
+underappreciated in reactor control engineering. It was often assumed that proper training
+and “common sense” would be sufficient for managing system faults. However, the TMI-
+2 accident highlighted that even well-trained operators can make catastrophic errors when
+systems are not designed to support human cognition under stress.
+As a result of this accident, the field of human factors engineering gained legitimacy within
+the nuclear industry. Organizational and industrial psychology—especially the study of
+ergonomics and operator-centered design—emerged as crucial disciplines in instrumentation
+and control system development [6]. Operators needed interfaces not merely for function, but
+for situational awareness and decision-making under pressure. This marked a turning point
+in control room design philosophies, ushering in era-defining updates to alarm management,
+display hierarchy, and interface logic}
+\end{frame}
+
+\begin{frame}{Chernobyl: Uncontrolled Power Surge}
+  \centering
+  \includegraphics[width=0.65\textwidth]{chernobylafter.jpg}
+  \note{The test intended to verify whether the rotational inertia
+of the turbine could temporarily power the emergency coolant pumps in the event of a
+grid power failure [7]. The RBMK-1000 reactor involved was a Soviet-designed graphite-
+moderated, water-cooled system—flawed by both design and execution}
+\end{frame}
+
+\begin{frame}{Chernobyl: Uncontrolled Power Surge}
+  \begin{itemize}
+    \item April 26, 1986 in Pripyat, USSR.
+    \item Unsafe test conducted at low power with flawed RBMK reactor.
+    \item Positive void coefficient caused rapid reactivity increase.
+    \item Control rods exacerbated the surge due to graphite tips.
+    \item Reactor exploded; massive radioactive release.
+  \end{itemize}
+  \note{The core overheated instantly, rupturing fuel channels and vaporizing coolant. A vi-
+olent steam explosion lifted the reactor’s 2,000-ton upper biological shield. A second ex-
+plosion—possibly from hydrogen or steam—further breached the structure and exposed the
+graphite moderator, which caught fire. Figure 6 shows the aftermath of the explosion mere
+hours after the explosion. The fire lofted radioactive particles into the upper atmosphere,
+affecting much of Europe }
+\end{frame}
+
+\begin{frame}
+  \centering
+  \includegraphics[width=0.65\textwidth]{elephantsfoot.png}
+  \note{The aftermath of Chernobyl catalyzed global reevaluation of nuclear safety, reactor de-
+sign, and operator training. The international community pushed for increased transparency,
+safety audits, and enhanced containment strategies. The photograph in Figure 7 famously
+shows the “Elephant’s Foot,” a deadly corium formation. The strange visual static in the
+image is not lightning, but film degradation from intense radiation—testament to the un-
+precedented radioactive environment at the site}
+\end{frame}
+
+\begin{frame}{Fukushima Daiichi: Station Blackout}
+  \begin{itemize}
+    \item March 11, 2011 in Japan.
+    \item Magnitude 9.0 earthquake triggered tsunami.
+    \item Backup diesel generators flooded—loss of core cooling.
+    \item Hydrogen buildup led to explosions in three reactors.
+  \end{itemize}
+  \note{The Fukushima-Daiichi Nuclear Power Station, operated by TEPCO, was critically
+affected. Although the three operating reactors—Units 1, 2, and 3—successfully SCRAMed
+(shut down automatically), the tsunami flooded the site and disabled both the offsite grid
+connections and emergency diesel generators [12]. This unprecedented loss of power across all
+units is illustrated in Figure 8, which shows the reactor buildings after successive hydrogen
+explosions.}
+\end{frame}
+
+\begin{frame}
+  \centering
+  \includegraphics[width=\textwidth]{fukushimaafter.jpg}
+  \note{Operators at Fukushima recognized the danger but were rendered effectively blind by
+the loss of power. Unlike U.S. Navy reactor systems—where portable test equipment such
+as Fluke digital multimeters can be attached to terminals for passive thermocouple read-
+ings—the Fukushima design lacked redundant manual monitoring options. In a desperate
+act of ingenuity, personnel retrieved car batteries from nearby vehicles and wired them in
+series to restore minimal voltage to power essential instrumentation. Unfortunately, by the
+time this improvised power source was implemented, core damage in Units 1, 2, and 3 was
+already underway, with partial or full meltdown confirmed in post-event analyses}
+\end{frame}
+
+\section{Human and Ethical Design}
+
+\begin{frame}{Human Operators: Essential Links}
+  \centering
+  \includegraphics[width=0.75\textwidth]{maneuvering-watch.jpg}
+  \note{}
+\end{frame}
+
+\begin{frame}{Human Operators: Essential Links}
+  \begin{itemize}
+    \item Human error was central to SL-1, TMI, and Chernobyl.
+    \item Operators are responsible for interpreting ambiguous or conflicting signals.
+    \item Training, vigilance, and mental workload matter.
+  \end{itemize}
+  \note{Despite automation, human judgment determines plant safety. Accidents often trace back to either misinterpretation or improper training.}
+\end{frame}
 
 \begin{frame}{Human Factors Engineering}
-  \speak{Human factors engineering is vital in designing systems that support, not hinder, decision making.}
   \begin{itemize}
-    \item Ergonomic control rooms
-    \item Full-scope simulation
-    \item Situational awareness tools
+    \item Post-TMI, HFE became a formal discipline in nuclear plant design.
+    \item Goals: reduce confusion, prevent overload, clarify alarms.
+    \item INPO and NRC led redesigns of control interfaces.
   \end{itemize}
+  \note{Human-machine interface redesign became essential after TMI. Alarm logic, indicator layout, and system feedback were all reengineered.}
 \end{frame}
 
-\begin{frame}{Ethics of Automation}
-  \speak{Automation must be supervised. Tunnel vision and outdated assumptions can cause failure without human oversight.}
-  \begin{itemize}
-    \item Humans mitigate uncertainty and change
-    \item Ethics demand human-in-the-loop
-  \end{itemize}
+\begin{frame}{Operator Training}
+  \centering
+  \includegraphics[width=0.75\textwidth]{simulators.jpg}
 \end{frame}
 
-\section{Economic Challenges}
-
-\begin{frame}{Waste and Infrastructure}
-  \speak{The economic burden of waste management and infrastructure decline limits nuclear's future.}
+\begin{frame}{Operator Training}
   \begin{itemize}
-    \item High refueling and decommissioning costs
-    \item No permanent U.S. waste site
-    \item Industry contraction post-Fukushima
+    \item U.S. Navy trains operators with year-long theoretical and practical curriculum.
+    \item Qualification cards enforce step-by-step system mastery.
+    \item Real-time simulators mimic full plant behavior—including failures.
   \end{itemize}
+  \note{Simulators allow safe repetition of emergency scenarios. This procedural rigor is unmatched in civilian sectors.}
 \end{frame}
 
-\begin{frame}{Hopeful Developments}
-  \speak{New efforts like Vogtle and SMRs show a path forward. Nuclear remains vital for base-load generation.}
+\begin{frame}{Ethical Oversight in Automation}
+  \centering
+  \includegraphics[width=0.65\textwidth]{ethicaldesign.jpg}
+\end{frame}
+
+\begin{frame}{Ethical Oversight in Automation}
   \begin{itemize}
-    \item Vogtle Unit 4 completed
-    \item Passive safety designs
-    \item Small Modular Reactors
+    \item Systems must support—not replace—human oversight.
+    \item Overreliance on automation can erode accountability.
+    \item Ethical design considers failure modes, transparency, and operator input.
   \end{itemize}
+  \note{Designing for ethical oversight means respecting human limits while reinforcing responsibility. The worst outcomes often arise when operators are sidelined.}
+\end{frame}
+
+\section{The Road Ahead}
+
+\begin{frame}{Why Nuclear Declined}
+  \subsubsection*{Policy, Perception, and Paralysis}
+  \begin{itemize}
+    \item Each accident—from SL-1 to Fukushima—prompted strict new regulations.
+    \item Longer licensing cycles delayed new builds by decades.
+    \item Public opposition and fear, not technical failure, stalled the industry.
+    \item Skilled labor and supply chains diminished as construction halted.
+  \end{itemize}
+  \note{The world walked away not because nuclear failed—but because faith in its management eroded. Regulation became slower, and expertise drifted away.}
+\end{frame}
+
+\begin{frame}{Challenges Today}
+  \begin{itemize}
+    \item Most operating reactors in the U.S. are past mid-life.
+    \item Replacing the aging nuclear workforce is a growing challenge.
+    \item Engineering firms that once supported nuclear have pivoted to renewables.
+    \item Safety margins remain—but the supporting infrastructure has weakened.
+  \end{itemize}
+  \note{Many of the companies and capabilities that built the first generation of plants no longer exist in their original form.}
 \end{frame}
@@
 
-\begin{frame}{Final Thoughts}
-  \speak{With humility and innovation, nuclear energy can be safe, ethical, and essential in the fight against climate change.}
+\begin{frame}{Conclusion: A Deliberate Future}
+  \centering
+  \includegraphics[width=0.55\textwidth]{vogtle}
+\end{frame}
+
+\begin{frame}{Conclusion: A Deliberate Future}
   \begin{itemize}
-    \item Ethics + engineering = resilient systems
-    \item Human vigilance must never be replaced
-    \item The atom, split with care, still holds promise
+    \item Nuclear safety is engineered, not assumed.
+    \item Instrumentation and human oversight must evolve together.
+    \item Ethical design puts operators in control, not out of the loop.
+    \item The tools are available—what remains is the will to rebuild trust.
   \end{itemize}
+  \note{Safe nuclear power isn’t an inevitability—it’s a discipline. What matters is not just how reactors are built, but how they are overseen.}
 \end{frame}

### 2025-06-11 Slides/PaperTwoPresentation/presentation.tex
work on presentation

@@
 \documentclass{beamer}
-\setbeameroption{show notes}
+%\setbeameroption{show only notes}
 \usetheme{Warsaw}
@@
   \tableofcontents
-  \note{Here is the roadmap for this presentation. Each section builds on the last, culminating in a call for renewed responsibility and innovation in the nuclear field.}
+   \note{I will be going through some motivation as to why nuclear power is still safe, explain a little bit about how the fission process works and the instrumentation that is used to control the reaction.
+I will then go on to emphasize the broader impacts and tell the story as to why nuclear power has not been as widely adopted as it could have been.
+I will discuss the improvements to human factors engineering because of these, and talk about the economic, environmental, and societal impacts that these accidents have had on the nuclear industry as a whole.}
 \end{frame}
@@
 \section{Introduction}
+
 \begin{frame}
-  \centering
+  \begin{columns}
+    \column{0.48\textwidth}
   \includegraphics[width=\textwidth]{enterprise}
-  \notes{i've been hearing a lot of misunderstanding when it comes to this epic photograph taken on july 31, 1964 with the uss enterprise, the uss long island, and the uss bainbridge.
-e = mc^2 refers to the fact that as the mass of the fuel increases, the potential for it to create energy is proprotional with that multiplied times the speed of light squared.
-the us navy was a big push for nuclear power and the enterprise alone had 8 submarine reactors powering her.}
+    \column{0.48\textwidth}
+    \begin{itemize}
+      \item Nuclear power plants are designed to be safe and efficient.
+      \item Instrumentation plays a critical role in monitoring and control.
+      \item Human oversight is essential for safe operation.
+    \end{itemize}
+  \note{I've been hearing a lot of misunderstanding when it comes to this epic photograph taken on july 31, 1964 with the uss enterprise, the uss long beach, and the uss bainbridge.
+People generally think that the sailors aboard irrelevantly spelled $e = mc^2$.
+This is just a response out of ignorance however.
+The formula refers to the fact that as the mass of the fuel increases, the potential for it to create energy is proprotional with that multiplied times the speed of light squared.
+This was a demonstration of force and showing the world the US Navy's capabilities in the nuclear age.
+The Enterprise was the first nuclear-powered aircraft carrier with 8 submarine reactors.
+The uss long beach and uss bainbridge were also powered by dual surface developed reactors.}
+\end{columns}
 \end{frame}
+
 \begin{frame}{Motivation}
@@
   \end{itemize}
-  \notes{Despite fears, nuclear remains one of the most efficient and clean energy sources. But it requires precision and care at every level.}
+  \note{Despite fears, nuclear remains one of the most efficient and clean energy sources. But it requires precision and care at every level.}
 \end{frame}
@@
   \end{itemize}
-  \note{Instrumentation exists to measure and control reactivity—avoiding runaway reactions like those at SL-1 and Chernobyl.}
+  \note{This is our friend Keff.
+The effective multiplication factor.
+The effective multiplication factor is just a measure of the average amount of thermal neutrons produced in one neutron lifecycle.
+The ratio of keff - 1 divided by keff forms a measurement called reactivity.
+A critical reactor has a keff of 1, a subcritical reactor has a keff of less than 1, and a supercritical reactor has a keff of greater than 1.
+}
 \end{frame}
@@
 
-\begin{frame}{Types of Reactors}
+\begin{frame}{Types of Reactors and PWR Instrumentation}
+  \begin{columns}
+    \column{0.48\textwidth}
+    \includegraphics[width=\textwidth]{instrumentation.png}
+
+    \column{0.48\textwidth}
     \begin{itemize}
@@
     \end{itemize}
-  \note{Each design uses different coolants and moderators, which affect safety logic and monitoring strategies.}
-\end{frame}
-
-\begin{frame}{PWR Instrumentation}
-  \centering
-  \includegraphics[width=0.8\textwidth]{instrumentation.png}
-  \note{PWRs are the most common worldwide. Sensors track neutron flux, coolant temperature, rod position, pressure, and flow rate.}
+  \note{
+    Each design uses different coolants and moderators, which affect safety logic and monitoring strategies.\\
+    PWRs are the most common worldwide. Sensors track neutron flux, coolant temperature, rod position, pressure, and flow rate.
+  }
+  \end{columns}
 \end{frame}
@@
   \subsubsection*{SCRAMs and Interlocks}
-  \note{Even the best operator can’t always react fast enough.
-That’s where automatic systems step in.
-SCRAMs instantly insert control rods.
-Protection logic compares real-time data to safety limits, triggering shutdowns, alarms, or backup actuation.}
+  \begin{columns}
+    \column{0.49\textwidth}
+    \includegraphics[width=0.95\textwidth]{safetylogic}
+
+    \column{0.48\textwidth}
   \begin{itemize}
@@
   \end{itemize}
-\end{frame}
-\begin{frame}{Control Logic}
-  \centering
-  \includegraphics[width=0.95\textwidth]{safetylogic}
+  \note{Even the best operator can’t always react fast enough.
+That’s where automatic systems step in.
+SCRAMs instantly insert control rods.
+Protection logic compares real-time data to safety limits, triggering shutdowns, alarms, or backup actuation.}
+\end{columns}
 \end{frame}
 
-\section{Historical Accidents}
+\section{Broader Impacts}
+\subsection{Historical Accidents: The Environmental Impact}
+\begin{frame}{SL-1: Prompt Critical Accident}
+  \begin{columns}
+    \column{0.48\textwidth}
+  \includegraphics[width=\textwidth]{sl1diagram.jpg}
 
-\begin{frame}{SL-1: Prompt Critical Accident}
-  \centering
-  \includegraphics[width=0.4\textwidth]{sl1diagram.jpg}
+  \column{0.48\textwidth}
+  \begin{itemize}
+    \item Occurred January 3, 1961 in Idaho Falls.
+    \item A single control rod was withdrawn manually beyond safe limits.
+    \item Caused an instantaneous power excursion and steam explosion.
+    \item All three operators died; first fatal U.S. nuclear accident.
+  \end{itemize}
   \note{Local firefighters arrived in response to a fire alarm and found the facility abandoned.
@@
 prompt critical event.}
-\end{frame}
-
-\begin{frame}{SL-1: Prompt Critical Accident}
-  \begin{itemize}
-    \item Occurred January 3, 1961 in Idaho Falls.
-    \item A single control rod was withdrawn manually beyond safe limits.
-    \item Caused an instantaneous power excursion and steam explosion.
-    \item All three operators died; first fatal U.S. nuclear accident.
-  \end{itemize}
-  \note{Despite repeated reports of sticking rods, the system lacked mechanical or procedural
-interlocks to prevent rapid withdrawal. A single operator retained full manual control, with
-no redundancy or supervisory lockouts. The SL-1 event shows the critical role of human
-factors engineering, particularly in minimizing design interfaces that allow unsafe manual
-operations}
+\end{columns}
 \end{frame}
@@
 \begin{frame}{Three Mile Island: Partial Core Meltdown}
-  \centering
-  \includegraphics[width=0.75\textwidth]{tmicontrolroom.png}
+  \begin{columns}
+    \column{0.48\textwidth}
+  \includegraphics[width=\textwidth]{tmicontrolroom.png}
+
+\column{0.48\textwidth}
+  \begin{itemize}
+    \item March 28, 1979 in Pennsylvania.
+    \item Equipment failure: relief valve stuck open.
+    \item Operator misinterpretation led to coolant pump shutdown.
+    \item Reactor overheated—partial meltdown of core.
+  \end{itemize}
+\end{columns}
+
   \note{The Three Mile Island Unit 2 (TMI-2) accident occurred on March 28, 1979, near Harrisburg,
@@
 
-\begin{frame}{Three Mile Island: Partial Core Meltdown}
+\begin{frame}{Chernobyl: Uncontrolled Power Surge}
+  \begin{columns}
+    \column{0.48\textwidth}
+  \includegraphics[width=\textwidth]{chernobylafter.jpg}
+\column{0.48\textwidth}
   \begin{itemize}
-    \item March 28, 1979 in Pennsylvania.
-    \item Equipment failure: relief valve stuck open.
-    \item Operator misinterpretation led to coolant pump shutdown.
-    \item Reactor overheated—partial meltdown of core.
+    \item April 26, 1986 in Pripyat, USSR.
+    \item Unsafe test conducted at low power with flawed RBMK reactor.
+    \item Control rods exacerbated the surge due to graphite tips.
+    \item Reactor exploded; massive radioactive release.
   \end{itemize}
-  \note{Prior to this event, the relationship between human performance and system design was
-underappreciated in reactor control engineering. It was often assumed that proper training
-and “common sense” would be sufficient for managing system faults. However, the TMI-
-2 accident highlighted that even well-trained operators can make catastrophic errors when
-systems are not designed to support human cognition under stress.
-As a result of this accident, the field of human factors engineering gained legitimacy within
-the nuclear industry. Organizational and industrial psychology—especially the study of
-ergonomics and operator-centered design—emerged as crucial disciplines in instrumentation
-and control system development [6]. Operators needed interfaces not merely for function, but
-for situational awareness and decision-making under pressure. This marked a turning point
-in control room design philosophies, ushering in era-defining updates to alarm management,
-display hierarchy, and interface logic}
-\end{frame}
 
-\begin{frame}{Chernobyl: Uncontrolled Power Surge}
-  \centering
-  \includegraphics[width=0.65\textwidth]{chernobylafter.jpg}
   \note{The test intended to verify whether the rotational inertia
@@
 grid power failure [7]. The RBMK-1000 reactor involved was a Soviet-designed graphite-
-moderated, water-cooled system—flawed by both design and execution}
-\end{frame}
-
-\begin{frame}{Chernobyl: Uncontrolled Power Surge}
-  \begin{itemize}
-    \item April 26, 1986 in Pripyat, USSR.
-    \item Unsafe test conducted at low power with flawed RBMK reactor.
-    \item Positive void coefficient caused rapid reactivity increase.
-    \item Control rods exacerbated the surge due to graphite tips.
-    \item Reactor exploded; massive radioactive release.
-  \end{itemize}
-  \note{The core overheated instantly, rupturing fuel channels and vaporizing coolant. A vi-
+moderated, water-cooled system—flawed by both design and executionhe core overheated instantly, rupturing fuel channels and vaporizing coolant. A vi-
 olent steam explosion lifted the reactor’s 2,000-ton upper biological shield. A second ex-
@@
 affecting much of Europe }
+\end{columns}
 \end{frame}
@@
 \begin{frame}{Fukushima Daiichi: Station Blackout}
+  \begin{columns}
+    \column{0.48\textwidth}
+  \includegraphics[width=\textwidth]{fukushimaafter.jpg}
+  \column{0.48\textwidth}
   \begin{itemize}
@@
 explosions.}
-\end{frame}
-
-\begin{frame}
-  \centering
-  \includegraphics[width=\textwidth]{fukushimaafter.jpg}
-  \note{Operators at Fukushima recognized the danger but were rendered effectively blind by
-the loss of power. Unlike U.S. Navy reactor systems—where portable test equipment such
-as Fluke digital multimeters can be attached to terminals for passive thermocouple read-
-ings—the Fukushima design lacked redundant manual monitoring options. In a desperate
-act of ingenuity, personnel retrieved car batteries from nearby vehicles and wired them in
-series to restore minimal voltage to power essential instrumentation. Unfortunately, by the
-time this improvised power source was implemented, core damage in Units 1, 2, and 3 was
-already underway, with partial or full meltdown confirmed in post-event analyses}
-\end{frame}
-
-\section{Human and Ethical Design}
-
-\begin{frame}{Human Operators: Essential Links}
-  \centering
-  \includegraphics[width=0.75\textwidth]{maneuvering-watch.jpg}
-  \note{}
+\end{columns}
 \end{frame}
@@
 \begin{frame}{Human Operators: Essential Links}
+  \begin{columns}
+    \column{0.48\textwidth}
+  \includegraphics[width=\textwidth]{maneuvering-watch.jpg}
+    \note{Human operators are the last line of defense.
+They interpret complex data, make decisions under pressure, and ensure systems operate safely.
+The operators in this picutre are in a room named maneuvering.
+In this tight cramped space, you can see the control panel to control the reactor, and further back the electric plant control panel of a S5W reactor powered ship.}
+
+    \column{0.48\textwidth}
     \begin{itemize}
-    \item Human error was central to SL-1, TMI, and Chernobyl.
-    \item Operators are responsible for interpreting ambiguous or conflicting signals.
-    \item Training, vigilance, and mental workload matter.
+      \item Operators monitor instrumentation and respond to alarms.
+      \item They must understand system behavior and safety limits.
+      \item Training and vigilance are critical for safe operation.
     \end{itemize}
-  \note{Despite automation, human judgment determines plant safety. Accidents often trace back to either misinterpretation or improper training.}
+  \end{columns}
 \end{frame}
@@
 \begin{frame}{Human Factors Engineering}
+  \begin{columns}
+    \column{0.48\textwidth}
+  \includegraphics[width=\textwidth]{simulators.jpg}
+  \column{0.48\textwidth}
   \begin{itemize}
@@
   \end{itemize}
-  \note{Human-machine interface redesign became essential after TMI. Alarm logic, indicator layout, and system feedback were all reengineered.}
-\end{frame}
-
-\begin{frame}{Operator Training}
-  \centering
-  \includegraphics[width=0.75\textwidth]{simulators.jpg}
-\end{frame}
-
-\begin{frame}{Operator Training}
-  \begin{itemize}
-    \item U.S. Navy trains operators with year-long theoretical and practical curriculum.
-    \item Qualification cards enforce step-by-step system mastery.
-    \item Real-time simulators mimic full plant behavior—including failures.
-  \end{itemize}
-  \note{Simulators allow safe repetition of emergency scenarios. This procedural rigor is unmatched in civilian sectors.}
+  \note{Human-machine interface redesign became essential after TMI.
+Alarm logic, indicator layout, and system feedback were all reengineered.
+The need to provide complex and realistic simulation training for operators has increased tenfold since the 1980s.}
+\end{columns}
 \end{frame}
@@
 \begin{frame}{Ethical Oversight in Automation}
-  \centering
-  \includegraphics[width=0.65\textwidth]{ethicaldesign.jpg}
-\end{frame}
-
-\begin{frame}{Ethical Oversight in Automation}
+  \begin{columns}
+    \column{0.48\textwidth}
+  \includegraphics[width=\textwidth]{ethicaldesign.jpg}
+  \column{0.48\textwidth}
   \begin{itemize}
@@
   \note{Designing for ethical oversight means respecting human limits while reinforcing responsibility. The worst outcomes often arise when operators are sidelined.}
+\end{columns}
 \end{frame}
 
-\section{The Road Ahead}
-
+\subsection{The Societal Impact of Nuclear Power}
 \begin{frame}{Why Nuclear Declined}
-  \subsubsection*{Policy, Perception, and Paralysis}
+  \subsubsection*{Societal Impact: Policy, Perception, and Paralysis}
   \begin{itemize}
@@
 \end{frame}
-
+\subsection{The Economic Impact}
 \begin{frame}{Challenges Today}
@@
   \end{itemize}
-  \note{Many of the companies and capabilities that built the first generation of plants no longer exist in their original form.}
+  \note{Many of the companies and capabilities that built the first generation of plants no longer exist in their original form.
+As a result, the ability to construct new certified plants has greatly diminished.
+The nuclear industry is now at a crossroads, where the lessons of the past must inform the future.}
 \end{frame}
@@
 \begin{frame}{Conclusion: A Deliberate Future}
-  \centering
-  \includegraphics[width=0.55\textwidth]{vogtle}
+  \begin{columns}
+    \column{0.45\textwidth}
+    \includegraphics[width=\textwidth]{vogtle}
+    \note{The future of nuclear power depends on our ability to learn from the past, innovate responsibly, and rebuild trust.}
+
+    \column{0.55\textwidth}
+    \begin{itemize}
+      \item Nuclear power is safe when designed, operated, and overseen with care.
+      \item Instrumentation and human oversight must evolve together.
+      \item Ethical design puts operators in control, not out of the loop.
+      \item The tools are available, what remains is the will to rebuild trust.
+    \end{itemize}
+  \end{columns}
 \end{frame}

### 2025-06-12 Slides/PaperTwoPresentation/presentation.tex
Finished the course

@@
 \setbeamertemplate{navigation symbols}{}
+\setbeamertemplate{footline}[frame number]
 \usepackage{graphicx}
@@
     \end{itemize}
-  \note{I've been hearing a lot of misunderstanding when it comes to this epic photograph taken on july 31, 1964 with the uss enterprise, the uss long beach, and the uss bainbridge.
-People generally think that the sailors aboard irrelevantly spelled $e = mc^2$.
+  \note{I've been hearing a lot of misunderstanding when it comes to this photograph taken on july 31, 1964 with the uss enterprise, the uss long beach, and the uss bainbridge.
+I've heard  generally think that the sailors aboard irrelevantly spelled $e = mc^2$.
 This is just a response out of ignorance however.
@@
 
+\subsection{Technological Impact}
 \begin{frame}{Human Operators: Essential Links}
